### 2024-08-02 paper/main.tex
added paper, trimmed overfit test and updated output video

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2024

% ready for submission
\usepackage{neurips_2024}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2024}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2024}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2024}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{float}
\usepackage{bm}

\title{Interpretable Risk Minimization -- Preliminaries}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  David S.~Hippocampus\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}

\maketitle

% \begin{abstract}
%   The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
%   both the left- and right-hand margins. Use 10~point type, with a vertical
%   spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
%   bold, and in point size 12. Two line spaces precede the abstract. The abstract
%   must be limited to one paragraph.
% \end{abstract}

\section{Background \& Motivation}

The standard optimization pipeline for image classification predominantly consists of using a CNN-backbone with a fully-connected layer pointing to the classification targets.

This setup is vulnerable to spurious correlation caused by features like a common background, causing the model to learn `tricks', bypassing actually learning the target features. A simple example is illustrated in the second paragraph of the introduction in \cite{arjovsky2020invariantriskminimization}.

A popular post-hoc technique -- Class Activation Mappings \citep{zhou2016learning} -- helps diagnose this by generating a heatmap that visualizes the regions of the image used by the model to generate the prediction. Further research uncovered that the interpretation used by CAMs, and it's derivative GradCAMs were unfaithful as a consequence of the averaging step.

HiResCAM \citep{draelos2020use} is a variant of GradCAM \citep{selvaraju2017grad} that demonstrates that their visualization \textbf{provably recovers the input-dependent abnormality score} from the final classification. Therefore, there is a guarantee associated with the faithfulness of the visualization.

This approach derives an interpretable training objective using HiResCAMs, which replaces and outperforms cross-entropy loss to build an intrinsically interpretable, performant model.

\section{Methodology}

The experimental setup is a CNN-backbone (here, ResNet-50) with a single fully connected softmax-activated layer for classification. The backbone outputs feature maps $\bm{A} \in \mathbb{R}^{F \times D_1 \times D_2}$, which is flattened to $\bm{A}_F \in \mathbb{R}^{FD_1D_2 \times 1}$ and fed to the linear layer with $\bm{w}_c \in \mathbb{R}^{1 \times FD_1D_2}$. The logits obtained in $\mathbb{R}^{c}$, where $c$ is the total number of classes is then passed through softmax to generate the final probabilities.

\subsection{Observation: HiResCAMs Explain Class Score, Not Class Probabilites}

Per \cite{draelos2020use}, a HiResCAM is defined as follows:
\begin{gather}
	\tilde{\mathcal{\bm{A}}}_c^{\text{HiResCAM}} = \sum^F_{f=1} \frac{\partial s_c}{\partial \bm{A}^f} \odot \bm{A}^f
\end{gather}

Where $s_c$ denotes the logit for a single class. $\tilde{\mathcal{\bm{A}}}_c^{\text{HiResCAM}}$ generates a class-specific CAM. For the CAM architecture, \cite{draelos2020use} (section 3.6.2) shows that:
\begin{gather}
	\sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
\end{gather}
Therefore, we can recreate the entire input-dependent part of the score using HiResCAMs. Doing this for each class $c$, we obtain the bias-removed logit vector $\in \mathbb{R}^{c}$.

The standard interpretation of this result uses the obtained activation maps as explanations for the prediction. Popular practice is to follow this up by ReLU and normalization \citep{draelos2020use}. However, since softmax is computed over the logit vector, this interpretation is flawed.

We can consider the backbone as a kernel \citep{jacot2018neural} that fundamentally attempts to linearly separate the correct class, and the softmax-activated final layer being logistic regression.

In this setup, each class has it's own set of weights that feedforwards independently, but trains collaboratively. It assigns a score for it's specific value, which is independent since one set of weights does not affect the other over a forward pass. If we take as example logits $[1,2]$ and $[-2,-1]$, we see that $\sigma([1,2]) = \sigma([-2,1]) = [0.2689, 0.7311]$. The prediction probabilities remain the same, despite different values for the same score $s_c$. Therefore, comparing \textit{absolute score} is a misrepresentation of the internal model state.

\subsection{Defining the Contrastive Optimization Objective}

We start by \textbf{comparing the contrast} of the activation maps with respect to one another. If for a single point $\bm{A}_{i,j}$, $\bm{w}_c > \bm{w}_{c'}$, that point is more likely to represent the object belonging to class $c$, even if the absolute value of $\bm{w}_{c'}$ is high. This information is lost during the flattening, which embeds positional invariance.

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values, which is an activation map instead. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

\subsection{Constraining Optimization to an Interpretable Basis}

The distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.

Segmentations maps provide this exact information. It describes the region over which the object is present. We therefore define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \\
	\mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{y}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{y'}^{\text{HiResCAM}} \\
	\mathcal{\bm{A}}^{\text{background}}_{i,j} := \begin{cases}
		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{D_1,D_2}_{i,j} \mathcal{\bm{A}}^{\text{foreground}}_{i,j}
\end{gather}

The cost function has a threefold purpose:
\begin{enumerate}
	\item Minimize the divergence of the contrastive activation map to the segmentation map.
	\item Minimize the absolute value of the contrastive map where the segmentation map is the background.
	\item Maximize the value of the contrastive map where the segmentation map is the foreground.
\end{enumerate}

$\lambda_1$ \& $\lambda_2$ are hyperparameters that scale the activation maps, since softmax computer over highly similar values close to zero produces uniform probability distributions. $\lambda_3$ weights the importance of the background not affecting the final prediction. For our experiments, $\lambda_1 = 10, \lambda_2 = 50, \lambda_3 = 100$ were set empirically, without rigorous optimization.

\subsection{Training Detail}

The Oxford-IIIT Pets dataset contains a classification goal alongside high-quality segmentation maps, and the binary classification case (image containing a dog or cat) was trained on. The architecture used for training was ResNet-50, initialized with random parameters and two key modifications:

\paragraph{Removed final downsampling} The final downsampling layer converts the latent feature maps from $D_1 = D_2 = 14$ to $7$. This prohibitively reduces the size of the activation map, and makes it hard to capture relevant features. Therefore, we replace the stride of the final downsampling convolution to $(1,1)$, matching that of the definition used through the rest of ResNet.

\paragraph{Removed final bias} The linear layer's bias is not used in the computation of the class activation map, and therefore cannot be optimized by the training procedure, as $\nabla_b \mathcal{\bm{J}}(\theta, \mathcal{D})$ = 0. We omit the bias from the final model architecture.

\paragraph{Removed final BatchNormalization \& ReLU} Since the HiResCAM definition for CAM architectures assumes the convolution followed by global average pooling and the final linear layer only, the current explanation does not directly explain the class score. We therefore also neutralize the final two-layers, which successfully recovers the faithfulness guarantee.

The model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits. The total time for training was $\approx$1hr.

\paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization}.

\section{Results}

An encouraging result from this experiment is that the model trained on the contrastive objective outperforms the exact same model architecture (with the final bias, batchnorm and ReLU re-added in) on cross entropy loss, even though the control case was trained explicitly to reduce cross entropy.

Our approach's performance on the Oxford-IIIT Pets Dataset to the default is as follows:

\begin{table}[h]
	\centering
	\begin{tabular}{c|ccc}
		\toprule
		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
		\midrule
		Cross-Entropy & 0.628 & 68.2\% & 64.1\% \\
		Interpretable (Ours) & \bf 0.354 & \bf 96.2\% & \bf 89.2\% \\
		\bottomrule
	\end{tabular}
\end{table}

This dataset contains a class imbalance (4978 dogs to 2371 cats), however no training modification was made to correct the class imbalance.

\begin{figure}[H]
	\centering
	\includegraphics[width=.8\textwidth]{img/plots.png}
	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy. The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_dd3fe52afeca42539f87d5da6ac4ea3f\%22,\%22m_e21074d967b741dfb772746b113fe761\%22]}{here}. The orange line represent cross-entropy loss, while the blue line represents the contrastive approach.}
\end{figure}

Crucially, on comparing contrastive activation maps, we can see that each of the cases of the random sample, the model uses the target region for prediction and not spurious features. In the case of the standard training procedure, this is not the case, and the model does not focus on the target in making it's prediction.

\begin{figure}[H]
	\centering
	\includegraphics[width=.4\textwidth]{img/default_cam.png}
	\hspace{4em}
	\includegraphics[width=.4\textwidth]{img/contrastive_cam.png}
	\caption{Left: contrastive activation maps from the default training run. Right: contrastive activation maps from the contrastive training run.}
\end{figure}

\section{Upcoming / In Progress Tasks}

\paragraph{Extending to the Multiclass Setting} Currently, the implementation computes contrastive activation maps from the target to each other class. However, the best approach to optimization needs to be tested. The likely approach would be to compute the mean error over each contrastive activation map.

\paragraph{Evaluating Generalization Capabilities} If we take another dog / cat dataset, could a well trained model perform better than traditional optimization, since it isn't relying as much on memorization of noise or distribution specific data?

\paragraph{Evaluating Adversarial Robustness} Could the model looking only at the object to affect the classification score improve adversarial performance?

\paragraph{Mechanistic Interpretability Study} Does the better focus of the model encourage the creation of new image circuits? How does the model perform in cases when it contains neither cat nor dog? Can we make claims about the confidence of the model?

\paragraph{Non-Reduction of Feature Maps} One step of computing HiResCAMs involves summing each attention map to create a single activation map. Could the multiple foci be useful, and something we can further optimize?

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2024-08-02 src/model/overfit.py

@@ -18,7 +18,6 @@ if __name__ == '__main__':
     loss_fn = ContrastiveLoss(model.get_contrastive_cams)
 
     idx = random.randint(0, len(data) - 1)
-    print(idx)
     X = data[idx][0].unsqueeze(0)
     y = [x.unsqueeze(0) for x in data[idx][1]]
 
@@ -53,6 +52,3 @@ if __name__ == '__main__':
 
     plt.tight_layout()
     plt.show()
-
-    from IPython import embed
-    embed()

### 2024-08-04 src/model/loss.py
removed penalty for higher foreground activation from KLD

@@ -57,4 +57,9 @@ class ContrastiveLoss(nn.Module):
         cc = self.get_contrastive_cam(y[1], y_pred[1])
         heatmap = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
 
-        return self.kld(F.softmax((cc * 1E1).flatten(start_dim=2), dim=2).reshape(*cc.shape).log(), F.softmax((y[0] * 5E1).flatten(start_dim=1), dim=1).reshape(cc.shape[0], *cc.shape[-2:]).repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)) + 5E2 * cc[heatmap == 0].pow(2).mean() - cc[heatmap != 0].mean()
+        cc_log_probs = F.softmax((cc * 1E1).flatten(start_dim=2), dim=2).reshape(*cc.shape).log()
+        heatmap_probs = F.softmax(((y[0] + 5) * 5E1).flatten(start_dim=1), dim=1).reshape(cc.shape[0], *cc.shape[-2:]).repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+        heatmap_log_probs = heatmap_probs.log()
+        heatmap_log_probs[heatmap != 0] == 0
+
+        return (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0) + 50 * cc[heatmap == 0].pow(2).mean() - cc.mean()

### 2024-08-04 src/model/train.py

@@ -31,8 +31,6 @@ def fit(model, optimizer, loss, train, val):
             cse_loss = torch.empty(1, device=const.DEVICE)
 
             for train_batch, valid_batch in zip(train, val):
-                optimizer.zero_grad()
-
                 X_train, y_train, X_valid, y_valid = [*train_batch, *valid_batch]
                 y_pred_train = model(X_train)
                 y_pred_valid = model(X_valid)
@@ -47,6 +45,7 @@ def fit(model, optimizer, loss, train, val):
                 valid_batch_loss = loss(y_pred_valid, y_valid) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred_valid[0], y_valid[1])
                 valid_loss = torch.vstack([valid_loss, torch.tensor(valid_batch_loss)])
 
+                optimizer.zero_grad()
                 train_batch_loss.backward()
                 optimizer.step()
 

### 2024-08-05 paper/main.tex
updated model weights, plots and paper score report
explained D_KL modification on writeup

@@
 \begin{gather}
-	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+	\max_{\theta} s_c + b_c - s_{c'} - b_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} + b_c - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} - b_{c'}
 \end{gather}
@@
 If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.
+Ignoring the bias terms, our new optimization objective is:
+\begin{gather}
+	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+\end{gather}
 
@@
 	\end{cases} \\
+	D_{KL}(P || Q) := P \cdot \log \frac{P'}{Q} \text{ where }
+	P'_{i,j} := \begin{cases}
+		1 & P_{i,j} \neq 0 \\
+		P_{i,j} & \text{otherwise}
+	\end{cases} \\
 	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{D_1,D_2}_{i,j} \mathcal{\bm{A}}^{\text{foreground}}_{i,j}
@@
 \end{enumerate}
+
+$D_{KL}$ is modified such that the likelihood comparator sets point probability to 1 if it is part of the foreground.
+While this violates the probability distribution definition, it motivates the optimizer to maximize the foreground activation.
+During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification negates.
 
@@
 		Cross-Entropy & 0.628 & 68.2\% & 64.1\% \\
-		Interpretable (Ours) & \bf 0.354 & \bf 96.2\% & \bf 89.2\% \\
+		Interpretable (Ours) & \bf 0.327 & \bf 98.4\% & \bf 90.0\% \\
 		\bottomrule

### 2024-08-06 src/const.py
added gradient accumulation, checkpointing and multiclass run results

@@ -10,9 +10,12 @@ MODELS_DIR = BASE_DIR / 'models'
 
 # training
 DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
+TRAIN_CUTOFF = 12600  # max training time in seconds
+GRAD_ACCUMULATION_STEPS = 5
+CHECKPOINTING = True
 LEARNING_RATE = 1E-3
 SELECT_BEST = True
-BATCH_SIZE = 32
+BATCH_SIZE = 12
 EPOCHS = 150
 
 # dataset

### 2024-08-06 src/model/train.py

@@ -7,30 +7,33 @@ from .arch import Model
 from src import const
 import mlflow
 import torch
+import json
+import time
 import sys
 
 
-def fit(model, optimizer, loss, train, val):
+def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_id=None):
     if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
-    best = {'param': model.state_dict(),
-            'epoch': 0,
-            'acc': 0.0}
+    start_time = time.time()
+    best = best or {'param': model.state_dict(),
+                    'epoch': 0,
+                    'acc': 0.0}
 
-    with mlflow.start_run():
+    with mlflow.start_run(mlflow_run_id):
         # log hyperparameters
         mlflow.log_params({k: v for k, v in const.__dict__.items() if k == k.upper() and all(s not in k for s in ['DIR', 'PATH'])})
         mlflow.log_param('optimizer_fn', 'Adam')
 
         interval = max(1, (const.EPOCHS // 10))
-        for epoch in range(const.EPOCHS):
-            if not (epoch+1) % interval: print('-' * 10)
+        for epoch in range(init_epoch, const.EPOCHS + init_epoch):
+            if not (epoch) % interval: print('-' * 10)
             train_loss = torch.empty(1, device=const.DEVICE)
             valid_loss = torch.empty(1, device=const.DEVICE)
             train_acc = torch.empty(1, device=const.DEVICE)
             valid_acc = torch.empty(1, device=const.DEVICE)
             cse_loss = torch.empty(1, device=const.DEVICE)
 
-            for train_batch, valid_batch in zip(train, val):
+            for batch_idx, (train_batch, valid_batch) in enumerate(zip(train, val)):
                 X_train, y_train, X_valid, y_valid = [*train_batch, *valid_batch]
                 y_pred_train = model(X_train)
                 y_pred_valid = model(X_valid)
@@ -45,9 +48,11 @@ def fit(model, optimizer, loss, train, val):
                 valid_batch_loss = loss(y_pred_valid, y_valid) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred_valid[0], y_valid[1])
                 valid_loss = torch.vstack([valid_loss, torch.tensor(valid_batch_loss)])
 
-                optimizer.zero_grad()
                 train_batch_loss.backward()
[…]

### 2024-08-08 paper/main.tex
added theory for multiclass classification, fixed checkpoint recovery bug
committed from 30,000 ft :)

@@
 Therefore, comparing \textit{absolute score} is a misrepresentation of the internal model state.
+Instead, the objective is to maximize the \textbf{difference between class logits}.
 
@@
 \end{gather}
-This is equivalent to the original ERM optimization objective, with the difference being the resultant values, which is an activation map instead.
-If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.
 Ignoring the bias terms, our new optimization objective is:
 \begin{gather}
-	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+	\max_{\theta} \mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+\end{gather}
+This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map.
+If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.
+
+For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class.
+The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
+\begin{gather}
+	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
+\end{gather}
+This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.
+
+To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result.
+Therefore, the multiclass optimization objective under this paradigm is as follows:
+\begin{gather}
+	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{i}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
 \end{gather}
@@
 
-The distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information.
+As discussed in the previous section, the distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information.
 We know at a pointwise level how each feature contributes to the final prediction.
@@
 We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.
+The model through the process of training should learn to distinguish and subsequently ignore the background in determining the final prediction.
 
-Segmentations maps provide this exact information. It describes the region over which the object is present. We therefore define a constrastive cost function as follows:
+Segmentations maps describes the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
 \begin{gather}
-	\{(x, (h, y))\}^N \in \mathcal{D} \\
-	\mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{y}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{y'}^{\text{HiResCAM}} \\
-	\mathcal{\bm{A}}^{\text{background}}_{i,j} := \begin{cases}
-		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & h_{i,j} = 0 \\
+	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
+		1 & \text{pixel contains subject} \\
 		0 & \text{otherwise}
 	\end{cases} \\
-	\mathcal{\bm{A}}^{\text{foreground}}_{i,j} := \begin{cases}
+	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
+		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
+		0 & \text{otherwise}
+	\end{cases} \\
+	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
 		0 & h_{i,j} = 0 \\
-		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & \text{otherwise}
+		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
 	\end{cases} \\
-	D_{KL}(P || Q) := P \cdot \log \frac{P'}{Q} \text{ where }
+	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
 	P'_{i,j} := \begin{cases}
@@
 		P_{i,j} & \text{otherwise}
-	\end{cases} \\
-	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{D_1,D_2}_{i,j} \mathcal{\bm{A}}^{\text{foreground}}_{i,j}
+	\end{cases} \label{d_kl} \\
+	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
 \end{gather}
@@
 
-$D_{KL}$ is modified such that the likelihood comparator sets point probability to 1 if it is part of the foreground.
+$D_{KL}$ is modified per (10), such that the likelihood comparator sets point probability to 1 if it is part of the foreground.
 While this violates the probability distribution definition, it motivates the optimizer to maximize the foreground activation.
-During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification negates.
+During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification corrects.
+In addition, it also averages the divergence from the target segmenetation map to each contrastive activation map.
 
@@
 
-The Oxford-IIIT Pets dataset contains a classification goal alongside high-quality segmentation maps, and the binary classification case (image containing a dog or cat) was trained on.
-The architecture used for training was ResNet-50, initialized with random parameters and two key modifications:
+The Oxford-IIIT Pets dataset is an image dataset with a classification objective alongside high-quality segmentation maps.
+We train the model on the binary classification case (image containing a dog or cat) as well as the fine-grained, breed specific 37 class split.
+The architecture used for training was ResNet-50, initialized with random parameters and three key modifications:
 
@@
 
-The model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
-The total time for training was $\approx$1hr.
-
 \paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization}.
@@
 \section{Results}
+
+\subsection{Binary Classification}
+
+For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
+The total time for training was $\approx$1hr.
 
@@
 
+\subsection{Multiclass Classification}
+
+Our approach's performance on the Oxford-IIIT Pets Dataset to the default is as follows:
+
+%TODO: update results
+\begin{table}[h]
+	\centering
+	\begin{tabular}{c|ccc}
+		\toprule
+		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
+		\midrule
+		Cross-Entropy & 0.000 & 3.0\% & 3.0\% \\
+		Interpretable (Ours) & \bf 0.000 & \bf 92.0\% & \bf 50.0\% \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
 \section{Upcoming / In Progress Tasks}
 
-\paragraph{Extending to the Multiclass Setting} Currently, the implementation computes contrastive activation maps from the target to each other class.
-However, the best approach to optimization needs to be tested.
-The likely approach would be to compute the mean error over each contrastive activation map.
+\paragraph{Understanding the Generalization Gap} While the model for the multiclass paradigm is able to fit the training dataset excellently -- achieving $\approx 92\%$ training accuracy -- the validation accuracy is significantly behind at $\approx 50\%$.
+This could imply memorization of samples, which was something that the approach was intended to mitigate.
+Further investigation could help support or disprove this result.
 

### 2024-08-08 src/model/train.py

@@ -13,7 +13,6 @@ import sys
 
 
 def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_id=None):
-    if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
     start_time = time.time()
     best = best or {'param': model.state_dict(),
                     'epoch': 0,
@@ -52,7 +51,6 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
                 if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS:
                     optimizer.step()
                     optimizer.zero_grad()
-                    break
 
             train_acc = train_acc[1:]
             valid_acc = valid_acc[1:]
@@ -86,6 +84,7 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
 
 if __name__ == '__main__':
     const.MODEL_NAME = sys.argv[1]
+    if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
 
     model = Model(const.IMAGE_SHAPE, is_contrastive=const.MODEL_NAME != 'default')  # initialize before loss functions to ensure accurate cam size configuration
     train, val, test = get_generators()

### 2024-08-11 paper/main.tex
completed multiclass section, added multiclass model weights

@@
 
-Segmentations maps describes the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
+Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
 \begin{gather}
@@
 	\end{cases} \label{d_kl} \\
-	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
+	\begin{split}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 h)) \\
+		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 \\
+		&- \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
+	\end{split}
 \end{gather}
@@
 $D_{KL}$ is modified per (10), such that the likelihood comparator sets point probability to 1 if it is part of the foreground.
-While this violates the probability distribution definition, it motivates the optimizer to maximize the foreground activation.
+While this violates the probability distribution under which KL Divergence is defined, it motivates the optimizer to maximize the foreground activation, which is especially crucial for cases where the image contains large sections of foreground.
 During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification corrects.
-In addition, it also averages the divergence from the target segmenetation map to each contrastive activation map.
+In addition, it also averages the divergence from the target segmentation map to each contrastive activation map.
 
@@
 For our experiments, $\lambda_1 = 10, \lambda_2 = 50, \lambda_3 = 100$ were set empirically, without rigorous optimization.
+
+In addition to classifying the input image, the model returns a provably accurate attention map used to distinguish between the two classes.
+This can be obtained by simply computing the absolute values of the contrastive map compared against any two classes, and subsequently performing normalization.
 
@@
 
+For multiclass classification, the objective is to distinguish between the features of species within the cat and dog families.
+There are a total of 37 classes, with a success of rate of $\approx 2.7\%$ with random guessing.
+
 Our approach's performance on the Oxford-IIIT Pets Dataset to the default is as follows:
 
-%TODO: update results
 \begin{table}[h]
@@
 		\midrule
-		Cross-Entropy & 0.000 & 3.0\% & 3.0\% \\
-		Interpretable (Ours) & \bf 0.000 & \bf 92.0\% & \bf 50.0\% \\
+		Cross-Entropy & 3.627 & 2.8\% & 3.1\% \\
+		Interpretable (Ours) & \bf 2.743 & \bf 93.0\% & \bf 51.5\% \\
 		\bottomrule
@@
 
+We can observe that the default model is unable to learn any valuable information, the contrastive model is able to learn a significant portion of the information, although with a substantial generalization gap.
+
+The model was trained for 731 epochs with the Adam optimizer with a learning rate of $10^{-3}$ with a batch size of 12, with 5 gradient accumulation steps (effective batch size of 60) on a 32GB V100 on a randomly shuffled dataset using the original splits.
+The default model was only trained for 30 epochs, by which point it had already converged.
+For posterity, we will run the default setup for 731 epochs as well before publication.
+
 \section{Upcoming / In Progress Tasks}
 
-\paragraph{Understanding the Generalization Gap} While the model for the multiclass paradigm is able to fit the training dataset excellently -- achieving $\approx 92\%$ training accuracy -- the validation accuracy is significantly behind at $\approx 50\%$.
-This could imply memorization of samples, which was something that the approach was intended to mitigate.
-Further investigation could help support or disprove this result.
+\paragraph{Understanding the Generalization Gap} While the model for the multiclass paradigm is able to fit the training dataset excellently -- achieving $\approx 93\%$ training accuracy -- the validation accuracy is significantly behind at $\approx 51\%$.
+This could imply memorization of samples, which was something that this approach was intended to mitigate.
+Further investigation could help support or disprove this hypothesis.
 

### 2024-08-16 src/data/pet_image.py
added accuracy eval, filtered dataset

@@ -18,7 +18,7 @@ class Dataset(torch.utils.data.Dataset):
         X = X / 255  # normalization
 
         y = torch.zeros((2))
-        y[int('cat' in self.images[idx])] = 1
+        y[int('dog' in self.images[idx])] = 1
 
         return X.to(const.DEVICE), y.to(const.DEVICE)
 

### 2024-08-16 src/model/evaluate.py

@@ -1,6 +1,7 @@
 #!/usr/bin/env python3
 
-from ..data.oxford_iiit_pet import Dataset
+from ..data.oxford_iiit_pet import Dataset as oxford_iiit_pets
+from ..data.pet_image import Dataset as pet_image
 import torch.nn.functional as F
 import matplotlib.pyplot as plt
 from .arch import Model
@@ -34,6 +35,17 @@ def visualize(model, gen):
     plt.show()
 
 
+def accuracy(model, gen, samples=1000):
+    score = 0
+
+    for sample in random.sample(range(len(gen)), samples):
+        X, y = gen[sample]
+        y_pred, cam = model(X.unsqueeze(0))
+        score += (y.argmax(dim=0) == y_pred.argmax(dim=1)).sum()
+
+    return score / samples
+
+
 if __name__ == '__main__':
     name = sys.argv[1]
     random.seed(const.SEED)
@@ -43,4 +55,5 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    visualize(model, Dataset('test'))
+    if sys.argv[2] == 'visualize': visualize(model, oxford_iiit_pets('test'))
+    else: print(accuracy(model, pet_image()))

### 2024-08-17 paper/main.tex
added BCE based loss, updated paper with OOD generalization result, updated hyperparameter

@@
 
+\subsection{Out-of-Distribution Generalization}
+
+To test how well the approach generalizes out of the training distribution, we evaluate the performance of the binary classification contrastive model on random samples ($n=1000$) from the \href{https://www.microsoft.com/en-us/download/details.aspx?id=54765}{Kaggle Cats and Dogs Dataset}:
+\begin{table}[h]
+	\centering
+	\begin{tabular}{c|c}
+		\toprule
+		\textbf{Method} & \textbf{Accuracy} \\
+		\midrule
+		Cross-Entropy & 49.5\% \\
+		Interpretable (Ours) & \bf 78.5\% \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+The next goal is to further understand and minimize this generalization gap.
+
 \section{Upcoming / In Progress Tasks}
@@
 Further investigation could help support or disprove this hypothesis.
-
-\paragraph{Evaluating Generalization Capabilities} If we take another dog / cat dataset, could a well trained model perform better than traditional optimization, since it isn't relying as much on memorization of noise or distribution specific data?
 

### 2024-08-17 src/const.py

@@ -11,7 +11,7 @@ MODELS_DIR = BASE_DIR / 'models'
 # training
 DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
-LAMBDAS = [1E1, 5E1, 1, 5E4, 1]
+LAMBDAS = [1E1, 5E1, 1, 5E1, 1]
 GRAD_ACCUMULATION_STEPS = 5
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3

### 2024-08-17 src/model/loss.py

@@ -6,6 +6,38 @@ from torch import nn
 import torch
 
 
+class SoftmaxKLDLoss(nn.Module):
+    def __init__(self, get_contrastive_cam_fn):
+        super().__init__()
+        self.get_contrastive_cam = get_contrastive_cam_fn
+        self.kld = nn.KLDivLoss(reduction='batchmean')
+
+    def forward(self, y_pred, y, debug=False):
+        cc = self.get_contrastive_cam(y[1], y_pred[1])
+        heatmap = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+
+        cc_log_probs = F.softmax((cc * const.LAMBDAS[0]).flatten(start_dim=2), dim=2).reshape(*cc.shape).log()
+        heatmap_probs = F.softmax(((y[0] + 5) * const.LAMBDAS[1]).flatten(start_dim=1), dim=1).reshape(cc.shape[0], *cc.shape[-2:]).repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+        heatmap_log_probs = heatmap_probs.log()
+        heatmap_log_probs[heatmap != 0] == 0
+
+        return const.LAMBDAS[2] * (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0) + const.LAMBDAS[3] * cc[heatmap == 0].pow(2).mean() - const.LAMBDAS[4] * cc.mean()
+
+
+class ContrastiveLoss(nn.Module):
+    def __init__(self, get_contrastive_cam_fn):
+        super().__init__()
+        self.get_contrastive_cam = get_contrastive_cam_fn
+
+    def forward(self, y_pred, y):
+        cc = self.get_contrastive_cam(y[1], y_pred[1])
+        heatmap = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+        heatmap[heatmap < .3] = .5
+        heatmap[heatmap >= 1] = 1.  # glitching torch, my theory -> some values are 1.0000 and extra 0's are counted as extending the [0,1] range
+
+        return F.binary_cross_entropy(F.sigmoid(cc), heatmap) + const.LAMBDAS[3] * cc[heatmap == .5].pow(2).mean() - const.LAMBDAS[4] * cc.mean()
+
+
 class PieceWiseLoss(nn.Module):
[…]

### 2024-08-17 src/model/loss.py
set all low focus (<.5) pixels to .5 for sigmoid target

@@ -32,7 +32,7 @@ class ContrastiveLoss(nn.Module):
     def forward(self, y_pred, y):
         cc = self.get_contrastive_cam(y[1], y_pred[1])
         heatmap = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
-        heatmap[heatmap < .3] = .5
+        heatmap[heatmap < .5] = .5
         heatmap[heatmap >= 1] = 1.  # glitching torch, my theory -> some values are 1.0000 and extra 0's are counted as extending the [0,1] range
 
         return F.binary_cross_entropy(F.sigmoid(cc), heatmap) + const.LAMBDAS[3] * cc[heatmap == .5].pow(2).mean() - const.LAMBDAS[4] * cc.mean()

### 2024-08-18 paper/main.tex
corrected some notation in paper, reverted to softmax-KLD based loss

@@
 \begin{gather}
-	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
+	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
 \end{gather}
@@
 		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 h)) \\
-		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 \\
+		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
 		&- \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}

### 2024-08-18 src/model/loss.py

@@ -6,7 +6,7 @@ from torch import nn
 import torch
 
 
-class SoftmaxKLDLoss(nn.Module):
+class ContrastiveLoss(nn.Module):
     def __init__(self, get_contrastive_cam_fn):
         super().__init__()
         self.get_contrastive_cam = get_contrastive_cam_fn
@@ -24,7 +24,7 @@ class SoftmaxKLDLoss(nn.Module):
         return const.LAMBDAS[2] * (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0) + const.LAMBDAS[3] * cc[heatmap == 0].pow(2).mean() - const.LAMBDAS[4] * cc.mean()
 
 
-class ContrastiveLoss(nn.Module):
+class BCELoss(nn.Module):
     def __init__(self, get_contrastive_cam_fn):
         super().__init__()
         self.get_contrastive_cam = get_contrastive_cam_fn

### 2024-08-18 src/const.py
added SGD trained models, added specific subsection logging
clean logging implementation desperately required

@@ -12,11 +12,11 @@ MODELS_DIR = BASE_DIR / 'models'
 DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
 LAMBDAS = [1E1, 5E1, 1, 5E1, 1]
-GRAD_ACCUMULATION_STEPS = 5
+GRAD_ACCUMULATION_STEPS = 1
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3
 SELECT_BEST = True
-BATCH_SIZE = 12
+BATCH_SIZE = 32
 EPOCHS = 150
 
 # dataset

### 2024-08-18 src/model/loss.py

@@ -21,7 +21,12 @@ class ContrastiveLoss(nn.Module):
         heatmap_log_probs = heatmap_probs.log()
         heatmap_log_probs[heatmap != 0] == 0
 
-        return const.LAMBDAS[2] * (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0) + const.LAMBDAS[3] * cc[heatmap == 0].pow(2).mean() - const.LAMBDAS[4] * cc.mean()
+        kld = (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0)
+        background = cc[heatmap == 0].pow(2).mean()
+        foreground = cc.mean()
+
+        self.prev = (kld, background, foreground)
+        return const.LAMBDAS[2] * kld + const.LAMBDAS[3] * background - const.LAMBDAS[4] * foreground
 
 
 class BCELoss(nn.Module):

### 2024-08-18 src/model/train.py

@@ -26,14 +26,19 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
             if not (epoch) % interval: print('-' * 10)
-            train_loss = torch.empty(1, device=const.DEVICE)
-            valid_loss = torch.empty(1, device=const.DEVICE)
+            train_contrast_loss = torch.empty(1, device=const.DEVICE)
+            train_kld_loss = torch.empty(1, device=const.DEVICE)
+            train_background_loss = torch.empty(1, device=const.DEVICE)
+            train_foreground_loss = torch.empty(1, device=const.DEVICE)
+            valid_contrast_loss = torch.empty(1, device=const.DEVICE)
+            valid_kld_loss = torch.empty(1, device=const.DEVICE)
+            valid_background_loss = torch.empty(1, device=const.DEVICE)
+            valid_foreground_loss = torch.empty(1, device=const.DEVICE)
             train_acc = torch.empty(1, device=const.DEVICE)
             valid_acc = torch.empty(1, device=const.DEVICE)
             cse_loss = torch.empty(1, device=const.DEVICE)
 
-            for batch_idx, (train_batch, valid_batch) in enumerate(zip(train, val)):
-                X_train, y_train, X_valid, y_valid = [*train_batch, *valid_batch]
+            for batch_idx, ((X_train, y_train), (X_valid, y_valid)) in enumerate(zip(train, val)):
                 y_pred_train = model(X_train)
                 y_pred_valid = model(X_valid)
 
@@ -41,11 +46,17 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
                 valid_acc = torch.vstack([valid_acc, (torch.argmax(y_valid[1], dim=1) == torch.argmax(y_pred_valid[0], dim=1)).unsqueeze(1)])
 
                 train_batch_loss = loss(y_pred_train, y_train) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred_train[0], y_train[1])
-                train_loss = torch.vstack([train_loss, torch.tensor(train_batch_loss)])
+                train_contrast_loss = torch.vstack([train_contrast_loss, torch.tensor(train_batch_loss)])
[…]

### 2024-08-18 src/model/train.py
offloaded logging from gpu

@@ -26,37 +26,39 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
             if not (epoch) % interval: print('-' * 10)
-            train_contrast_loss = torch.empty(1, device=const.DEVICE)
-            train_kld_loss = torch.empty(1, device=const.DEVICE)
-            train_background_loss = torch.empty(1, device=const.DEVICE)
-            train_foreground_loss = torch.empty(1, device=const.DEVICE)
-            valid_contrast_loss = torch.empty(1, device=const.DEVICE)
-            valid_kld_loss = torch.empty(1, device=const.DEVICE)
-            valid_background_loss = torch.empty(1, device=const.DEVICE)
-            valid_foreground_loss = torch.empty(1, device=const.DEVICE)
-            train_acc = torch.empty(1, device=const.DEVICE)
-            valid_acc = torch.empty(1, device=const.DEVICE)
-            cse_loss = torch.empty(1, device=const.DEVICE)
+            train_contrast_loss = torch.empty(1, device='cpu')
+            train_kld_loss = torch.empty(1, device='cpu')
+            train_background_loss = torch.empty(1, device='cpu')
+            train_foreground_loss = torch.empty(1, device='cpu')
+            valid_contrast_loss = torch.empty(1, device='cpu')
+            valid_kld_loss = torch.empty(1, device='cpu')
+            valid_background_loss = torch.empty(1, device='cpu')
+            valid_foreground_loss = torch.empty(1, device='cpu')
+            train_acc = torch.empty(1, device='cpu')
+            valid_acc = torch.empty(1, device='cpu')
+            cse_loss = torch.empty(1, device='cpu')
 
             for batch_idx, ((X_train, y_train), (X_valid, y_valid)) in enumerate(zip(train, val)):
                 y_pred_train = model(X_train)
                 y_pred_valid = model(X_valid)
 
-                train_acc = torch.vstack([train_acc, (torch.argmax(y_train[1], dim=1) == torch.argmax(y_pred_train[0], dim=1)).unsqueeze(1)])
-                valid_acc = torch.vstack([valid_acc, (torch.argmax(y_valid[1], dim=1) == torch.argmax(y_pred_valid[0], dim=1)).unsqueeze(1)])
[…]

### 2024-08-19 src/const.py
minor memory optimizations

@@ -23,6 +23,7 @@ EPOCHS = 150
 SEED = 1024
 SPLITS = ['train', 'valid', 'test']
 IMAGE_SIZE = (224, 224)
+CAM_SIZE = (14, 14)
 N_CHANNELS = 3
 BINARY_CLS = True
 N_CLASSES = 2 if BINARY_CLS else 37

### 2024-08-19 src/model/arch.py

@@ -18,18 +18,13 @@ class Model(torch.nn.Module):
             self.backbone.layer4[-1].bn3 = nn.Identity()
             self.backbone.layer4[-1].relu = nn.Identity()
         self.backbone.layer4[-1].conv3.register_forward_hook(self._hook)
-
         self.backbone.fc = nn.Identity()
-
-        self.feature_grad = None
-        self.feature_rect = None
 
         self.linear = nn.LazyLinear(const.N_CLASSES, bias=not is_contrastive)
         self.softmax = nn.Softmax(dim=1)  # ~equivalent to sigmoid since classes = 2; relevant for CAMs
 
         self.to(const.DEVICE)
         self(torch.randn(1, *input_shape).to(const.DEVICE))  # initialization
-        const.CAM_SIZE = tuple(self.feature_rect.shape[2:])
 
     def _hook(self, model, i, o):
         def assign(grad):
@@ -57,6 +52,10 @@ class Model(torch.nn.Module):
             for class_idx in range(logits.shape[1]):
                 logits[img_idx, class_idx].backward(retain_graph=True, inputs=self.backbone.layer4[-1].conv3.weight)
                 cams[img_idx, class_idx] = (self.feature_rect * self.feature_grad).sum(dim=1)[img_idx]
+
+        self.feature_grad = None
+        self.feature_rect = None
+
         return cams
 
 

### 2024-08-19 paper/main.tex
corrected eq 11, added pipeline args
formalized logit-difference influence on class probabilities

@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{float}
 \usepackage{bm}
+
+\newtheorem{theorem}{Proposition}
+\newtheorem{corollary}{Corollary}[theorem]
 
@@
 
-We can consider the backbone as a kernel \citep{jacot2018neural} that fundamentally attempts to linearly separate the correct class, and the softmax-activated final layer being logistic regression.
+We can consider the backbone as a kernel that attempts to linearly separate the correct class, with the softmax-activated final layer just being logistic regression over the final feature map.
 
@@
 It assigns a score for it's specific value, which is independent since one set of weights does not affect the other over a forward pass.
-If we take as example logits $[1,2]$ and $[-2,-1]$, we see that $\sigma([1,2]) = \sigma([-2,1]) = [0.2689, 0.7311]$.
+A consequence of performing softmax over the final logits is that the normalization steps renders only the difference between each score $s_c$ relevant for the final prediction probability.
+
+\begin{theorem}
+	Individual class probabilites $\sigma_c(\bm{s})$ depend on the difference(s) between class logits.
+\end{theorem}
+\begin{proof} Individual class probabilities for logit vector $\bm{s}$ are defined as:
+	\begin{gather}
+		\sigma_c(\bm{s}) = \frac{e^{s_c}}{\sum_i e^{s_i}}
+	\end{gather}
+	We define our logit vector in terms of the elementwise difference to a target class $c$:
+	\begin{gather}
+		d = [s_i - s_c]_i \\
+		\bm{s} = [s_c + d_i]_i
+	\end{gather}
+	Based on this definition, class probabilities can now be computed as:
+	\begin{align}
+		\sigma_c(\bm{s}) &= \frac{e^{s_c}}{\sum_i e^{s_c+d_i}} \\
+				 &= \frac{e^{s_c}}{e^{s_c} \sum_i e^{d_i}} = \frac{1}{\sum_i e^{d_i}}
+	\end{align}
+	This recontextualizes softmax as a direct function of the differences of class logits.
+\end{proof}
+\begin{corollary} Since class probabilities depend on logit differences, the class probabilities are identitical for any two pairs of logit vectors with identical logit differences.
+	\begin{gather}
+		\frac{e^{s_c}}{\sum_i e^{s_i}} = \frac{e^{s_c+r}}{\sum_i e^{s_i+r}} \forall r \in \mathbb{R}
+	\end{gather}
+\end{corollary}
+\begin{proof} Recontextualizing $\bm{s}$ per the previous proposition, we get:
+	\begin{align}
+		\frac{e^{s_c}}{\sum_i e^{s_c + d_i}} &= \frac{e^{s_c+r}}{\sum_i e^{s_c+d_i+r}} \\
+		\frac{e^{s_c}}{e^{s_c}\sum_i e^{d_i}} &= \frac{e^{s_c+r}}{e^{s_c + r} \sum_i e^{d_i}} \\
+		\frac{1}{\sum_i e^{d_i}} &= \frac{1}{\sum_i e^{d_i}}
+	\end{align}
+	Therefore, the absolute scale of a logit does not influence the final class probability.
+\end{proof}
+
+As an example, for logits $[1,2]$ and $[-2,-1]$, $\sigma([1,2]) = \sigma([-2,1]) = [0.2689, 0.7311]$.
 The prediction probabilities remain the same, despite different values for the same score $s_c$.
-Therefore, comparing \textit{absolute score} is a misrepresentation of the internal model state.
+Therefore, making inferences based on \textit{absolute score} results in a misrepresentation of the internal model state.
 Instead, the objective is to maximize the \textbf{difference between class logits}.
@@
 	P'_{i,j} := \begin{cases}
-		1 & P_{i,j} \neq 0 \\
+		1 & h_{i,j} \neq 0 \\
 		P_{i,j} & \text{otherwise}

### 2024-08-21 src/const.py
offloaded logging for memory optimizations; fixed leaky memory, improved writeup notation

@@ -16,7 +16,7 @@ GRAD_ACCUMULATION_STEPS = 1
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3
 SELECT_BEST = True
-BATCH_SIZE = 32
+BATCH_SIZE = 44
 EPOCHS = 150
 
 # dataset

### 2024-08-21 src/model/loss.py

@@ -25,7 +25,7 @@ class ContrastiveLoss(nn.Module):
         background = cc[heatmap == 0].pow(2).mean()
         foreground = cc.mean()
 
-        self.prev = (kld, background, foreground)
+        self.prev = (kld.item(), background.item(), foreground.item())
         return const.LAMBDAS[2] * kld + const.LAMBDAS[3] * background - const.LAMBDAS[4] * foreground
 
 

### 2024-08-21 src/model/train.py

@@ -5,6 +5,7 @@ from .loss import ContrastiveLoss
 import torch.nn.functional as F
 from .arch import Model
 from src import const
+import numpy as np
 import mlflow
 import torch
 import json
@@ -26,67 +27,36 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
             if not (epoch) % interval: print('-' * 10)
-            train_contrast_loss = torch.empty(1, device='cpu')
-            train_kld_loss = torch.empty(1, device='cpu')
-            train_background_loss = torch.empty(1, device='cpu')
-            train_foreground_loss = torch.empty(1, device='cpu')
-            valid_contrast_loss = torch.empty(1, device='cpu')
-            valid_kld_loss = torch.empty(1, device='cpu')
-            valid_background_loss = torch.empty(1, device='cpu')
-            valid_foreground_loss = torch.empty(1, device='cpu')
-            train_acc = torch.empty(1, device='cpu')
-            valid_acc = torch.empty(1, device='cpu')
-            cse_loss = torch.empty(1, device='cpu')
+            metrics = {metric: [] for metric in ['train_contrast_loss', 'train_kld_loss', 'train_background_loss', 'train_foreground_loss', 'train_acc', 'cse_loss',
+                                                 'valid_contrast_loss', 'valid_kld_loss', 'valid_background_loss', 'valid_foreground_loss', 'valid_acc']}
 
             for batch_idx, ((X_train, y_train), (X_valid, y_valid)) in enumerate(zip(train, val)):
                 y_pred_train = model(X_train)
                 y_pred_valid = model(X_valid)
 
-                train_acc = torch.vstack([train_acc, (torch.argmax(y_train[1], dim=1) == torch.argmax(y_pred_train[0], dim=1)).unsqueeze(1).to('cpu')])
-                valid_acc = torch.vstack([valid_acc, (torch.argmax(y_valid[1], dim=1) == torch.argmax(y_pred_valid[0], dim=1)).unsqueeze(1).to('cpu')])
+                metrics['train_acc'].extend((torch.argmax(y_train[1], dim=1) == torch.argmax(y_pred_train[0], dim=1)).unsqueeze(1).tolist())
[…]

### 2024-08-22 paper/main.tex
offloaded logging for memory optimizations; fixed leaky memory, improved writeup notation
updated plots, evals, added model weights, moved code to SGD for evaluation

@@
 
-\title{Interpretable Risk Minimization -- Preliminaries}
+\title{Contrastive Optimization -- Preliminaries}
 
@@
 \begin{gather}
-	\sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
-	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_1}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
+	\sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
+	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
 \end{gather}
@@
 \begin{gather}
-	\max_{\theta} s_c + b_c - s_{c'} - b_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} + b_c - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} - b_{c'}
+	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
 \end{gather}
@@
 \begin{gather}
-	\max_{\theta} \mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+	\max_{\theta} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}}
 \end{gather}
@@
 \begin{gather}
-	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
+	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
 \end{gather}
@@
 \begin{gather}
-	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{i}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
+	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
 \end{gather}
@@
 	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
-		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
+		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
 		0 & \text{otherwise}
@@
 		0 & h_{i,j} = 0 \\
-		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
+		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
 	\end{cases} \\
@@
 
-For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
+For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $44$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
 The total time for training was $\approx$1hr.
@@
 		\midrule
-		Cross-Entropy & 0.628 & 68.2\% & 64.1\% \\
-		Interpretable (Ours) & \bf 0.327 & \bf 98.4\% & \bf 90.0\% \\
+		Cross-Entropy & 0.646 & 66.1\% & 64.5\% \\
+		Interpretable (Ours) & \bf 0.326 & \bf 98.6\% & \bf 92.3\% \\
 		\bottomrule
@@
 
+\vspace{-2em}
 \begin{figure}[H]
 	\centering
-	\includegraphics[width=.8\textwidth]{img/plots.png}
+	\includegraphics[width=.4\textwidth]{img/accuracy.png}
+	\hspace{4em}
+	\includegraphics[width=.4\textwidth]{img/loss.png}
 	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy.
-The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_dd3fe52afeca42539f87d5da6ac4ea3f\%22,\%22m_e21074d967b741dfb772746b113fe761\%22]}{here}.
-The orange line represent cross-entropy loss, while the blue line represents the contrastive approach.}
+The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_f4f7dea52fa74739be788e5f47da1195\%22,\%22m_1301bf1ba9bf4db48e09e7497c8cd214\%22]}{here}.
+The blue line represents cross-entropy loss, while the orange line represents the contrastive approach.}
 \end{figure}
@@
 	\includegraphics[width=.4\textwidth]{img/default_cam.png}
-	\hspace{4em}
+	\hspace{3em}
 	\includegraphics[width=.4\textwidth]{img/contrastive_cam.png}
@@
 There are a total of 37 classes, with a success of rate of $\approx 2.7\%$ with random guessing.
+There is virtually no class imbalance in this case.
 

### 2024-08-22 src/const.py

@@ -17,6 +17,7 @@ CHECKPOINTING = True
 LEARNING_RATE = 1E-3
 SELECT_BEST = True
 BATCH_SIZE = 44
+MOMENTUM = 0.9
 EPOCHS = 150
 
 # dataset

### 2024-08-22 src/model/evaluate.py

@@ -2,6 +2,7 @@
 
 from ..data.oxford_iiit_pet import Dataset as oxford_iiit_pets
 from ..data.pet_image import Dataset as pet_image
+from matplotlib.colors import Normalize
 import torch.nn.functional as F
 import matplotlib.pyplot as plt
 from .arch import Model
@@ -17,7 +18,7 @@ def kernel_diff(model):
     plt.show()
 
 
-def visualize(model, gen):
+def visualize(model, gen, norm=None):
     fig = plt.figure(figsize=(14, 14),
                      facecolor='white')
 
@@ -28,7 +29,7 @@ def visualize(model, gen):
         fig.add_subplot(4, 4, idx + 1)
         plt.xlabel(f'Pred: {str(y_pred[0].argmax().item())}, Actual: {str(y.argmax().item())}')
         plt.imshow(X.permute(1, 2, 0).cpu().detach(), alpha=0.5)
-        plt.imshow(F.interpolate(cam, const.IMAGE_SIZE, mode='bilinear')[0, 0].cpu(), cmap='jet', alpha=0.5)
+        plt.imshow(F.interpolate(cam, const.IMAGE_SIZE, mode='bilinear')[0, 0].cpu(), cmap='jet', alpha=0.5, norm=norm)
 
     plt.tight_layout()
     fig.savefig(const.DATA_DIR / 'evals' / f'{model.name}_cam.png')
@@ -56,4 +57,5 @@ if __name__ == '__main__':
     model.eval()
 
     if sys.argv[2] == 'visualize': visualize(model, oxford_iiit_pets('test'))
+    elif sys.argv[2] == 'visualize-normed': visualize(model, oxford_iiit_pets('test'), Normalize(vmin=-.5, vmax=.5) )
     else: print(accuracy(model, pet_image()))

### 2024-08-22 src/model/train.py

@@ -22,7 +22,7 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
     with mlflow.start_run(mlflow_run_id):
         # log hyperparameters
         mlflow.log_params({k: v for k, v in const.__dict__.items() if k == k.upper() and all(s not in k for s in ['DIR', 'PATH'])})
-        mlflow.log_param('optimizer_fn', 'Adam')
+        mlflow.log_param('optimizer_fn', 'SGD')
 
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
@@ -83,8 +83,9 @@ if __name__ == '__main__':
 
     model = Model(const.IMAGE_SHAPE, is_contrastive=const.MODEL_NAME != 'default')  # initialize before loss functions to ensure accurate cam size configuration
     train, val, test = get_generators()
-    optimizer = torch.optim.Adam(model.parameters(),
-                                 lr=const.LEARNING_RATE)
+    optimizer = torch.optim.SGD(model.parameters(),
+                                lr=const.LEARNING_RATE,
+                                momentum=const.MOMENTUM)
     loss = ContrastiveLoss(model.get_contrastive_cams) if const.MODEL_NAME != 'default' else torch.nn.CrossEntropyLoss()
 
     checkpoint_args = {'init_epoch': 1,

### 2024-08-27 paper/main.tex
fixed needlessly adversarial terms from loss, corrected loss formulation in writeup

@@
 \usepackage{float}
+\usepackage{bbm}
 \usepackage{bm}
@@
 \begin{gather}
-	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
+	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{contrastive}}\ \forall i \in \mathbb{Z}_{+}(|c| - 1)
 \end{gather}
@@
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 h)) \\
-		&+ \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
-		&- \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_3 h)) \\
+		&+ \frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
+		&- \frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
 	\end{split}
@@
 
-$\lambda_1$ \& $\lambda_2$ are hyperparameters that scale the activation maps, since softmax computer over highly similar values close to zero produces uniform probability distributions.
-$\lambda_3$ weights the importance of the background not affecting the final prediction.
-For our experiments, $\lambda_1 = 10, \lambda_2 = 50, \lambda_3 = 100$ were set empirically, without rigorous optimization.
+$\lambda_2$ \& $\lambda_3$ are hyperparameters that scale the activation maps, since softmax computed over values with small differences produce largely uniform probability distributions.
+$\lambda_1$ weighs the importance of the divergence, $\lambda_4$ weights the importance of the background not affecting the final prediction \& $\lambda_5$ weighs the importance of the foreground being positive.
+For our experiments, $\bm{\lambda} = \{1, 50, 1, 50, 1\}$ were set for $i = \{1, \ldots, 5\}$ respectively.
+These were identified empirically, without rigorous optimization.
 

### 2024-08-27 src/model/loss.py

@@ -23,7 +23,7 @@ class ContrastiveLoss(nn.Module):
 
         kld = (heatmap_probs * (heatmap_log_probs - cc_log_probs)).sum() / cc.size(0)
         background = cc[heatmap == 0].pow(2).mean()
-        foreground = cc.mean()
+        foreground = cc[heatmap == 1].mean()
 
         self.prev = (kld.item(), background.item(), foreground.item())
         return const.LAMBDAS[2] * kld + const.LAMBDAS[3] * background - const.LAMBDAS[4] * foreground

### 2024-08-27 src/model/train.py

@@ -32,10 +32,9 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
             for split, dataloader in zip(const.SPLITS[:2], (train, val)):
                 for batch_idx, (X, y) in enumerate(dataloader):
                     y_pred = model(X)
-
-                    metrics[f'{split}_acc'].extend((torch.argmax(y[1], dim=1) == torch.argmax(y_pred[0], dim=1)).unsqueeze(1).tolist())
-
                     batch_loss = loss(y_pred, y) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred[0], y[1])
+
+                    metrics[f'{split}_acc'].extend(y[1].argmax(1).eq(y_pred[0].argmax(1)).unsqueeze(1).tolist())
                     metrics[f'{split}_contrast_loss'].append(batch_loss.item())
                     metrics[f'{split}_cse_loss'].append(F.cross_entropy(y_pred[0], y[1]).item())
 

### 2024-08-27 src/model/train.py
split zero grad & optimizer step
following https://github.com/kuangliu/pytorch-cifar

@@ -31,6 +31,8 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
 
             for split, dataloader in zip(const.SPLITS[:2], (train, val)):
                 for batch_idx, (X, y) in enumerate(dataloader):
+                    if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.zero_grad()
+
                     y_pred = model(X)
                     batch_loss = loss(y_pred, y) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred[0], y[1])
 
@@ -50,9 +52,7 @@ def fit(model, optimizer, loss, train, val, best=None, init_epoch=1, mlflow_run_
                         batch_loss.backward()
                         mlflow.log_metric(f'{split}_batchwise_loss', batch_loss.item(), step=(epoch-1) * len(dataloader) + batch_idx)
 
-                        if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS:
-                            optimizer.step()
-                            optimizer.zero_grad()
+                        if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.step()
 
             metrics = {metric: np.mean(metrics[metric]) for metric in metrics}
             mlflow.log_metrics(metrics, step=epoch-1)

### 2024-09-11 src/model/train.py
added cosine annealing, weight decay, consistency changes

@@ -13,7 +13,9 @@ import time
 import sys
 
 
-def fit(model, optimizer, loss, train, val, selected=None, init_epoch=1, mlflow_run_id=None):
+def fit(model, optimizer, scheduler, criterion, train, val,
+        selected=None, init_epoch=1, mlflow_run_id=None):
+    model.train()
     start_time = time.time()
     selected = selected or {'last': model.state_dict(),
                             'epoch': init_epoch,
@@ -34,7 +36,7 @@ def fit(model, optimizer, loss, train, val, selected=None, init_epoch=1, mlflow_
                     if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.zero_grad()
 
                     y_pred = model(X)
-                    batch_loss = loss(y_pred, y) if loss._get_name() != 'CrossEntropyLoss' else loss(y_pred[0], y[1])
+                    batch_loss = criterion(y_pred, y) if criterion._get_name() != 'CrossEntropyLoss' else criterion(y_pred[0], y[1])
 
                     metrics[f'{split}_acc'].extend(y[1].argmax(1).eq(y_pred[0].argmax(1)).unsqueeze(1).tolist())
                     metrics[f'{split}_contrast_loss'].append(batch_loss.item())
@@ -43,16 +45,17 @@ def fit(model, optimizer, loss, train, val, selected=None, init_epoch=1, mlflow_
                     del y_pred, X, y
                     torch.cuda.empty_cache()
 
-                    if loss._get_name() != 'CrossEntropyLoss':
-                        metrics[f'{split}_kld_loss'].append(loss.prev[0])
-                        metrics[f'{split}_background_loss'].append(loss.prev[1])
-                        metrics[f'{split}_foreground_loss'].append(loss.prev[2])
+                    if criterion._get_name() != 'CrossEntropyLoss':
+                        metrics[f'{split}_kld_loss'].append(criterion.prev[0])
+                        metrics[f'{split}_background_loss'].append(criterion.prev[1])
+                        metrics[f'{split}_foreground_loss'].append(criterion.prev[2])
 
                     if split == 'train':
                         batch_loss.backward()
                         mlflow.log_metric(f'{split}_batchwise_loss', batch_loss.item(), step=(epoch-1) * len(dataloader) + batch_idx)
 
[…]

### 2024-09-11 paper/main.tex
corrected bug in math

@@
 \begin{gather}
-	\max_{\theta} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}}
+	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
 \end{gather}
@@
 \begin{gather}
-	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_t'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
+	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
 \end{gather}
@@
 \begin{gather}
-	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{contrastive}}\ \forall i \in \mathbb{Z}_{+}(|c| - 1)
+	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|c| - 1)
 \end{gather}
@@
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_3 h)) \\
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}})) \\
 		&+ \frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\

### 2024-09-27 paper/main.tex
added ablated cross entropy formulation to paper, normalized fg/bg split influence plots to full heatmap

@@
 
-\begin{theorem}
+\begin{theorem}\label{classprobdiff}
 	Individual class probabilites $\sigma_c(\bm{s})$ depend on the difference(s) between class logits.
@@
 
+\subsubsection{Divergence Formulation}
 Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
@@
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}})) \\
-		&+ \frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F \\
-		&- \frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}}))}_{\text{minimize divergence from target shape}} \\
+		&+ \underbrace{\frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F}_{\text{minimize background influence}} \\
+		&- \underbrace{\frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}}_{\text{maximize foreground influence}}
 	\end{split}
 \end{gather}
-
-The cost function has a threefold purpose:
-\begin{enumerate}
-	\item Minimize the divergence of the contrastive activation map to the segmentation map.
-	\item Minimize the absolute value of the contrastive map where the segmentation map is the background.
-	\item Maximize the value of the contrastive map where the segmentation map is the foreground.
-\end{enumerate}
 
@@
 These were identified empirically, without rigorous optimization.
+
+\subsubsection{Ablated Cross-Entropy}
+
+A more mathematically rigorous approach involves performing an ablation over class logits to penalize background influence.
+
+\begin{theorem} We can integrate background invariance to the definition of cross-entropy using the following formulation:
+	\begin{gather}
+		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+	\end{gather}
+\end{theorem}
+\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from \ref{classprobdiff}, we have:
+	\begin{align}
+		d &= [s_i - s_c]_i \\
+		&= \left[\sum^{D_1,D_2}_{i,j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{c,j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right] \\
+		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
+	\end{align}
+	Using our result from \ref{classprobdiff}, we have:
+	\begin{gather}
+		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
+	\end{gather}
+	For a target class $c$, cross-entropy loss is defined as:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right)
+	\end{gather}
+	We can break down each contrastive map to foreground and background components:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} - \sum (1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)} \right)
+	\end{gather}
+	The current loss formulation motivates improving contrast using either the foreground or the background.
+We penalize contrast on the background, motivating learning invariance to noise in the form of the background.
+	\begin{gather}
+		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+	\end{gather}
+	This gives us our ablated cross-entropy formulation.
+\end{proof}
 
@@
 
-\section{Results}
+\section{Results [OUTDATED, IGNORE FOR NOW]}
 
@@
 
-\vspace{-2em}
 \begin{figure}[H]

### 2024-09-27 src/const.py

@@ -12,13 +12,13 @@ MODELS_DIR = BASE_DIR / 'models'
 DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
 LAMBDAS = [1E1, 5E1, 1, 5E1, 1]
-GRAD_ACCUMULATION_STEPS = 1
+GRAD_ACCUMULATION_STEPS = 4
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3
-EVAL_BATCH_SIZE = 16
+EVAL_BATCH_SIZE = 10
 WEIGHT_DECAY = 5E-4
 SELECT_BEST = True
-BATCH_SIZE = 80
+BATCH_SIZE = 20
 MOMENTUM = 0.9
 EPOCHS = 50
 
@@ -28,7 +28,7 @@ SPLITS = ['train', 'valid', 'test']
 IMAGE_SIZE = (224, 224)
 CAM_SIZE = (14, 14)
 N_CHANNELS = 3
-BINARY_CLS = False
+BINARY_CLS = True
 N_CLASSES = 2 if BINARY_CLS else 37
 IMAGE_SHAPE = (N_CHANNELS,) + IMAGE_SIZE
 

### 2024-09-27 src/evals/accuracy.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python3
 
-
 from ..data.pet_image import Dataset
 from ..model.arch import Model
 from src import const

### 2024-09-27 src/evals/average_prediction_noise.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python3
 
-
 from ..data.oxford_iiit_pet import Dataset
 from torch.utils.data import DataLoader
 from ..model.arch import Model

### 2024-09-27 src/evals/influence_comparison.py

@@ -1,7 +1,7 @@
 #!/usr/bin/env python3
 
-
 from ..data.oxford_iiit_pet import Dataset
+from matplotlib.colors import Normalize
 from torch.utils.data import DataLoader
 import matplotlib.pyplot as plt
 from ..model.arch import Model
@@ -36,10 +36,11 @@ def evaluate_influence(model, gen, misclassified_only=False):
             plt.imshow((X_i * heatmap_i).cpu().permute(1, 2, 0))
 
             heatmap_i = torchvision.transforms.functional.resize(heatmap_i, (14, 14), antialias=False).squeeze(0)
+            norm = Normalize(vmin=heatmap_i.min(), vmax=heatmap_i.max())
             fig.add_subplot(2, 2, 3)
-            plt.imshow((cc_i[y_i] * heatmap_i).cpu())
+            plt.imshow((cc_i[y_i] * heatmap_i).cpu(), norm=norm)
             fig.add_subplot(2, 2, 4)
-            plt.imshow((cc_i[y_i] * (1 - heatmap_i)).cpu())
+            plt.imshow((cc_i[y_i] * (1 - heatmap_i)).cpu(), norm=norm)
 
             fig.suptitle(f'Net Influence: {cc_i.sum()}\nForeground Influence: {cc_i[0][heatmap_i == 1].sum()}\nBackground Influence: {cc_i[0][heatmap_i != 1].sum()}')
             plt.tight_layout()
@@ -56,4 +57,4 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    evaluate_influence(model, DataLoader(Dataset('valid'), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 3)
+    evaluate_influence(model, DataLoader(Dataset('train'), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 3)

### 2024-09-27 src/evals/kernel_diff.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python3
 
-
 import matplotlib.pyplot as plt
 from ..model.arch import Model
 from src import const

### 2024-09-27 src/evals/sample_contrastive_cams.py

@@ -1,6 +1,5 @@
 #!/usr/bin/env python3
 
-
 from ..data.oxford_iiit_pet import Dataset
 from matplotlib.colors import Normalize
 import torch.nn.functional as F

### 2024-09-27 src/model/train.py

@@ -29,7 +29,7 @@ def fit(model, optimizer, scheduler, criterion, train, val,
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
             if not (epoch) % interval: print('-' * 10)
-            metrics = {metric: [] for metric in [f'{split}_{report}' for report in ['contrast_loss', 'background_loss', 'foreground_loss', 'acc', 'cse_loss'] for split in const.SPLITS[:2]]}
+            metrics = {metric: [] for metric in [f'{split}_{report}' for report in ['contrast_loss', 'acc', 'cse_loss'] for split in const.SPLITS[:2]]}
 
             for split, dataloader in zip(const.SPLITS[:2], (train, val)):
                 for batch_idx, (X, y) in enumerate(dataloader):

### 2024-09-28 src/const.py
created entropy-divergence loss

@@ -11,7 +11,7 @@ MODELS_DIR = BASE_DIR / 'models'
 # training
 DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
-LAMBDAS = [1E1, 5E1, 1, 5E1, 1]
+LAMBDAS = [1E1, 5E1, 1, 1E-3]
 GRAD_ACCUMULATION_STEPS = 4
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3

### 2024-09-28 src/evals/influence_comparison.py

@@ -57,4 +57,4 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    evaluate_influence(model, DataLoader(Dataset('train'), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 3)
+    evaluate_influence(model, DataLoader(Dataset('valid'), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 3)

### 2024-09-28 src/model/loss.py

@@ -10,33 +10,22 @@ class ContrastiveLoss(nn.Module):
     def __init__(self, get_contrastive_cam_fn, debug=False):
         super().__init__()
         self.get_contrastive_cam = get_contrastive_cam_fn
-        self.debug = debug
-
-    def kld(self, cc, fg_mask, reverse=False):
-        fg_mask[fg_mask > .5] = 1
-
-        cc_log_probs = F.sigmoid(cc)
-        fg_mask_probs = F.softmax((fg_mask * const.LAMBDAS[1]).flatten(start_dim=1), dim=1).reshape(cc.shape[0], *cc.shape[-2:]).repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
-        fg_mask[fg_mask <= .5] = .5
-        fg_mask_log_probs = fg_mask.log()
-        # fg_mask_log_probs[fg_mask > .5] = 0
-
-        # fg_mask_probs[fg_mask == 1] = .8 / (fg_mask == 1).sum()
-        # fg_mask_probs[fg_mask != 1] = .2 / (fg_mask != 1).sum()
-
-        return fg_mask_probs * (fg_mask_log_probs - cc_log_probs)
 
     def forward(self, y_pred, y):
         cc = self.get_contrastive_cam(y[1], y_pred[1])
         fg_mask = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+        ablation = (-cc * fg_mask + cc.abs() * (1 - fg_mask)).sum(dim=[2, 3])
 
-        kld = self.kld(cc.abs(), fg_mask.clone())
-        kld_prime = self.kld((-cc).abs(), -fg_mask.clone() + 1, reverse=True)
-        foreground = cc[fg_mask != 0].mean()
+        cc_log_probs = F.softmax((cc * const.LAMBDAS[0]).flatten(start_dim=2), dim=2).reshape(*cc.shape).log()
+        fg_mask_probs = F.softmax(((y[0]) * const.LAMBDAS[1]).flatten(start_dim=1), dim=1).reshape(cc.shape[0], *cc.shape[-2:]).repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
+        fg_mask_log_probs = fg_mask_probs.log()
+        fg_mask_log_probs[fg_mask != 0] = 0
 
[…]

### 2024-09-28 src/model/overfit.py

@@ -2,7 +2,7 @@
 
 from src.data.oxford_iiit_pet import Dataset
 import matplotlib.animation as animation
-from .loss import PieceWiseLoss
+from .loss import ContrastiveLoss
 import torch.nn.functional as F
 import matplotlib.pyplot as plt
 from .arch import Model
@@ -15,7 +15,7 @@ if __name__ == '__main__':
     data = Dataset('train')
     model = Model(const.IMAGE_SHAPE)
     optim = torch.optim.Adam(model.parameters(), lr=1E-3)
-    criterion = torch.nn.CrossEntropyLoss()
+    criterion = ContrastiveLoss(model.get_contrastive_cams)
 
     idx = random.randint(0, len(data) - 1)
     print(idx)
@@ -31,12 +31,7 @@ if __name__ == '__main__':
         y_pred = model(X)
         cc = model.get_contrastive_cams(y[1], y_pred[1])
 
-        if model.is_contrastive:
-            bg_masks = 1 - y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
-            out = (-cc * (1 - bg_masks) + cc.abs() * bg_masks).sum(dim=[2, 3])
-        else: out = y_pred[0]
-
-        loss = criterion(out, y[1])
+        loss = criterion(y_pred, y)
         loss.backward()
 
         frames.append([plt.imshow(F.interpolate(y[0][None], const.IMAGE_SIZE, mode='bilinear')[0][0].cpu(), cmap='jet', alpha=0.5, animated=True),

### 2024-09-28 src/model/train.py

@@ -1,7 +1,7 @@
 #!/usr/bin/env python3
 
 from ..data.oxford_iiit_pet import get_generators
-from .loss import PieceWiseLoss
+from .loss import ContrastiveLoss
 import torch.nn.functional as F
 from .arch import Model
 from src import const
@@ -29,26 +29,25 @@ def fit(model, optimizer, scheduler, criterion, train, val,
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch):
             if not (epoch) % interval: print('-' * 10)
-            metrics = {metric: [] for metric in [f'{split}_{report}' for report in ['contrast_loss', 'acc', 'cse_loss'] for split in const.SPLITS[:2]]}
+            metrics = {metric: [] for metric in [f'{split}_{report}' for report in ['contrast_loss', 'acc', 'divergence_loss', 'ablated_ce_loss', 'cse_loss'] for split in const.SPLITS[:2]]}
 
             for split, dataloader in zip(const.SPLITS[:2], (train, val)):
                 for batch_idx, (X, y) in enumerate(dataloader):
                     if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.zero_grad()
 
                     y_pred = model(X)
-                    if model.is_contrastive:
-                        cc = model.get_contrastive_cams(y[1], y_pred[1])
-                        fg_masks = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
-                        out = (-cc * fg_masks + cc.abs() * (1 - fg_masks)).sum(dim=[2, 3])
-                    else: out = y_pred[0]
-                    batch_loss = criterion(out, y[1])
+                    batch_loss = criterion(y_pred, y) if criterion._get_name() != 'CrossEntropyLoss' else criterion(y_pred[0], y[1])
 
                     metrics[f'{split}_acc'].extend(y[1].argmax(1).eq(y_pred[0].argmax(1)).unsqueeze(1).tolist())
                     metrics[f'{split}_contrast_loss'].append(batch_loss.item())
-                    metrics[f'{split}_cse_loss'].append(criterion(y_pred[0], y[1]).item())
+                    metrics[f'{split}_cse_loss'].append(F.cross_entropy(y_pred[0], y[1]).item())
 
                     del y_pred, X, y
                     torch.cuda.empty_cache()
+
+                    if criterion._get_name() != 'CrossEntropyLoss':
[…]

### 2024-10-17 paper/main.tex
updated paper, bugfixed evals (again)

@@
 
-\begin{theorem} We can integrate background invariance to the definition of cross-entropy using the following formulation:
+\begin{theorem}\label{ablatedce}
+	We can integrate background invariance to the definition of cross-entropy using the following formulation:
 	\begin{gather}
@@
 
-\section{Results [OUTDATED, IGNORE FOR NOW]}
+\section{Results}
 
-\subsection{Binary Classification}
+\subsection{Training Invariance to Spurious Correlation}
 
-For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $44$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
-The total time for training was $\approx$1hr.
+One significant concern uncovered both theoretically from \ref{ablatedce} and empirically from \ref{contr-cam} was that the standard classification pipeline motivates learning spurious correlations between the model, provided the spurious information is able to improve accuracy.
+This motivates learning distribution-specific correlations, which worsens it's ability to generalize.
+As an example
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=.3\textwidth]{img/default_cam.png}
+	\hspace{1em}
+	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
+	\hspace{1em}
+	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
+	\caption{Left: contrastive activation maps from the default training run.
+Middle: contrastive activations maps from the Abalted CE+KLD training run.
+Right: contrastive activation maps from the Abalted CE training run.}
+\end{figure}
+
+\subsection{Benchmark: Binary Classification}
+
+For binary classification, the model was trained for $100$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $80$ on a 32GB V100 on a randomly shuffled dataset using the original data splits.
+The total time for training was $\approx$4hr.
 
@@
 		\toprule
-		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
+		\textbf{Method}  & \textbf{Training CE Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
 		\midrule
-		Cross-Entropy & 0.646 & 66.1\% & 64.5\% \\
-		Interpretable (Ours) & \bf 0.326 & \bf 98.6\% & \bf 92.3\% \\
+		Cross-Entropy & 0.361 & 95.0\% & 82.8\% \\
+		Ablated CE (Ours) & 0.333 & 98.4\% & 88.2\% \\
+		Ablated CE + KLD (Ours) & \bf 0.322 & \bf 99.2\% & \bf 91.5\% \\
 		\bottomrule
@@
 	\caption{Plots comparing benchmark cross-entropy loss, training and validation accuracy.
-The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_f4f7dea52fa74739be788e5f47da1195\%22,\%22m_1301bf1ba9bf4db48e09e7497c8cd214\%22]}{here}.
-The blue line represents cross-entropy loss, while the orange line represents the contrastive approach.}
+The web version of this comparison can be found \href{https://dagshub.com/jinensetpal/contrastive-optimization/experiments\#/compare?experiments=[\%22m_cea5cf4fcbbb4c3f984b2396927cf80c\%22,\%22m_101287828f604122a092a2501ec3facb\%22]}{here}.
+The orange line represents cross-entropy loss, while the blue line represents the contrastive approach.}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{img/default_cam.png}
-	\hspace{3em}
-	\includegraphics[width=.4\textwidth]{img/contrastive_cam.png}
-	\caption{Left: contrastive activation maps from the default training run. Right: contrastive activation maps from the contrastive training run.}
+	\includegraphics[width=.3\textwidth]{img/default_cam.png}
+	\hspace{1em}
+	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
+	\hspace{1em}
+	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
+	\caption{Left: contrastive activation maps from the default training run.
+Middle: contrastive activations maps from the Abalted CE+KLD training run.
+Right: contrastive activation maps from the Abalted CE training run.}
 \end{figure}
 
-\subsection{Multiclass Classification}
+\subsection{Benchmark: Multiclass Classification}
 
@@
 		\midrule
-		Cross-Entropy & 3.627 & 2.8\% & 3.1\% \\
-		Interpretable (Ours) & \bf 2.743 & \bf 93.0\% & \bf 51.5\% \\
+		Cross-Entropy & 3.605 & 3.7\% & 5.2\% \\
+		Abltated CE (Ours) & 3.610 & 3.1\% & 4.9\% \\
+		Abltated CE + KLD (Ours) & \bf 2.743 & \bf 93.0\% & \bf 51.5\% \\
 		\bottomrule
@@
 
-The model was trained for 731 epochs with the Adam optimizer with a learning rate of $10^{-3}$ with a batch size of 12, with 5 gradient accumulation steps (effective batch size of 60) on a 32GB V100 on a randomly shuffled dataset using the original splits.
-The default model was only trained for 30 epochs, by which point it had already converged.
-For posterity, we will run the default setup for 731 epochs as well before publication.
+The contrastive model was trained for 227 epochs with the Adam optimizer with a learning rate of $10^{-3}$ with a batch size of 20, with 4 gradient accumulation steps (effective batch size of 80) on a 32GB V100 on a randomly shuffled dataset using the original splits.
+The default model was only trained for 33 epochs, while the ablated cross-entropy loss was only trained for 98 epochs by which point both had converged.
+For posterity, we will run each setup for 227 epochs as well before publication.
 
-\subsection{Out-of-Distribution Generalization}
+\subsection{Benchmark: Out-of-Distribution Generalization}
 
@@
 		\midrule
-		Cross-Entropy & 49.5\% \\
-		Interpretable (Ours) & \bf 78.5\% \\
+		Cross-Entropy & 77.0\% \\
+		Abltated CE (Ours) & 75.7\% \\
+		Abltated CE + KLD (Ours) & \bf 83.4\% \\
 		\bottomrule

### 2024-10-17 src/evals/influence_comparison.py

@@ -26,10 +26,10 @@ def evaluate_influence(model, gen, misclassified_only=False):
             X = X[mask]
             cc = cc[mask]
             heatmap = heatmap[mask]
-            y_pred = [y_pred_i[mask] for y_pred_i in y_pred]
 
-        for idx, (X_i, cc_i, heatmap_i, y_pred_i) in enumerate(zip(X, cc, heatmap, y_pred[0].argmax(1))):
+        for idx, (X_i, cc_i, heatmap_i) in enumerate(zip(X, cc, heatmap)):
             fig = plt.figure()
+            cc_i = cc_i.sum(dim=0)
 
             fig.add_subplot(2, 2, 1)
             plt.imshow(X_i.cpu().permute(1, 2, 0))
@@ -38,8 +38,8 @@ def evaluate_influence(model, gen, misclassified_only=False):
 
             heatmap_i = torchvision.transforms.functional.resize(heatmap_i, (14, 14), antialias=False).squeeze(0)
             norm = Normalize(vmin=cc_i.min(), vmax=cc_i.max())
-            foreground = cc_i[y_pred_i] * heatmap_i
-            background = cc_i[y_pred_i] * (1 - heatmap_i)
+            foreground = cc_i * heatmap_i
+            background = cc_i * (1 - heatmap_i)
 
             fig.add_subplot(2, 2, 3)
             plt.imshow(foreground.cpu(), norm=norm)

### 2024-10-17 src/evals/sample_contrastive_cams.py

@@ -16,9 +16,10 @@ def visualize(model, gen, norm=None):
                      facecolor='white')
 
     for idx, sample in enumerate(random.sample(range(len(gen)), 16)):
+        print(sample)
         X, (heatmap, y) = gen[sample]
         y_pred, cam = model(X.unsqueeze(0))
-        cam = model.get_contrastive_cams(y.unsqueeze(0), cam).detach()[0, y.argmin(0)]
+        cam = model.get_contrastive_cams(y.unsqueeze(0), cam).detach()[0, y.argmin(0)].abs()
         fig.add_subplot(4, 4, idx + 1)
         plt.xlabel(f'Pred: {str(y_pred[0].argmax().item())}, Actual: {str(y.argmax().item())}')
         plt.imshow(X.permute(1, 2, 0).cpu().detach(), alpha=0.5)
@@ -38,5 +39,5 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    if len(sys.argv) == 3 and sys.argv[2] == 'normed': visualize(model, Dataset('test'), Normalize(vmin=-.5, vmax=.5))
-    else: visualize(model, Dataset('test'))
+    if len(sys.argv) == 3 and sys.argv[2] == 'normed': visualize(model, Dataset('train'), Normalize(vmin=-.5, vmax=.5))
+    else: visualize(model, Dataset('train'))

### 2024-10-17 src/model/arch.py
pre-trained weights retained instead of overwritten, logging metrics before-training for baselining

@@ -7,25 +7,25 @@ import torch
 
 
 class Model(nn.Module):
-    def __init__(self, input_shape, is_contrastive=True, no_downsampling=False):
+    def __init__(self, input_shape, is_contrastive=True, no_downsampling=False, pretrained=False):
         super().__init__()
 
         self.is_contrastive = is_contrastive
-        self.backbone = torchvision.models.resnet50(weights=None)
+        self.backbone = torchvision.models.resnet50(weights=torchvision.models.ResNet50_Weights.IMAGENET1K_V2 if pretrained else None)
 
         if no_downsampling:
-            self.backbone.conv1 = nn.Conv2d(3, 64, kernel_size=(7, 7), stride=(1, 1), padding=(3, 3), bias=False)
-            self.backbone.maxpool = nn.MaxPool2d(kernel_size=3, stride=1, padding=1, dilation=1, ceil_mode=False)
+            self.backbone.conv1.stride = (1, 1)
+            self.backbone.maxpool.stride = 1
 
-            self.backbone.layer2[0].conv2 = nn.Conv2d(128, 128, kernel_size=(3, 3), stride=(1, 1), padding=(1,1), bias=False)
-            self.backbone.layer2[0].downsample[0] = nn.Conv2d(256, 512, kernel_size=(1, 1), stride=(1, 1), bias=False)
+            self.backbone.layer2[0].conv2.stride = (1, 1)
+            self.backbone.layer2[0].downsample[0].stride = (1, 1)
 
-            self.backbone.layer3[0].conv2 = nn.Conv2d(256, 256, kernel_size=(3, 3), stride=(1, 1), padding=(1,1), bias=False)
-            self.backbone.layer3[0].downsample[0] = nn.Conv2d(512, 1024, kernel_size=(1, 1), stride=(1, 1), bias=False)
+            self.backbone.layer3[0].conv2.stride = (1, 1)
+            self.backbone.layer3[0].downsample[0].stride = (1, 1)
 
         # one ablated downsampling is required for (224, 224) input
-        self.backbone.layer4[0].conv2 = nn.Conv2d(512, 512, kernel_size=(3, 3), stride=(1, 1), padding=(1,1), bias=False)
-        self.backbone.layer4[0].downsample[0] = nn.Conv2d(1024, 2048, kernel_size=(1, 1), stride=(1, 1), bias=False)
+        self.backbone.layer4[0].conv2.stride = (1, 1)
[…]

### 2024-10-17 src/model/train.py

@@ -14,7 +14,7 @@ import sys
 
 
 def fit(model, optimizer, scheduler, criterion, train, val,
-        selected=None, init_epoch=1, mlflow_run_id=None):
+        selected=None, init_epoch=0, mlflow_run_id=None):
     model.train()
     start_time = time.time()
     selected = selected or {'last': model.state_dict(),
@@ -49,9 +49,9 @@ def fit(model, optimizer, scheduler, criterion, train, val,
                         metrics[f'{split}_ablated_ce_loss'].append(criterion.prev[0])
                         metrics[f'{split}_divergence_loss'].append(criterion.prev[1])
 
-                    if split == 'train':
+                    if split == 'train' and epoch > 0:  # epoch 0 is for evaluating performance on initalization
                         batch_loss.backward()
-                        mlflow.log_metric(f'{split}_batchwise_loss', batch_loss.item(), step=(epoch-1) * len(dataloader) + batch_idx)
+                        mlflow.log_metric(f'{split}_batchwise_loss', batch_loss.item(), step=epoch * len(dataloader) + batch_idx)
 
                         if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.step()
             scheduler.step()
@@ -71,6 +71,11 @@ def fit(model, optimizer, scheduler, criterion, train, val,
             if const.TRAIN_CUTOFF is not None and time.time() - start_time >= const.TRAIN_CUTOFF: break
 
         selected['last'] = model.state_dict()
+        if const.SELECT_BEST and 'best' not in selected:
+            selected['best'] = model.state_dict()
+            selected['epoch'] = epoch
+            selected['acc'] = metrics['valid_acc']
+
         if const.SELECT_BEST:
             mlflow.log_metrics({'selected_epoch': selected['epoch'],
                                 'selected_valid_acc': selected['acc']}, step=epoch)
@@ -87,7 +92,7 @@ if __name__ == '__main__':
     const.MODEL_NAME = sys.argv[1]
     if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
 
-    model = Model(const.IMAGE_SHAPE, is_contrastive=const.MODEL_NAME != 'default')
+    model = Model(const.IMAGE_SHAPE, is_contrastive='default' not in const.MODEL_NAME, pretrained='pretrained' in const.MODEL_NAME)
 
[…]

### 2024-10-21 src/const.py
added finetuning with frozen parameters

@@ -13,11 +13,13 @@ DEVICE = 'cuda' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
 LAMBDAS = [1E1, 5E1, 1, 1E-3]
 GRAD_ACCUMULATION_STEPS = 1
+PRETRAINED_BACKBONE = False
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3
 EVAL_BATCH_SIZE = 10
 WEIGHT_DECAY = 5E-4
 SELECT_BEST = True
+FINETUNING = True
 BATCH_SIZE = 80
 MOMENTUM = 0.9
 EPOCHS = 50

### 2024-10-21 src/model/arch.py

@@ -7,11 +7,11 @@ import torch
 
 
 class Model(nn.Module):
-    def __init__(self, input_shape, is_contrastive=True, no_downsampling=False, pretrained=False):
+    def __init__(self, input_shape, is_contrastive=True, no_downsampling=False):
         super().__init__()
 
         self.is_contrastive = is_contrastive
-        self.backbone = torchvision.models.resnet50(weights=torchvision.models.ResNet50_Weights.IMAGENET1K_V2 if pretrained else None)
+        self.backbone = torchvision.models.resnet50(weights=torchvision.models.ResNet50_Weights.IMAGENET1K_V2 if const.PRETRAINED_BACKBONE else None)
 
         if no_downsampling:
             self.backbone.conv1.stride = (1, 1)

### 2024-10-21 src/model/train.py

@@ -27,7 +27,7 @@ def fit(model, optimizer, scheduler, criterion, train, val,
         mlflow.log_param('optimizer_fn', 'SGD')
 
         interval = max(1, (const.EPOCHS // 10))
-        for epoch in range(init_epoch, const.EPOCHS + init_epoch):
+        for epoch in range(init_epoch, const.EPOCHS + init_epoch + int(init_epoch == 0)):
             if not (epoch) % interval: print('-' * 10)
             metrics = {metric: [] for metric in [f'{split}_{report}' for report in ['contrast_loss', 'acc', 'divergence_loss', 'ablated_ce_loss', 'cse_loss'] for split in const.SPLITS[:2]]}
 
@@ -90,12 +90,15 @@ def fit(model, optimizer, scheduler, criterion, train, val,
 if __name__ == '__main__':
     # Usage: $ python -m path.to.script model_name --nocheckpoint
     const.MODEL_NAME = sys.argv[1]
+    const.PRETRAINED_BACKBONE = 'pretrained' in const.MODEL_NAME
     if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
 
-    model = Model(const.IMAGE_SHAPE, is_contrastive='default' not in const.MODEL_NAME, pretrained='pretrained' in const.MODEL_NAME)
+    model = Model(const.IMAGE_SHAPE, is_contrastive='default' not in const.MODEL_NAME)
 
     train, val, test = get_generators()
-    optimizer = torch.optim.SGD(model.parameters(),
+    optimizer = torch.optim.SGD([*model.linear.parameters(),
+                                 *model.backbone.layer4[0].conv2.parameters(),
+                                 *model.backbone.layer4[0].downsample[0].parameters()] if const.FINETUNING else model.parameters(),
                                 lr=const.LEARNING_RATE,
                                 momentum=const.MOMENTUM,
                                 weight_decay=const.WEIGHT_DECAY)

### 2024-10-21 src/data/cifar10.py
added cifar-10

@@ -1,0 +1,31 @@
+#!/usr/bin/env python3
+
+import torchvision.transforms as transforms
+import torchvision
+import torch
+
+
+def get_generators():
+    transform_train = transforms.Compose([
+        transforms.RandomCrop(32, padding=4),
+        transforms.RandomHorizontalFlip(),
+        transforms.ToTensor(),
+        transforms.Normalize((0.4914, 0.4822, 0.4465), (0.2023, 0.1994, 0.2010)),
+    ])
+
+    transform_test = transforms.Compose([
+        transforms.ToTensor(),
+        transforms.Normalize((0.4914, 0.4822, 0.4465), (0.2023, 0.1994, 0.2010)),
+    ])
+
+    trainset = torchvision.datasets.CIFAR10(
+        root='./data', train=True, download=True, transform=transform_train)
+    trainloader = torch.utils.data.DataLoader(
+        trainset, batch_size=128, shuffle=True, num_workers=2)
+
+    testset = torchvision.datasets.CIFAR10(
+        root='./data', train=False, download=True, transform=transform_test)
+    testloader = torch.utils.data.DataLoader(
+        testset, batch_size=100, shuffle=False, num_workers=2)
+
+    return trainloader, testloader, None

### 2024-10-21 paper/main.tex
notation & typo fixes

@@
 \begin{gather}
-	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
+	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\}^{|C|-1}_{c_{t'} \in c - c_t}
 \end{gather}
@@
 \begin{gather}
-	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|c| - 1)
+	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|C| - 1)
 \end{gather}
@@
 
-\subsubsection{Divergence Formulation}
-Segmentations maps describe the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
+\subsubsection{Divergence Minimization}
+Segmentation maps describe the region over which the classification target is present. We can leverage this information to define a constrastive cost function as follows:
 \begin{gather}
@@
 	\end{cases} \\
-	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
-		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
-		0 & \text{otherwise}
-	\end{cases} \\
-	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
-		0 & h_{i,j} = 0 \\
-		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
-	\end{cases} \\
-	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
+	D_{KL}(P || Q) := \sum_{i,j} P \cdot \log \frac{P'}{Q} \text{ where } 
 	P'_{i,j} := \begin{cases}
@@
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \mathcal{\bm{A}}^{\text{contrastive}}))}_{\text{minimize divergence from target shape}} \\
-		&+ \underbrace{\frac{\lambda_4}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 0)} ||\mathcal{\bm{A}}^{\text{background}}||^2_F}_{\text{minimize background influence}} \\
-		&- \underbrace{\frac{\lambda_5}{\sum^{D_1,D_2}_{i,j}\mathbbm{1}(h_{i,j} = 1)} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}}_{\text{maximize foreground influence}}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)}))}_{\text{minimize divergence from target shape}} \\
+		&+ \underbrace{\frac{\lambda_4}{(|C| - 1) \left(D_1 D_2 - \sum^{D_1,D_2}_{i,j}h_{i,j}\right)} || (1 - h) \odot \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)}||^2_F}_{\text{minimize background influence}} \\
+		&- \underbrace{\frac{\lambda_5}{(|C| - 1) \sum^{D_1,D_2}_{i,j}h_{i,j}} \sum^{|C|-1,D_1,D_2}_{i,j,k} \left(\tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c,i)} \odot h\right)_{j,k} }_{\text{maximize foreground influence}}
 	\end{split}
@@
 	\end{align}
+	As an aside: we note that when $i = c$, the contrastive map is the zero matrix $\in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
 Using our result from \ref{classprobdiff}, we have:
@@
 	\caption{Left: contrastive activation maps from the default training run.
-Middle: contrastive activations maps from the Abalted CE+KLD training run.
-Right: contrastive activation maps from the Abalted CE training run.}
+Middle: contrastive activations maps from the Ablated CE+KLD training run.
+Right: contrastive activation maps from the Ablated CE training run.}
 \end{figure}
@@
 		Cross-Entropy & 3.605 & 3.7\% & 5.2\% \\
-		Abltated CE (Ours) & 3.610 & 3.1\% & 4.9\% \\
-		Abltated CE + KLD (Ours) & \bf 2.743 & \bf 93.0\% & \bf 51.5\% \\
+		Ablated CE (Ours) & 3.610 & 3.1\% & 4.9\% \\
+		Ablated CE + KLD (Ours) & \bf 2.743 & \bf 93.0\% & \bf 51.5\% \\
 		\bottomrule
@@
 		Cross-Entropy & 77.0\% \\
-		Abltated CE (Ours) & 75.7\% \\
-		Abltated CE + KLD (Ours) & \bf 83.4\% \\
+		Ablated CE (Ours) & 75.7\% \\
+		Ablated CE + KLD (Ours) & \bf 83.4\% \\
 		\bottomrule

### 2024-10-23 paper/main.tex
ce/ace now have representative comparison

@@
 		d &= [s_i - s_c]_i \\
-		&= \left[\sum^{D_1,D_2}_{i,j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{c,j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right] \\
+		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
 		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
 	\end{align}
-	As an aside: we note that when $i = c$, the contrastive map is the zero matrix $\in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
+	As an aside: we note that when $i = c$, the contrastive map is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps. \newline \\
 	Using our result from \ref{classprobdiff}, we have:

### 2024-10-23 src/model/arch.py

@@ -49,7 +49,8 @@ class Model(nn.Module):
         x = self.backbone(x)
         logits = self.linear(x)
 
-        return self.softmax(logits), self._compute_hi_res_cam(logits)
+        if self.training: return logits, self._compute_hi_res_cam(logits)
+        else: return self.softmax(logits), self._compute_hi_res_cam(logits)
 
     @staticmethod
     def get_contrastive_cams(y, cams):

### 2024-10-26 src/const.py
bp-free highres-cams, optimized finetuning
fun fact: till now, the last layer was *always* just the same as when it was initialized randomly

@@ -19,9 +19,9 @@ LEARNING_RATE = 1E-3
 EVAL_BATCH_SIZE = 10
 WEIGHT_DECAY = 5E-4
 SELECT_BEST = True
-FINETUNING = True
+FINETUNING = False
 OPTIMIZER = 'SGD'
-BATCH_SIZE = 80
+BATCH_SIZE = 32
 MOMENTUM = 0.9
 EPOCHS = 50
 
@@ -31,7 +31,7 @@ SPLITS = ['train', 'valid', 'test']
 IMAGE_SIZE = (224, 224)
 CAM_SIZE = (14, 14)
 N_CHANNELS = 3
-BINARY_CLS = True
+BINARY_CLS = False
 N_CLASSES = 2 if BINARY_CLS else 37
 IMAGE_SHAPE = (N_CHANNELS,) + IMAGE_SIZE
 

### 2024-10-26 src/evals/average_prediction_noise.py

@@ -2,6 +2,7 @@
 
 from ..data.oxford_iiit_pet import Dataset
 from torch.utils.data import DataLoader
+import matplotlib.pyplot as plt
 from ..model.arch import Model
 from src import const
 import random
@@ -10,6 +11,7 @@ import sys
 
 
 def average_prediction_noise(model, gen, misclassified_only=False):
+    pixelwise_influence = torch.zeros(const.CAM_SIZE, device=const.DEVICE)
     avgs = torch.tensor([0., 0., 0.], device=const.DEVICE)
     n_miscls = torch.tensor([0.], device=const.DEVICE)
 
@@ -24,12 +26,17 @@ def average_prediction_noise(model, gen, misclassified_only=False):
             heatmap = heatmap[mask]
             n_miscls += mask.sum()
 
+        pixelwise_influence += cc.sum(dim=0).sum(dim=0)
         avgs[0] += cc.sum()
         avgs[1] += cc[heatmap == 1].sum()
         avgs[2] += cc[heatmap != 1].sum()
 
     avgs /= n_miscls if misclassified_only else len(gen.dataset)
     print('\n'.join([f'{feature} Average Influence: {metric}' for (feature, metric) in zip(('Net', 'Foreground', 'Background'), avgs.tolist())]))
+
+    plt.imshow(pixelwise_influence.detach().cpu())
+    plt.savefig(const.DATA_DIR / 'evals' / f'{model.name}_influence.png')
+    plt.show()
 
 
 if __name__ == '__main__':

### 2024-10-26 src/model/arch.py

@@ -49,8 +49,8 @@ class Model(nn.Module):
         x = self.backbone(x)
         logits = self.linear(x)
 
-        if self.training: return logits, self._compute_hi_res_cam(logits)
-        else: return self.softmax(logits), self._compute_hi_res_cam(logits)
+        if self.training: return logits, self._bp_free_hi_res_cams(logits)
+        else: return self.softmax(logits), self._hi_res_cams(logits)
 
     @staticmethod
     def get_contrastive_cams(y, cams):
@@ -59,7 +59,18 @@ class Model(nn.Module):
 
         return contrastive_cams
 
-    def _compute_hi_res_cam(self, logits):
+    def _bp_free_hi_res_cams(self, logits):  # required to obtain gradients on self.linear.weight
+        cams = torch.zeros(*logits.shape, *self.feature_rect.shape[2:], device=const.DEVICE)
+        for img_idx in range(logits.shape[0]):
+            for class_idx, weight in enumerate(self.linear.weight):
+                cams[img_idx, class_idx] = (weight[None, None].repeat(14, 14, 1).permute(2, 0, 1) * self.feature_rect[img_idx]).sum(dim=0)
+        cams /= 14**2
+
+        self.feature_rect = None
+
+        return cams
+
+    def _hi_res_cams(self, logits):
         cams = torch.zeros(*logits.shape, *self.feature_rect.shape[2:], device=const.DEVICE)
         for img_idx in range(logits.shape[0]):
             for class_idx in range(logits.shape[1]):

### 2024-10-26 src/model/train.py

@@ -50,7 +50,7 @@ def fit(model, optimizer, scheduler, criterion, train, val,
                         metrics[f'{split}_divergence_loss'].append(criterion.prev[1])
 
                     if split == 'train' and epoch > 0:  # epoch 0 is for evaluating performance on initalization
-                        batch_loss.backward()
+                        batch_loss.backward(inputs=optimizer.param_groups[0]['params'])
                         mlflow.log_metric(f'{split}_batchwise_loss', batch_loss.item(), step=epoch * len(dataloader) + batch_idx)
 
                         if not (batch_idx+1) % const.GRAD_ACCUMULATION_STEPS: optimizer.step()
@@ -93,6 +93,7 @@ if __name__ == '__main__':
     const.FINETUNING = 'finetuned' in const.MODEL_NAME
     const.OPTIMIZER = 'Adam' if 'adam' in const.MODEL_NAME else 'SGD'
     const.PRETRAINED_BACKBONE = 'pretrained' in const.MODEL_NAME
+    if 'ablated_only' in const.MODEL_NAME: const.LAMBDAS[-1] = 0
 
     if const.LOG_REMOTE: mlflow.set_tracking_uri(const.MLFLOW_TRACKING_URI)
 

### 2024-10-28 paper/main.tex
bp-free highres-cams, optimized finetuning
fun fact: till now, the last layer was *always* just the same as when it was initialized randomly
added experiments to establish motivation for contrastive approach, added models, made minor bugfixes

@@
 \usepackage{amsthm}
+\usepackage{subfig}
 \usepackage{float}
@@
 \begin{gather}
-	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} ~~\text{ where }~~ \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c')} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
 \end{gather}
-This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map.
+This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information in the form of the activation map instead of flattening the feature map.
 If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.
@@
 \begin{gather}
-	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\}^{|C|-1}_{c_{t'} \in c - c_t}
+	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c - c_t}
 \end{gather}
@@
 
+\subsection{Observation: Cross-Entropy Motivates Learning from Spurious Correlations}
+
+Using contrastive activation maps, we can evaluate the attention maps generated by a \underline{well-trained} model (ImageNet weights, fine-tuned to $99.9\%$ training accuracy, $99.6\%$ validation accuracy) on Oxford-IIIT Pets:
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{img/default_cams.png}
+\end{figure}
+
+In this random sample, we observe that the regions within each of the images that influence the final prediction are misaligned to the classification target.
+Particularly, the top-left corner of the final two images contributed significantly to the final prediction, despite not containing the classification target.
+
+Interestingly, this dependence holds even when we average influence plots generated from inference across the dataset:
+\begin{figure}[h!]
+	\centering
+	\subfloat[Pre-Trained on Train]{\includegraphics[width=.24\textwidth]{img/default_pretrained_finetuned_best_train_influence.png}}
+	\subfloat[Pre-Trained on Valid]{\includegraphics[width=.24\textwidth]{img/default_pretrained_finetuned_best_valid_influence.png}}
+	\subfloat[Random Init on Valid]{\includegraphics[width=.24\textwidth]{img/default_best_valid_influence.png}}
+	\subfloat[Random Init on Train]{\includegraphics[width=.24\textwidth]{img/default_best_train_influence.png}}
+\end{figure}
+
+Further, we can use the segmentation trimaps from the Oxford-IIIT Pets dataset as foreground and background delimiters to bifurcate the regions that influence final prediction logits.
+On average, we observe the following:
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{c|cc|cc}
+		\toprule
+		\bf Average & \multicolumn{2}{c}{\bf Validation} & \multicolumn{2}{c}{\bf Training} \\
+		\bf Pixel Influence & \textbf{Random} & \textbf{Pre-Trained} & \textbf{Random} & \textbf{Pre-Trained} \\
+		\midrule
+		Foreground & 6.834 & \bf 2.622 & 6.673 & \bf 2.579 \\
+		Background & \bf 8.934 & 1.347 & \bf 8.738 & 1.332 \\
+		\midrule
+		Net & 15.769 & 3.970 & 15.410 & 3.912 \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+Crucially, we observe that weights optimized from random initialization leverage more of the background than the actual target to inform the final prediction.
+In training classification models, an implicit assumption is that the best approximator of the class label is the target itself, which experimentally are observed to not hold.
+Despite fine-tuning on pre-trained weights mitigating the problem, it still learnt a feature that can be exploited adversarially.
+
+This is especially concerning for cases wherein the target is far away from the camera, in which case the emphasis is objectively placed on learning the best spurious surrogate to the actual target, rather than obtaining an approximation of the target's feature representation using the fewer available pixels.
+
+% TODO: adversarial ablation
+
 \subsection{Constraining Optimization to an Interpretable Basis}
 
-As discussed in the previous section, the distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information.
+The primary distinction between the contrastive optimization objective over standard risk is that it preserves spatial information.
 We know at a pointwise level how each feature contributes to the final prediction.
@@
 The model through the process of training should learn to distinguish and subsequently ignore the background in determining the final prediction.
+
+% TODO: formal analysis -- empirical performance test on scale augmentation + norm-change in feature representation.
+
+In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target also motivates learning scale-robust representations of the target.
 
@@
 \begin{theorem}\label{ablatedce}
-	We can integrate background invariance to the definition of cross-entropy using the following formulation:
+	We can integrate backgroud invariance to the definition of cross-entropy using the following formulation:
 	\begin{gather}
@@
 	\end{align}
-	As an aside: we note that when $i = c$, the contrastive map is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps. \newline \\
+	As an aside: we note that when $i = c$, the contrastive map obtained is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
+\newline \\
 	Using our result from \ref{classprobdiff}, we have:
@@
 	\begin{gather}
-		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
 	\end{gather}
@@
 
+\subsubsection{ACE-Divergence Formulation}
+
+An empirical observation from the behavior of cross-entropy loss is that it displays a tendency to generate contrast only to regions where feature differences are present within the dataset, which places an assumption for the same set of differenciating features being present across the underlying task.
+In addition, the peaky nature of the generated contrastive maps also encourages vulnerability to adversarial pertubation.
+
+We therefore combined the divergence based loss with ablated cross-entropy.
+The ablated cross-entropy term ensures accurate predictions using only the region where the classification target is present, and the divergence term mitigates the dataset-specific contrastive setting, motivating effectively learning an implicit segmentation map.
+
+\begin{gather}
+	\begin{split}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)})) \\
+		&- \lambda_4 \log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+	\end{split}
+\end{gather}
+
+For this setting, $\bm{\lambda} = \{1, 50, 1, 10^{-3}\}$ was set for $i = \{1, \ldots, 4\}$ respectively. As in the previous setting, these were identified empirically, without rigorous optimization.
+
 \subsection{Training Detail}
+
+% \begin{figure}
+% 	\centering
+% 	\includegraphics[width=\textwidth]{img/cams.pdf}
+% \end{figure}
 
@@
 We train the model on the binary classification case (image containing a dog or cat) as well as the fine-grained, breed specific 37 class split.
-The architecture used for training was ResNet-50, initialized with random parameters and three key modifications:
+The architecture used for training was ResNet-50, initialized either with random parameters or ImageNet, with three key modifications:
 
@@
 \section{Results}
-
-\subsection{Training Invariance to Spurious Correlation}
-
-One significant concern uncovered both theoretically from \ref{ablatedce} and empirically from \ref{contr-cam} was that the standard classification pipeline motivates learning spurious correlations between the model, provided the spurious information is able to improve accuracy.
-This motivates learning distribution-specific correlations, which worsens it's ability to generalize.
-As an example
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=.3\textwidth]{img/default_cam.png}
-	\hspace{1em}
-	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
-	\hspace{1em}
-	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
-	\caption{Left: contrastive activation maps from the default training run.
-Middle: contrastive activations maps from the Ablated CE+KLD training run.
-Right: contrastive activation maps from the Ablated CE training run.}
-\end{figure}
 
@@
 	\centering
-	\includegraphics[width=.3\textwidth]{img/default_cam.png}
+	\subfloat[Cross-Entropy (CE)]{\includegraphics[width=.3\textwidth]{img/default_cam.png}}
 	\hspace{1em}
-	\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}
+	\subfloat[Ablated CE]{\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}}
 	\hspace{1em}
-	\includegraphics[width=.3\textwidth]{img/contrastive_ace_cam.png}
-	\caption{Left: contrastive activation maps from the default training run.
-Middle: contrastive activations maps from the Ablated CE+KLD training run.
-Right: contrastive activation maps from the Ablated CE training run.}
+	\subfloat[Ablated CE+KLD]{\includegraphics[width=.3\textwidth]{img/contrastive_cam.png}}
+	\caption{Contrastive Class Activation Maps from each training setting.}
 \end{figure}

### 2024-10-28 src/const.py

@@ -31,7 +31,7 @@ SPLITS = ['train', 'valid', 'test']
 IMAGE_SIZE = (224, 224)
 CAM_SIZE = (14, 14)
 N_CHANNELS = 3
-BINARY_CLS = False
+BINARY_CLS = True
 N_CLASSES = 2 if BINARY_CLS else 37
 IMAGE_SHAPE = (N_CHANNELS,) + IMAGE_SIZE
 

### 2024-10-28 src/evals/average_prediction_noise.py

@@ -35,7 +35,7 @@ def average_prediction_noise(model, gen, misclassified_only=False):
     print('\n'.join([f'{feature} Average Influence: {metric}' for (feature, metric) in zip(('Net', 'Foreground', 'Background'), avgs.tolist())]))
 
     plt.imshow(pixelwise_influence.detach().cpu())
-    plt.savefig(const.DATA_DIR / 'evals' / f'{model.name}_influence.png')
+    plt.savefig(const.DATA_DIR / 'evals' / f'{model.name}_{gen.dataset.split}_influence.png')
     plt.show()
 
 
@@ -48,4 +48,4 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    average_prediction_noise(model, DataLoader(Dataset('valid'), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 3)
+    average_prediction_noise(model, DataLoader(Dataset(sys.argv[2]), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 4)

### 2024-10-28 src/evals/sample_contrastive_cams.py

@@ -21,7 +21,7 @@ def visualize(model, gen, norm=None):
         y_pred, cam = model(X.unsqueeze(0))
         cam = model.get_contrastive_cams(y.unsqueeze(0), cam).detach()[0, y.argmin(0)].abs()
         fig.add_subplot(4, 4, idx + 1)
-        plt.xlabel(f'Pred: {str(y_pred[0].argmax().item())}, Actual: {str(y.argmax().item())}')
+        plt.xlabel(f'Pred: {str(y_pred[0].argmax().item())}, Label: {str(y.argmax().item())}')
         plt.imshow(X.permute(1, 2, 0).cpu().detach(), alpha=0.5)
         plt.imshow(F.interpolate(cam[None, None, :], const.IMAGE_SIZE, mode='bilinear')[0, 0].cpu(), cmap='jet', alpha=0.5, norm=norm)
 

### 2024-10-28 src/model/arch.py

@@ -63,8 +63,8 @@ class Model(nn.Module):
         cams = torch.zeros(*logits.shape, *self.feature_rect.shape[2:], device=const.DEVICE)
         for img_idx in range(logits.shape[0]):
             for class_idx, weight in enumerate(self.linear.weight):
-                cams[img_idx, class_idx] = (weight[None, None].repeat(*self.CAM_SIZE, 1).permute(2, 0, 1) * self.feature_rect[img_idx]).sum(dim=0)
-        cams /= self.CAM_SIZE[0]**2
+                cams[img_idx, class_idx] = (weight[None, None].repeat(*const.CAM_SIZE, 1).permute(2, 0, 1) * self.feature_rect[img_idx]).sum(dim=0)
+        cams /= const.CAM_SIZE[0]**2
 
         self.feature_rect = None
 

### 2024-10-28 src/model/train.py

@@ -90,7 +90,7 @@ def fit(model, optimizer, scheduler, criterion, train, val,
 if __name__ == '__main__':
     # Usage: $ python -m path.to.script model_name --nocheckpoint
     const.MODEL_NAME = sys.argv[1]
-    const.BINARY_CLS = 'multiclass' in const.MODEL_NAME
+    const.BINARY_CLS = 'multiclass' not in const.MODEL_NAME
     const.N_CLASSES = 2 if const.BINARY_CLS else 37
     const.FINETUNING = 'finetuned' in const.MODEL_NAME
     const.OPTIMIZER = 'Adam' if 'adam' in const.MODEL_NAME else 'SGD'

### 2024-10-30 paper/main.tex
added models, updated evals, minor paper tweaks

@@
 
-\newtheorem{theorem}{Proposition}
-\newtheorem{corollary}{Corollary}[theorem]
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{assumption}[theorem]{Assumption}
+\theoremstyle{remark}
+\newtheorem{remark}[theorem]{Remark}
 
@@
 
-For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class.
-The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
+We can define ContrastiveCAMs as a generalization of the binary case.
+
+\begin{definition}
+	Given a set of classes $C$ with $c_t$ being the target class and $c_{t'}$ being a sampled non-target class, we define ContrastiveCAMs as follows:
 	\begin{gather}
-	\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c - c_t}
+		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c \setminus c_t}
 	\end{gather}
-This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.
+\end{definition}
+
+This enables us to evaluate ContrastiveCAMs in the multiclass setting.
+This is trivially a generalization of the previous case, as the same formulation applied to a binary classification task returns a set of contrastive maps containing a single element.
 

### 2024-10-30 src/const.py

@@ -16,7 +16,6 @@ GRAD_ACCUMULATION_STEPS = 1
 PRETRAINED_BACKBONE = False
 CHECKPOINTING = True
 LEARNING_RATE = 1E-3
-EVAL_BATCH_SIZE = 10
 WEIGHT_DECAY = 5E-4
 SELECT_BEST = True
 FINETUNING = False
@@ -24,6 +23,10 @@ OPTIMIZER = 'SGD'
 BATCH_SIZE = 32
 MOMENTUM = 0.9
 EPOCHS = 50
+
+# evals
+EVAL_BATCH_SIZE = 10
+CONFIDENCE_THRESHOLD = .85
 
 # dataset
 SEED = 1024

### 2024-10-30 src/evals/average_prediction_noise.py

@@ -10,7 +10,7 @@ import torch
 import sys
 
 
-def average_prediction_noise(model, gen, misclassified_only=False):
+def average_prediction_noise(model, gen, misclassified_low_confidence_only=False):
     pixelwise_influence = torch.zeros(const.CAM_SIZE, device=const.DEVICE)
     avgs = torch.tensor([0., 0., 0.], device=const.DEVICE)
     n_miscls = torch.tensor([0.], device=const.DEVICE)
@@ -20,8 +20,8 @@ def average_prediction_noise(model, gen, misclassified_only=False):
         cc = model.get_contrastive_cams(y[1], y_pred[1]).detach().abs()
         heatmap = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)
 
-        if misclassified_only:
-            mask = y[1].argmax(1) != y_pred[0].argmax(1)
+        if misclassified_low_confidence_only:
+            mask = (y[1].argmax(1) != y_pred[0].argmax(1)) | (y_pred[0].max(dim=1).values < const.CONFIDENCE_THRESHOLD)
             cc = cc[mask]
             heatmap = heatmap[mask]
             n_miscls += mask.sum()
@@ -31,7 +31,7 @@ def average_prediction_noise(model, gen, misclassified_only=False):
         avgs[1] += cc[heatmap == 1].sum()
         avgs[2] += cc[heatmap != 1].sum()
 
-    avgs /= n_miscls if misclassified_only else len(gen.dataset)
+    avgs /= n_miscls if misclassified_low_confidence_only else len(gen.dataset)
     print('\n'.join([f'{feature} Average Influence: {metric}' for (feature, metric) in zip(('Net', 'Foreground', 'Background'), avgs.tolist())]))
 
     plt.imshow(pixelwise_influence.detach().cpu())
@@ -48,4 +48,4 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    average_prediction_noise(model, DataLoader(Dataset(sys.argv[2]), batch_size=10, shuffle=False), misclassified_only=len(sys.argv) == 4)
+    average_prediction_noise(model, DataLoader(Dataset(sys.argv[2]), batch_size=10, shuffle=False), misclassified_low_confidence_only=len(sys.argv) == 4)

### 2024-10-30 src/evals/influence_comparison.py

@@ -12,8 +12,9 @@ import torch
 import sys
 
 
-def evaluate_influence(model, gen, misclassified_only=False):
-    (const.DATA_DIR / 'evals' / 'influence_plots' / gen.dataset.split / model.name).mkdir(exist_ok=True, parents=True)
+def evaluate_influence(model, gen, misclassified_low_confidence_only=False):
+    path = const.DATA_DIR / 'evals' / f'influence_plots{"_mscls" if misclassified_low_confidence_only else ""}' / gen.dataset.split / model.name
+    (path).mkdir(exist_ok=True, parents=True)
 
     const.CAM_SIZE = const.IMAGE_SIZE
     for batch_idx, (X, y) in enumerate(gen):
@@ -21,8 +22,8 @@ def evaluate_influence(model, gen, misclassified_only=False):
         cc = model.get_contrastive_cams(y[1], y_pred[1]).detach().abs()
         heatmap = y[0].repeat((cc.shape[1] - 1, 1, 1, 1)).permute(1, 0, 2, 3)
 
-        if misclassified_only:
-            mask = y[1].argmax(1) != y_pred[0].argmax(1)
+        if misclassified_low_confidence_only:
+            mask = (y[1].argmax(1) != y_pred[0].argmax(1)) | (y_pred[0].max(dim=1).values < const.CONFIDENCE_THRESHOLD)
             X = X[mask]
             cc = cc[mask]
             heatmap = heatmap[mask]
@@ -48,7 +49,8 @@ def evaluate_influence(model, gen, misclassified_only=False):
 
             fig.suptitle(f'Net Influence: {cc_i.sum()}\nForeground Influence: {foreground.sum()}\nBackground Influence: {background.sum()}')
             plt.tight_layout()
-            fig.savefig(const.DATA_DIR / 'evals' / 'influence_plots' / gen.dataset.split / model.name / f'{batch_idx * gen.batch_size + idx}.png')
+            fig.savefig(path / f'{batch_idx * gen.batch_size + idx}.png')
+            if background.sum() / foreground.sum() > 1: fig.savefig(path / f'{batch_idx * gen.batch_size + idx}_high_background.png')
             plt.close()
 
 
@@ -61,4 +63,4 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
[…]

### 2024-10-30 src/model/train.py

@@ -24,7 +24,6 @@ def fit(model, optimizer, scheduler, criterion, train, val,
     with mlflow.start_run(mlflow_run_id):
         # log hyperparameters
         mlflow.log_params({k: v for k, v in const.__dict__.items() if k == k.upper() and all(s not in k for s in ['DIR', 'PATH', 'SELECT_BEST'])})
-        mlflow.log_param('optimizer_fn', const.OPTIMIZER)
 
         interval = max(1, (const.EPOCHS // 10))
         for epoch in range(init_epoch, const.EPOCHS + init_epoch + int(init_epoch == 0)):

### 2024-12-17 paper/main.tex
added contrastive cutmix writeup to appendix

@@
 \usepackage{graphicx}
+\usepackage{amsfonts}
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{amsthm}
@@
 
-Crucially, we observe that weights optimized from random initialization leverage more of the background than the actual target to inform the final prediction.
+We observe that weights optimized from random initialization leverage more of the background than the actual target to inform the final prediction.
 In training classification models, an implicit assumption is that the best approximator of the class label is the target itself, which experimentally are observed to not hold.
@@
 \subsection{Constraining Optimization to an Interpretable Basis}
+\label{sec:constraining-optimization}
 
@@
 
-\begin{theorem}\label{ablatedce}
+\begin{theorem}\label{ace}
 	We can integrate backgroud invariance to the definition of cross-entropy using the following formulation:
 	\begin{gather}
-		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}|\right)} \right)
 	\end{gather}
@@
 
-Crucially, on comparing contrastive activation maps, we can see that each of the cases of the random sample, the model uses the target region for prediction and not spurious features.
+On comparing contrastive activation maps, we observe in each of the random samples, that the model uses the target region for prediction and not spurious features.
 In the case of the standard training procedure, this is not the case, and the model does not focus on the target in making it's prediction.
@@
 
+\paragraph{Scaling the Approach} Training ResNet-50 on ImageNet and PASCAL VOC can allow us to evaluate if the approach scales effectively, with the performance gain not being constrained to the Oxford-IIIT Pets Dataset.
+
 \paragraph{Understanding the Generalization Gap} While the model for the multiclass paradigm is able to fit the training dataset excellently -- achieving $\approx 93\%$ training accuracy -- the validation accuracy is significantly behind at $\approx 51\%$.
@@
 
+\appendix
+
+\section{Ablated Cross-Entropy Adaptations}
+
+In addition to cross-entropy, we also adapted training methodologies from \citet{imagenetrecipe}.
+These required additional modifications to the approach.
+These modifications are discussed in the following sections, remaining consistent with principles discussed in Section \ref{sec:constraining-optimization}.
+
+\subsection{Contrastive CutMix}
+
+CutMix\cite{yun2019cutmix} is a batch-wise augmentation technique that encourages better regularization by a) ``cutting'' out a randomized rectangle (randomized portion remaining consistent across the batch) of a given image and b) ``mixing'' the cut-out with it's neighbor.
+The corresponding labels are mixed by a randomly sampled parameter $\lambda$.
+For our experiments, $\lambda \sim \text{Uniform}(0,1)$.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\textwidth]{img/cutmix}
+\end{figure}
+
+CutMix can be effectively adapted to work with our setup: given a cut-out, we motivate differential contrast; if non-target pixels are contained in the new image, irresepective of the label, contrast should be generated in favor of the non-target class for the non-target, foreground subset of pixels.
+This affects the downstream prediction confidence, representative of the label being shifted from one-hot encoding to a randomized affine combination.
+
+Let segmentation mask $h$ take the following form:
+\begin{gather*}
+	h := \begin{cases}
+		-1 & \text{pixel does not contain any class} \\
+		c & \text{pixel contains class } c
+	\end{cases}
+\end{gather*}
+
+Then, Ablated Cross Entropy (\ref{ace}) is formulated as follows:
+\begin{gather}
+	\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum \mathbbm{1}_{c}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |\mathbbm{1}_{-1}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}| + \sum \mathbbm{1}_{i}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right)} \right)
+\end{gather}
+
+Where the third newly introduced term within the exponent expresses differential contrast.
+
 \end{document}

### 2024-12-31 src/model/overfit.py
added divergence-heavy model, generalized overfitting test for multilabel classification

@@ -1,7 +1,9 @@
 #!/usr/bin/env python3
+from matplotlib.colors import Normalize
 
-from src.data.oxford_iiit_pet import Dataset
+from ..data.oxford_iiit_pet import Dataset as oxford_iiit_pet
 import matplotlib.animation as animation
+from src.data.sbd import Dataset as sbd
 from .loss import ContrastiveLoss
 import torch.nn.functional as F
 import matplotlib.pyplot as plt
@@ -9,21 +11,24 @@ from .arch import Model
 from src import const
 import random
 import torch
+import sys
 
 
 if __name__ == '__main__':
-    data = Dataset('train')
+    multilabel = len(sys.argv) == 2
+    data = sbd('data/sbd', mode='segmentation') if multilabel else oxford_iiit_pet('train')
     model = Model(const.IMAGE_SHAPE)
     optim = torch.optim.Adam(model.parameters(), lr=1E-3)
-    criterion = ContrastiveLoss(model.get_contrastive_cams)
+    criterion = ContrastiveLoss(model.get_contrastive_cams, multilabel=multilabel, divergence=bool(const.LAMBDAS[-1]))
 
     idx = random.randint(0, len(data) - 1)
     print(idx)
     X = data[idx][0].unsqueeze(0)
     y = [x.unsqueeze(0) for x in data[idx][1]]
 
+    plt.imshow(y[0][0].cpu())
+
     frames = []
-    fig = plt.figure()
     for i in range(50):
         optim.zero_grad()
 
@@ -34,23 +39,28 @@ if __name__ == '__main__':
         loss = criterion(y_pred, y)
         loss.backward()
 
-        frames.append([plt.imshow(F.interpolate(y[0][None], const.IMAGE_SIZE, mode='bilinear')[0][0].cpu(), cmap='jet', alpha=0.5, animated=True),
-                       plt.imshow(F.interpolate(cc[:, y[1].argmin(1)], const.IMAGE_SIZE, mode='bilinear')[0, 0].detach().cpu(), cmap='jet', alpha=0.5, animated=True)])
+        if not multilabel:
+            frames.append([plt.imshow(F.interpolate(y[0][None], const.IMAGE_SIZE, mode='bilinear')[0][0].cpu(), cmap='jet', alpha=0.5, animated=True),
[…]

### 2025-01-01 paper/main.tex
added formulation for ablated BCE to paper

@@
 
+\subsubsection{Ablated Binary Cross-Entropy}
+
+For sigmoid-activated binary / multilabel classification tasks, we leverage similar principles to defined an ablation to binary cross-entropy.
+Since we do not have a contrastive process in softmax-activation, this definitions relies only on HiResCAMs.
+
+\begin{theorem}\label{abce}
+	We can integrate backgroud invariance to the definition of binary cross-entropy using the following formulation:
+	\begin{gather}
+		\begin{split}
+			H(\theta, \mathcal{D}) = &- y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) \\
+			&- (1 - y_i) \log \left(1 - \sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} + (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right)
+		\end{split}
+	\end{gather}
+\end{theorem}
+\begin{proof} For any logit $i$, binary cross-entropy loss is defined as:
+	\begin{align}
+		H(\theta, \mathcal{D}) &= -\left[ y_i \log (\sigma(\bm{s}_i)) + (1 - y_i) \log (1 - \sigma(\bm{s}_i)) \right] \\
+		&= -\left[ y_i \log \left(\sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) + (1 - y_i) \log \left(1 - \sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) \right]
+	\end{align}
+	Similar to Theorem \ref{ace}, we can break down each HiResCAM to foreground and background components:
+	\begin{gather}
+		\begin{split}
+			H(\theta, \mathcal{D}) = &- y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} + (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) \\
+			&- (1 - y_i) \log \left(1 - \sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} + (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right)
+		\end{split}
+	\end{gather}
+	The current formulation motivates activating either the foreground or the background for positive classification, and similarly motivates de-activating every pixel of the non-positive class.
+We therefore penalize activation on the background for the both cases, motivating learning invariance to noise in the form of the background despite it the improving the final logit score.
+	\begin{gather}
+		\begin{split}
+			H(\theta, \mathcal{D}) = &- y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) \\
+			&- (1 - y_i) \log \left(1 - \sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} + (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right)
+		\end{split}
+	\end{gather}
+	This gives us the ablated binary cross-entropy formulation.
+\end{proof}
+
 \subsection{Training Detail}

### 2025-01-01 src/model/loss.py
corrected implementation to reflect ablated bce formulation
also added more weights

@@ -20,7 +20,7 @@ class ContrastiveLoss(nn.Module):
         if self.multilabel:
             labels = ((torch.arange(const.N_CLASSES) + 1) * torch.ones(*const.CAM_SIZE, const.N_CLASSES)).T[None,].repeat(y[0].size(0), 1, 1, 1).to(const.DEVICE)
             fg_mask = (labels == y[0].repeat(1, const.N_CLASSES, 1).view(y[0].size(0), -1, *y[0].shape[1:])).to(torch.int)
-            ablation = (fg_mask * y_pred[1] - (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * y[1] + ((1 - fg_mask) * y_pred[1]).sum(dim=[2, 3]) * (1 - y[1])
+            ablation = (fg_mask * y_pred[1] - (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * y[1] + (fg_mask * y_pred[1] + (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * (1 - y[1])
         elif self.is_label_mask:
             cc = self.get_contrastive_cams(y[1], y_pred[1]).to(const.DEVICE)
 

### 2025-01-01 src/evals/ood_f1.py
added scheduler checkpointing, fixed ema, OOD actually testing on test now
updated model params

@@ -18,7 +18,7 @@ if __name__ == '__main__':
     model.name = name
     model.eval()
 
-    for test, test_loader in zip(('easy', 'hard'), get_generators('train')):
+    for test, test_loader in zip(('easy', 'hard'), get_generators('test')):
         preds = torch.empty(0, device=const.DEVICE)
         targets = torch.empty(0, device=const.DEVICE)
         for X, (heatmap, y) in test_loader:

### 2025-01-01 src/model/train.py

@@ -227,6 +227,7 @@ if __name__ == '__main__':
     if const.CHECKPOINTING and len(sys.argv) == 2:  # add extra sys.argv to signify first checkpointing run
         model.load_state_dict(torch.load(path / 'last.pt', map_location=torch.device(const.DEVICE), weights_only=True))
         optimizer.load_state_dict(torch.load(path / 'optim.pt', map_location=torch.device(const.DEVICE), weights_only=True))
+        scheduler.load_state_dict(torch.load(path / 'scheduler.pt', map_location=torch.device(const.DEVICE), weights_only=True))
         checkpoint_args = json.load(open(path / 'checkpoint_metadata.json'))
         prev_metrics = mlflow.get_run(checkpoint_args['mlflow_run_id']).data.metrics
 
@@ -235,8 +236,8 @@ if __name__ == '__main__':
                     'epoch': prev_metrics['selected_epoch'],
                     'acc': prev_metrics['selected_valid_acc']}
         if ema and (path / 'ema.pt').exists():
-            selected['ema'] = torch.load(path / 'ema.pt', map_location=torch.device(const.DEVICE), weights_only=False)
-            ema = selected['ema']
+            selected['ema'] = torch.load(path / 'ema.pt', map_location=torch.device(const.DEVICE), weights_only=True)
+            ema.load_state_dict(selected['ema'])
 
     completed_epochs, selected = fit(model, optimizer, scheduler, criterion, train, val, is_multilabel=is_multilabel, ema=ema, selected=selected, **checkpoint_args)
 
@@ -251,4 +252,5 @@ if __name__ == '__main__':
 
         if const.CHECKPOINTING:
             torch.save(optimizer.state_dict(), path / 'optim.pt')
+            torch.save(scheduler.state_dict(), path / 'scheduler.pt')
             json.dump({'init_epoch': completed_epochs+1, 'mlflow_run_id': mlflow.last_active_run().info.run_id}, open(path / 'checkpoint_metadata.json', 'w'))

### 2025-01-03 paper/main.tex
reverted ablated bce formulation to only constraint target-class cams

--- 2024-12-17 paper/main.tex
@@
 
+\subsubsection{Ablated Binary Cross-Entropy}
+
+For sigmoid-activated binary / multilabel classification tasks, we leverage similar principles to defined an ablation to binary cross-entropy.
+Since we do not have a contrastive process in softmax-activation, this definitions relies only on HiResCAMs.
+
+\begin{theorem}\label{abce}
+	We can integrate backgroud invariance to the definition of binary cross-entropy using the following formulation:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
+	\end{gather}
+\end{theorem}
+\begin{proof} For any logit $i$, binary cross-entropy loss is defined as:
+	\begin{align}
+		H(\theta, \mathcal{D}) &= -\left[ y_i \log (\sigma(\bm{s}_i)) + (1 - y_i) \log (1 - \sigma(\bm{s}_i)) \right] \\
+		&= -\left[ y_i \log \left(\sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) + (1 - y_i) \log \left(1 - \sigma\left(\sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) \right]
+	\end{align}
+	Similar to Theorem \ref{ace}, we can break down each HiResCAM to foreground and background components.
+For a non-target HiResCAM, anything that is not the target is inherently the background.
+Therefore, it's definition is retrained.
+	\begin{gather}
+		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
+	\end{gather}
+	The current formulation motivates activating either the foreground or the background for positive classification, and motivates de-activating every pixel of the non-positive class.
+We penalize activation on the background for the positive class only:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
+	\end{gather}
+	This gives us the ablated binary cross-entropy formulation.
+\end{proof}
+
 \subsection{Training Detail}

### 2025-01-03 scripts/train.sh

@@ -1,7 +1,7 @@
 #!/usr/bin/bash
 
 run=`ls .log/  | wc -l`
-echo "Writing to .logs/training-$run.log"
+echo "Writing to .log/training-$run.log"
 
 MLFLOW_TRACKING_USERNAME=jinensetpal \
 MLFLOW_TRACKING_PASSWORD=$MLFLOW_TOKEN \

### 2025-01-03 src/model/loss.py

@@ -20,7 +20,7 @@ class ContrastiveLoss(nn.Module):
         if self.multilabel:
             labels = ((torch.arange(const.N_CLASSES) + 1) * torch.ones(*const.CAM_SIZE, const.N_CLASSES)).T[None,].repeat(y[0].size(0), 1, 1, 1).to(const.DEVICE)
             fg_mask = (labels == y[0].repeat(1, const.N_CLASSES, 1).view(y[0].size(0), -1, *y[0].shape[1:])).to(torch.int)
-            ablation = (fg_mask * y_pred[1] - (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * y[1] + (fg_mask * y_pred[1] + (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * (1 - y[1])
+            ablation = (fg_mask * y_pred[1] - (1 - fg_mask) * y_pred[1].abs()).sum(dim=[2, 3]) * y[1] + y_pred[1].sum(dim=[2, 3]) * (1 - y[1])
         elif self.is_label_mask:
             cc = self.get_contrastive_cams(y[1], y_pred[1]).to(const.DEVICE)
 

### 2025-01-03 src/data/sbd.py
added average prediction eval for multilabel classification, logging module-specific ema weights separately
minor streamlining for SBD call

@@ -8,7 +8,7 @@ import torch
 
 class Dataset(torchvision.datasets.SBDataset):
     def __init__(self, *args, device=const.DEVICE, **kwargs):
-        super().__init__(*args, **kwargs)
+        super().__init__(const.DATA_DIR / 'sbd', *args, **kwargs)
         self.device = device
         self.reweight = torch.tensor([9.2773, 10.5697, 7.1412, 11.5777, 7.8395, 12.8758, 4.8783, 5.3312, 5.1691, 19.1396, 10.7298, 4.6035, 12.0710, 11.0651, 1.4098, 10.8117, 16.9283, 11.3914, 10.7028, 9.8814], device=const.DEVICE)  # computed over train to establish a .5 prior of detection for each class
 
@@ -27,7 +27,7 @@ def get_generators():
     torch.multiprocessing.set_start_method('spawn', force=True)
     const.SPLITS[1] = 'val'
 
-    dataloaders = *[torch.utils.data.DataLoader(Dataset(const.DATA_DIR / 'sbd', mode='segmentation', image_set=split, device='cpu', download=not (const.DATA_DIR / 'sbd').exists()),
+    dataloaders = *[torch.utils.data.DataLoader(Dataset(mode='segmentation', image_set=split, device='cpu', download=not (const.DATA_DIR / 'sbd').exists()),
                                                 num_workers=2, pin_memory=True, batch_size=const.BATCH_SIZE, shuffle=True) for split in const.SPLITS[:2]], None
     const.SPLITS[1] = 'valid'
     return dataloaders

### 2025-01-03 src/evals/sbd/average_prediction_noise.py

@@ -1,0 +1,52 @@
+#!/usr/bin/env python3
+
+from torch.utils.data import DataLoader
+from src.model.arch import Model
+from src.data.sbd import Dataset
+from src import const
+import random
+import torch
+import sys
+
+
+def average_prediction_noise(model, gen, failures_only=False):
+    avgs = torch.zeros(4, device=const.DEVICE)
+    n_miscls = torch.zeros(1, device=const.DEVICE)
+
+    for X, y in gen:
+        y_pred, cams = model(X)
+        conf = (y_pred - .5).abs() + .5
+
+        labels = ((torch.arange(const.N_CLASSES) + 1) * torch.ones(*const.CAM_SIZE, const.N_CLASSES)).T[None,].repeat(y[0].size(0), 1, 1, 1).to(const.DEVICE)
+        fg_mask = (labels == y[0].repeat(1, const.N_CLASSES, 1).view(y[0].size(0), -1, *y[0].shape[1:])).to(torch.int)
+
+        if failures_only:
+            mask = (y[1] != (y_pred > .5).to(torch.int)) | (conf < const.CONFIDENCE_THRESHOLD)
+            cams = cams[mask]
+            conf = conf[mask]
+            fg_mask = fg_mask[mask]
+            n_miscls += mask.sum()
+
+        avgs[0] += cams.sum().detach()
+        avgs[1] += (fg_mask * cams.abs()).detach().sum()
+        avgs[2] += ((1-fg_mask) * cams.abs()).detach().sum()
+        avgs[3] += conf.detach().sum()
+
+    tot = n_miscls.item() if failures_only else len(gen.dataset) * const.N_CLASSES
+    avgs /= tot
+    print(model.name)
+    print('\n'.join([f'{feature} Average Influence: {metric}' for (feature, metric) in zip(('Net', 'Foreground', 'Background'), avgs.tolist()[:-1])]))
+    print(f'#images: {tot}\nAverage Confidence: {avgs[-1].item()}')
+
+
+if __name__ == '__main__':
+    name = sys.argv[1]
+    random.seed(const.SEED)
+
+    model = Model(input_shape=const.IMAGE_SHAPE, is_contrastive='default' not in name)
+    model.load_state_dict(torch.load(const.MODELS_DIR / f'{name}.pt', map_location=const.DEVICE, weights_only=True))
+    model.name = name
+    model.eval()
+
+    torch.multiprocessing.set_start_method('spawn', force=True)
[…]

### 2025-01-03 src/model/overfit.py

@@ -16,7 +16,7 @@ import sys
 
 if __name__ == '__main__':
     multilabel = len(sys.argv) == 2
-    data = sbd('data/sbd', mode='segmentation') if multilabel else oxford_iiit_pet('train')
+    data = sbd(mode='segmentation') if multilabel else oxford_iiit_pet('train')
     model = Model(const.IMAGE_SHAPE)
     optim = torch.optim.Adam(model.parameters(), lr=1E-3)
     criterion = ContrastiveLoss(model.get_contrastive_cams, multilabel=multilabel, divergence=bool(const.LAMBDAS[-1]))

### 2025-01-03 src/model/train.py

@@ -145,7 +145,9 @@ def fit(model, optimizer, scheduler, criterion, train, val, is_multilabel=False,
             if const.TRAIN_CUTOFF is not None and time.time() - start_time >= const.TRAIN_CUTOFF: break
 
         if is_primary_rank:
-            if ema: selected['ema'] = ema.state_dict()
+            if ema:
+                selected['ema'] = ema.state_dict()
+                selected['ema_weights'] = ema.module.state_dict()
             selected['last'] = deepcopy(model.state_dict())
             if const.SELECT_BEST and 'best' not in selected:
                 selected['best'] = deepcopy(model.state_dict())
@@ -248,7 +250,9 @@ if __name__ == '__main__':
     if not const.DDP or (const.DDP and const.DEVICE == 0):
         torch.save(selected['last'], path / 'last.pt')
         if const.SELECT_BEST: torch.save(selected['best'], path / 'best.pt')
-        if const.EMA: torch.save(selected['ema'], path / 'ema.pt')
+        if const.EMA:
+            torch.save(selected['ema'], path / 'ema.pt')
+            torch.save(selected['ema_weights'], path / 'ema_weights.pt')
 
         if const.CHECKPOINTING:
             torch.save(optimizer.state_dict(), path / 'optim.pt')

### 2025-01-04 paper/main.tex
added ABCE_KLD formulation, minor bugfix

@@
 	\begin{gather}
-		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
+		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
 	\end{gather}
@@
 \end{proof}
+
+\subsubsection{Ablated Binary Cross-Entropy Divergence Formulation}
+
+Similar to ACE, we define a divergence term for the target class to motivate learning the foreground as part of the training process.
+\begin{gather}
+	\begin{split}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}_{i}^{\text{HiResCAM}})) \\
+		&- \lambda_2 \bigg[ y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) + (1 - y_i) \log \left( \bm{s}_i \right) \bigg]
+	\end{split}
+\end{gather}
 

### 2025-01-04 src/model/loss.py

@@ -49,7 +49,7 @@ class ContrastiveLoss(nn.Module):
             fg_mask_log_probs = fg_mask_probs.log()
             fg_mask_log_probs[fg_mask != 0] = 0
 
-            kld = fg_mask_probs * (fg_mask_log_probs - cam_log_probs) * fg_mask
+            kld = fg_mask_probs * (fg_mask_log_probs - cam_log_probs)
         else: kld = torch.tensor(0)
 
         self.prev = (ace.item(), (kld.sum() / y[0].size(0)).item())
@@ -112,7 +112,6 @@ class KLDPenaltyLoss(nn.Module):
         super().__init__()
         self.get_contrastive_cam = get_contrastive_cam_fn
         self.debug = debug
-
     def forward(self, y_pred, y):
         cc = self.get_contrastive_cam(y[1], y_pred[1])
         fg_mask = y[0].repeat((cc.shape[1], 1, 1, 1)).permute(1, 0, 2, 3)

### 2025-01-04 src/evals/sbd/ap.py
added trained models, AP eval

@@ -1,0 +1,35 @@
+#!/usr/bin/env python3
+
+from torcheval.metrics.functional import binary_auprc
+from torch.utils.data import DataLoader
+from src.data.sbd import Dataset
+from src.model.arch import Model
+from src import const
+import random
+import torch
+import sys
+
+
+def average_precision(model, gen):
+    score = 0
+
+    preds = torch.empty(0, device=const.DEVICE)
+    targets = torch.empty(0, device=const.DEVICE)
+    for X, (heatmap, y) in gen:
+        preds = torch.cat([preds, model(X.to(const.DEVICE))[0].detach().flatten()])
+        targets = torch.cat([targets, y.to(const.DEVICE).flatten()])
+
+    return binary_auprc(preds, targets)
+
+
+if __name__ == '__main__':
+    name = sys.argv[1]
+    random.seed(const.SEED)
+
+    model = Model(input_shape=const.IMAGE_SHAPE, is_contrastive='default' not in name)
+    model.load_state_dict(torch.load(const.MODELS_DIR / f'{name}.pt', map_location=const.DEVICE, weights_only=True))
+    model.name = name
+    model.eval()
+
+    torch.multiprocessing.set_start_method('spawn', force=True)
+    print(average_precision(model, DataLoader(Dataset(mode='segmentation', image_set=sys.argv[2]), batch_size=const.BATCH_SIZE, num_workers=4, shuffle=False)))

### 2025-02-03 paper/main.tex
switched hardinet to module-based invocation, updated paper title, drafted req. tables (w/o entries)
updated model weights, hardinet dataloader faulty `self.` call removed

@@
 
-\title{Contrastive Optimization -- Preliminaries}
+\title{Interpretable Casual Interventions \\
+Encourage Core Risk Minimization}
 
@@
 For a non-target HiResCAM, anything that is not the target is inherently the background.
-Therefore, it's definition is retrained.
+Therefore, it's definition is retained.
 	\begin{gather}
@@
 \end{gather}
+
+\subsubsection{Spatial Sliced Wasserstein Loss}
 
@@
 
-\paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization}.
+\paragraph{Reproducibility.} The source code, dataset, experiments, evals, and model weights can be found at \url{https://dagshub.com/jinensetpal/contrastive-optimization.git}.
 
 \section{Results}
+
+\subsection{Overview}
+
+We present training results on the following settings:
+\begin{enumerate}
+	\item \textbf{Oxford IIIT-Pets Dataset:} Binary Classification (Cat / Dog Classification) -- OOD Evaluation without semantic-shift.
+Multiclass Classification on 37-class fine-grained classification task.
+	\item \textbf{Semantic Boundaries Dataset:} 20-class multilabel classification, evaluation on VOC2012 Test Set reporting AP, mIOU.
+	\item \textbf{S-OOD ImageNet:} reporting in-distribution accuracy, 
+\end{enumerate}
 
@@
 
+\subsection{Benchmark: Hard-Imagenet}
+
+\begin{table}[H]
+	\centering
+	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
+	\begin{tabular}{c|c|ccc|c}
+		\toprule
+		\textbf{Method} & \textbf{None} $(\uparrow)$ & \textbf{Gray Mask FG} $(\downarrow)$ & \textbf{Gray BBOX FG} $(\downarrow)$ & \textbf{Tile FG} $(\downarrow)$ & \textbf{RFS} $(\uparrow)$ \\
+		\midrule
+		Cross-Entropy & 94.25 & 75.94 & 69.39 & 67.38 & -0.18 \\
+		\midrule
+		CORM & 92.91 & 76.20 & 69.12 & 68.32 & -0.08 \\
+		DRF & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
+		CORM + DFR & 91.31 & 72.59 & 63.64 & 63.90 & -0.23 \\
+		\midrule
+		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
+		ACE + SWD (Ours) & 91.20 & \bf 38.27 & \bf 27.20 & \bf 33.60 & \bf 0.31 \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
 \subsection{Benchmark: Out-of-Distribution Generalization}
 
+\subsubsection{No Semantic Shift}
 To test how well the approach generalizes out of the training distribution, we evaluate the performance of the binary classification contrastive model on random samples ($n=1000$) from the \href{https://www.microsoft.com/en-us/download/details.aspx?id=54765}{Kaggle Cats and Dogs Dataset}:
@@
 	\centering
-	\begin{tabular}{c|c}
+	\begin{tabular}{c|ccc}
 		\toprule
-		\textbf{Method} & \textbf{Accuracy} \\
+		\textbf{Method} & \textbf{Accuracy} & \textbf{Avg Conf | Misclassified} & \textbf{Avg Conf} \\
 		\midrule
-		Cross-Entropy & 77.0\% \\
-		Ablated CE (Ours) & 75.7\% \\
-		Ablated CE + KLD (Ours) & \bf 83.4\% \\
+		Cross-Entropy 		& 77.0\% 	& TBR & TBR \\
+		Ablated CE (Ours) 	& 75.7\% 	& TBR & TBR \\
+		Ablated CE + KLD (Ours) & \bf 83.4\% 	& TBR & TBR \\
+		Ablated CE + SWD (Ours) & TBR & TBR & TBR \\
 		\bottomrule
@@
 
+\subsubsection{With Semantic Shift}
+
+To evaluate the performance of the approach against semantic shifts, we can use the \citet{soodimagenet} dataset. It
+
+The descriptors for abbreviations used in \ref{sood-wo-pretrain}, \ref{sood-w-pretrain} are as follows:
+\begin{table}[h]
+	\centering
+	\begin{tabular}{c|c}
+		\toprule
+		\textbf{Key} & \textbf{Descriptor} \\
+		\midrule
+		F1 & F-1 Score \\
+		ID & In-Distribution \\ 
+		E & Easy Test Set -- Small Semantic Shift \\
+		H & Hard Test Set -- Significant Semantic Shift \\
+		AC & Average Confidence \\
+		M & Misclassified Samples Only \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+\begin{table}[h]
+	\label{sood-wo-pretrain}
+	\caption{Without pre-training:}
+	\centering
+	\hspace*{-2.5em}
+	\begin{tabular}{c|ccccccc}
+		\toprule
+		\textbf{Method} & \textbf{F1 | ID} & \textbf{F1 | E} & \textbf{F1 | H} & \textbf{AC | M, E} & \textbf{AC | E} & \textbf{AC | M, H} & \textbf{AC | H} \\
+		\midrule
+		Cross-Entropy 		& TBR & 69.6\% & 52.9\% & .871 & .725 & .833 & .741 \\
+		Ablated CE (Ours) 	& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
+		Ablated CE + KLD (Ours) & TBR & 72.0\% & 55.9\% & .909 & .790  & .879 & .805 \\
+		Ablated CE + SWD (Ours) & TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+\begin{table}[h]
+	\label{sood-w-pretrain}
+	\caption{With pre-training:}
+	\centering
+	\hspace*{-2.5em}
+	\begin{tabular}{c|ccccccc}
+		\toprule
+		\textbf{Method} & \textbf{F1 | ID} & \textbf{F1 | E} & \textbf{F1 | H} & \textbf{AC | M, E} & \textbf{AC | E} & \textbf{AC | M, H} & \textbf{AC | H} \\
+		\midrule
+		Cross-Entropy 		& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
+		Ablated CE (Ours) 	& TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
+		Ablated CE + KLD (Ours) & TBR & 78.1\% & 61.7\% & .904 & .747 & .864 & .766 \\
+		Ablated CE + SWD (Ours) & TBR & TBR & TBR & TBR & TBR  & TBR & TBR \\
+		\bottomrule
+	\end{tabular}
+\end{table}
+
+\subsection{Qualitative Evals}
+
+\subsubsection{Adversarial Robustness}
+
+\subsubsection{Automated Image Circuit Discovery}
+
 \section{Upcoming / In Progress Tasks}
-
-\paragraph{Scaling the Approach} Training ResNet-50 on ImageNet and PASCAL VOC can allow us to evaluate if the approach scales effectively, with the performance gain not being constrained to the Oxford-IIIT Pets Dataset.
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{plainnat}
 \bibliography{references}

### 2025-02-03 src/data/hard_imagenet.py

@@ -33,7 +33,7 @@ class Dataset(torch.utils.data.Dataset):
         self.collect_mask_paths()
         # self.mask_paths = glob.glob(_MASK_ROOT + split+'/*')
         self.num_classes = 15
-        self.eval_mode = self.eval_mode
+        self.eval_mode = eval_mode
         self.ft = ft
 
     def standard_resize_center_crop(self, img, mask, resize_shape=const.IMAGE_SIZE):

### 2025-02-15 src/const.py
added salient imagenet data + loaders

@@ -20,7 +20,7 @@ N_WORKERS = 3
 N_CHANNELS = 3
 BBOX_MAP = False
 BINARY_CLS = True
-DATASET = 'hardimagenet'
+DATASET = 'salientimagenet'
 CAM_SIZE = (28, 28)
 VAL_CROP_SIZE = 224
 VAL_RESIZE_SIZE = 232
@@ -29,12 +29,14 @@ IMAGE_SIZE = (224, 224)
 SPLITS = ['train', 'valid', 'test']
 IMAGE_SHAPE = (N_CHANNELS,) + IMAGE_SIZE
 
+SAL_INET_MASK_THRESHOLD = .75
 HARD_INET_BALANCED_SUBSET = False
 HARD_INET_TRIM_MASKS = False
 HARD_INET_TRIM_FACTOR = 12
 HARD_INET_CENTER_BIAS = 2
 
 if DATASET == 'imagenet': N_CLASSES = 1000
+elif DATASET == 'salientimagenet': N_CLASSES = 1000
 elif DATASET == 'hardimagenet': N_CLASSES = 15
 elif DATASET == 'soodimagenet': N_CLASSES = 56
 elif DATASET == 'sbd': N_CLASSES = 20
@@ -51,7 +53,7 @@ MODIFY_BN = True
 # training
 DEVICE = 'cuda:0' if torch.cuda.is_available() else 'cpu'
 TRAIN_CUTOFF = 12600  # max training time in seconds
-LAMBDAS = [5, 0, 1, 5]  # means different things for different approaches
+LAMBDAS = [5, 0, 0, 1]  # means different things for different approaches
 POS_ONLY = False and DATASET == 'sbd'  # (multilabel only) restrict divergence loss to just positive classes
 GRAD_ACCUMULATION_STEPS = 1
 PORT = get_open_port()

### 2025-02-15 src/data/salient_imagenet.py

@@ -1,0 +1,88 @@
+#!/usr/bin/env python3
+
+from .hard_imagenet import Dataset as hardimagenet
+from src.utils import DataLoader, trim_mask
+import torchvision.transforms.v2 as T
+from PIL import Image
+from src import const
+import numpy as np
+import random
+import torch
+import json
+
+
+# code adapted from: https://github.com/singlasahil14/salient_imagenet/blob/main/inspection_utils.py
+class Dataset(torch.utils.data.Dataset):
+    def __init__(self, trim_masks=False, mask_threshold=const.SAL_INET_MASK_THRESHOLD, resize_size=256, crop_size=const.IMAGE_SIZE, device=const.DEVICE):
+        self.transform = T.Compose([
+            T.Resize(resize_size),
+            T.CenterCrop(crop_size),
+            T.ToImage(),
+            T.ToDtype(torch.float32, scale=True)
+        ])
+
+        self.dataset_path = const.DATA_DIR / 'salientimagenet'
+        self.mask_threshold = mask_threshold
+        self.trim_masks = trim_masks
+        self.device = device
+
+        self.core_features_dict = json.load(open(self.dataset_path / 'core_features_dict.json', 'r'))
+        self.wordnet_dict = json.load(open(self.dataset_path / 'wordnet_dict.json', 'r'))
+
+        images_class_dict = json.load(open(self.dataset_path / 'image_class_dict.json', 'r'))
+        self.image_names = list(images_class_dict.keys())
+        self.image_labels = list(images_class_dict.values())
+
+    def __len__(self):
+        return len(self.image_names)
+
+    def __getitem__(self, index):
+        image_name = self.image_names[index]
+        curr_image_path = const.DATA_DIR / 'imagenet' / 'images' / 'train' / self.wordnet_dict[str(self.image_labels[index])] / (image_name + '.JPEG')
+
+        image = Image.open(curr_image_path).convert("RGB")
+        image_tensor = self.transform(image)
+
+        all_mask = np.zeros(image_tensor.shape[1:])
+        for feature_index in self.core_features_dict[str(self.image_labels[index])]:
+            curr_mask_path = self.dataset_path / self.wordnet_dict[str(self.image_labels[index])] / ('feature_' + str(feature_index)) / (image_name + '.JPEG')
+            if not curr_mask_path.exists(): continue
+
[…]

### 2025-02-15 src/model/arch.py

@@ -6,7 +6,7 @@ import torchvision
 import torch
 
 
-# modified from: https://github.com/ptrblck/pytorch_misc/blob/master/batch_norm_manual.py
+# code modified from: https://github.com/ptrblck/pytorch_misc/blob/master/batch_norm_manual.py
 class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
     def __init__(self, num_features, eps=1e-5, momentum=0.1, affine=True, track_running_stats=True):
         super(ModifiedBN2d, self).__init__(num_features, eps, momentum, affine, track_running_stats)

### 2025-02-15 src/model/train.py

@@ -1,5 +1,6 @@
 #!/usr/bin/env python3
 
+from ..data.salient_imagenet import get_generators as salientimagenet
 from ..data.oxford_iiit_pet import get_generators as oxford_iiit_pet
 from ..data.hard_imagenet import get_generators as hardimagenet
 from ..data.soodimagenet import get_generators as soodimagenet
@@ -37,6 +38,7 @@ def configure(model_name):
 
     if 'hardimagenet' in const.MODEL_NAME: const.DATASET = 'hardimagenet'
     elif 'soodimagenet' in const.MODEL_NAME: const.DATASET = 'soodimagenet'
+    elif 'salientimagenet' in const.MODEL_NAME: const.DATASET = 'salientimagenet'
     elif 'imagenet' in const.MODEL_NAME: const.DATASET = 'imagenet'
     elif 'sbd' in const.MODEL_NAME: const.DATASET = 'sbd'
     elif 'oxford' in const.MODEL_NAME: const.DATASET = 'oxford'
@@ -67,6 +69,10 @@ def configure(model_name):
         const.BINARY_CLS = False
         const.HARD_INET_TRIM_MASKS = 'trim_masks' in const.MODEL_NAME
         const.HARD_INET_BALANCED_SUBSET = 'bsubset' in const.MODEL_NAME
+    elif const.DATASET == 'salientimagenet':
+        const.N_CLASSES = 1000
+        const.BINARY_CLS = False
+        const.HARD_INET_TRIM_MASKS = 'trim_masks' in const.MODEL_NAME
     elif const.DATASET == 'soodimagenet':
         const.N_CLASSES = 56
         const.BINARY_CLS = False
@@ -220,6 +226,7 @@ if __name__ == '__main__':
 
     if const.DATASET == 'imagenet': train, val, _ = imagenet()
     elif const.DATASET == 'soodimagenet': train, val, _ = soodimagenet('train')
+    elif const.DATASET == 'salientimagenet': train, val, _ = salientimagenet()
     elif const.DATASET == 'hardimagenet': train, val, _ = hardimagenet()
     elif const.DATASET == 'sbd': train, val, _ = sbd()
     else: train, val, test = oxford_iiit_pet()

### 2025-02-15 paper/main.tex
updated paper; background and motivations now brief key contributions
TODO: detailed subsections

@@
 \title{Interpretable Casual Interventions \\
-Encourage Core Risk Minimization}
+Discourage Spurious Learning}
 
@@
 
-The standard optimization pipeline for image classification predominantly consists of using a CNN-backbone with a fully-connected layer pointing to the classification targets.
+Neural architectures for image classification tasks predominantly consist of: a) convolutional approaches, i.e.~convolutional backbones with fully-connected layers transforming feature maps to logit space, and b) transformer based approaches \citep{alexey2020image}.
+Our exposition follows the first setup.
 
-This setup is vulnerable to spurious correlation caused by features like a common background, causing the model to learn `tricks', bypassing actually learning the target features.
-A simple example is illustrated in the second paragraph of the introduction in \cite{arjovsky2020invariantriskminimization}.
+\cite{singla2022salient} demonstrate that convolutional architectures are vulnerable to learning spurious correlations caused by features such as a common background, causing the model to learn `tricks', preventing it from actually learning the target features, a consequence of which is poor out-of-distribution generalization.
+A simple example of spurious learning is illustrated in the second paragraph of the introduction in \cite{arjovsky2020invariantriskminimization}.
 
-A popular post-hoc technique -- Class Activation Mappings \citep{zhou2016learning} -- helps diagnose this by generating a heatmap that visualizes the regions of the image used by the model to generate the prediction.
-Further research uncovered that the interpretation used by CAMs, and it's derivative GradCAMs were unfaithful as a consequence of the averaging step.
+A popular post-hoc interpretability technique -- Class Activation Mappings (CAMs) \citep{zhou2016learning} -- helps diagnose this by generating a heatmap that visualizes the regions of the image used by the model to generate the prediction.
+\cite{draelos2020use} uncovered that the interpretation offered by CAMs and it's derivative GradCAMs \citep{selvaraju2017grad} were unfaithful as a consequence of the averaging step.
 
-HiResCAM \citep{draelos2020use} is a variant of GradCAM \citep{selvaraju2017grad} that demonstrates that their visualization \textbf{provably recovers the input-dependent abnormality score} from the final classification.
+HiResCAMs \citep{draelos2020use} is a variant of GradCAMs \citep{selvaraju2017grad} that provides an interpretation that \textbf{provably recovers the input-dependent abnormality score} from the final classification.
 Therefore, there is a guarantee associated with the faithfulness of the visualization.
 
-This approach derives an interpretable training objective using HiResCAMs, which replaces and outperforms cross-entropy loss to build an intrinsically interpretable, performant model.
+We demonstrate that the normalizing behavior of softmax during the forward pass introduces conditions where an interpretation obtained from HiResCAMs fails to convey the true factors that contribute towards a prediction.
+To correct this, we propose an interpretability technique -- \textbf{ContrastiveCAMs} -- which reconstructs provably faithful activation mappings in the multiclass setting.
+
+We prove using ContrastiveCAMs that the default training behavior of Batch Normalization \citep{ioffe2015batch} \underline{breaks the locality principle of convolutional networks}, and instead motivate the use of Running Batch Normalization \citep{jermeyhowardlesson10}, computed causally to recover locality.
+
+Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image.
+We further introduce an interpretable training objective with ContrastiveCAMs, which is equivalent to Core Risk Minimization \citep{singla2022core}.
+
+Next, we present a ContrastiveCAM-based intervention to the model's logits, applied during training, which enables cross-entropy loss to implicitly optimize for multiple objectives.
+It must: a) \textbf{learn an invariance} to a user-specified non-target region \underline{despite any presence of strong spurious factors}, and b) \textbf{learn to generate contrast} within the user-specified target region to fulfill the classification task.
+
+Finally, we observe that the Cross-Entropy Loss only motivates generating contrast over regions that describe dataset-specific inter-class differences, and therefore propose distance based loss functions that outperforms Cross-Entropy Loss in both in-distribution and out-of-distribution settings.
 
@@
 		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
-		ACE + SWD (Ours) & 91.20 & \bf 38.27 & \bf 27.20 & \bf 33.60 & \bf 0.31 \\
+		ACE + SWD (Ours) & 91.20 & \bf 36.13 & \bf 27.20 & \bf 31.60 & \bf 0.28 \\
+		ACE + SWD + RevUPs (Ours) & 75.86 & \bf 10.40 & \bf 09.86 & \bf 13.19 & \bf 0.57 \\
 		\bottomrule
@@
 
-CutMix\cite{yun2019cutmix} is a batch-wise augmentation technique that encourages better regularization by a) ``cutting'' out a randomized rectangle (randomized portion remaining consistent across the batch) of a given image and b) ``mixing'' the cut-out with it's neighbor.
+CutMix \citep{yun2019cutmix} is a batch-wise augmentation technique that encourages better regularization by a) ``cutting'' out a randomized rectangle (randomized portion remaining consistent across the batch) of a given image and b) ``mixing'' the cut-out with it's neighbor.
 The corresponding labels are mixed by a randomly sampled parameter $\lambda$.
@@
 
-CutMix can be effectively adapted to work with our setup: given a cut-out, we motivate differential contrast; if non-target pixels are contained in the new image, irresepective of the label, contrast should be generated in favor of the non-target class for the non-target, foreground subset of pixels.
+CutMix can be effectively adapted to work with our setup: given a cut-out, we motivate differential contrast; if non-target pixels are contained in the new image, irrespective of the label, contrast should be generated in favor of the non-target class for the non-target, foreground subset of pixels.
 This affects the downstream prediction confidence, representative of the label being shifted from one-hot encoding to a randomized affine combination.

### 2025-02-16 src/model/arch.py
corrected batchnorm modification, added configuration for reverse upsampling using model name

@@ -28,8 +28,8 @@ class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
                 else:  # use exponential moving average
                     exponential_average_factor = self.momentum
 
-        mean = self.running_mean.clone()
-        var = self.running_var.clone()
+        fp_mean = self.running_mean.clone()
+        fp_var = self.running_var.clone()
 
         # calculate running estimates
         if self.training:
@@ -44,7 +44,7 @@ class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
                 self.running_var = exponential_average_factor * var * n / (n - 1)\
                     + (1 - exponential_average_factor) * self.running_var
 
-        input = (input - mean[None, :, None, None]) / (torch.sqrt(var[None, :, None, None] + self.eps))
+        input = (input - fp_mean[None, :, None, None]) / (torch.sqrt(fp_var[None, :, None, None] + self.eps))
         if self.affine:
             input = input * self.weight[None, :, None, None] + self.bias[None, :, None, None]
 
@@ -95,7 +95,7 @@ class Model(nn.Module):
         self.linear = nn.Linear(2048, const.N_CLASSES, bias=not is_contrastive)
         self.probabilities = nn.Identity() if logits_only else nn.Softmax(dim=1) if not multilabel else nn.Sigmoid()
 
-        if const.DATASET == 'imagenet' and const.PRETRAINED_BACKBONE:
+        if const.DATASET in ['imagenet', 'salientimagenet'] and const.PRETRAINED_BACKBONE:
             self.linear.weight = self.backbone.fc.weight
             if not is_contrastive: self.linear.bias = self.backbone.fc.bias
         self.backbone.fc = nn.Identity()
@@ -153,7 +153,7 @@ class Model(nn.Module):
 if __name__ == '__main__':
     import matplotlib.pyplot as plt
 
-    model = Model(disable_bn=False, upsampling_level=const.UPSAMPLING_LEVEL)
+    model = Model(disable_bn=False, modified_bn=True, upsampling_level=const.UPSAMPLING_LEVEL)
     print(model)
 
     x = torch.rand(1, *const.IMAGE_SHAPE, device=const.DEVICE, requires_grad=True)

### 2025-02-16 src/model/train.py

@@ -50,6 +50,7 @@ def configure(model_name):
     const.XL_BACKBONE = 'largemodel' in const.MODEL_NAME
     const.PRETRAINED_BACKBONE = 'pretrained' in const.MODEL_NAME
 
+    if ('revups' in const.MODEL_NAME and const.UPSAMPLING_LEVEL > 0) or ('revups' not in const.MODEL_NAME and const.UPSAMPLING_LEVEL < 0): const.UPSAMPLING_LEVEL *= -1
     if 'label_smoothing' not in const.MODEL_NAME: const.LABEL_SMOOTHING = 0
 
     if 'ablated_only' in const.MODEL_NAME: const.LAMBDAS[-1] = 0

### 2025-02-16 src/model/arch.py
reset affine params and running stats for modified batchnorm

@@ -31,18 +31,14 @@ class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
         fp_mean = self.running_mean.clone()
         fp_var = self.running_var.clone()
 
-        # calculate running estimates
         if self.training:
             mean = input.mean([0, 2, 3])
-            # use biased var in train
             var = input.var([0, 2, 3], unbiased=False)
             n = input.numel() / input.size(1)
+
             with torch.no_grad():
-                self.running_mean = exponential_average_factor * mean\
-                    + (1 - exponential_average_factor) * self.running_mean
-                # update running_var with unbiased var
-                self.running_var = exponential_average_factor * var * n / (n - 1)\
-                    + (1 - exponential_average_factor) * self.running_var
+                self.running_mean = exponential_average_factor * mean + (1 - exponential_average_factor) * self.running_mean
+                self.running_var = exponential_average_factor * var * n / (n - 1) + (1 - exponential_average_factor) * self.running_var
 
         input = (input - fp_mean[None, :, None, None]) / (torch.sqrt(fp_var[None, :, None, None] + self.eps))
         if self.affine:
@@ -102,6 +98,12 @@ class Model(nn.Module):
 
         self.to(self.device)
         if disable_bn: self.disable_batchnorms()
+
+        if modified_bn:
+            for x in self.modules():
+                if x._get_name() == 'ModifiedBN2d':
+                    x.reset_parameters()
+                    x.reset_running_stats()
 
     def disable_batchnorms(self):
         for x in self.modules():

### 2025-02-24 paper/main.tex
wrote a proof to demonstrate cross-entropy motivates learning from spurious correlations
added some drawings, probably useful later

@@
 
-\subsection{Observation: Cross-Entropy Motivates Learning from Spurious Correlations}
+\subsection{Cross-Entropy Motivates Learning from Spurious Correlations}
 
+\begin{theorem}\label{ce-spur-corr}
+	Cross-Entropy motivates learning from spurious correlations.
+\end{theorem}
+\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from Theorem \ref{classprobdiff}, we have:
+	\begin{align}
+		d &= [s_i - s_c]_i \\
+		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
+		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
+	\end{align}
+	As an aside: we note that when $i = c$, the contrastive map obtained is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
+\newline \\
+	Using our result from Theorem~\ref{classprobdiff}, we have:
+	\begin{gather}
+		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
+	\end{gather}
+	For a target class $c$, cross-entropy loss is defined as:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right) \\
+		\nabla_\theta H(\theta, \mathcal{D}) = \left(\frac{\partial H}{\partial \bm{s}}\right)^T \cdot \frac{\partial \bm{s}}{\partial \theta} = (\bm{s} - \bm{y})^T \cdot \frac{\partial \bm{s}}{\partial \theta} = -\left(\sum^{D_1, D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}} + \bm{y}_{i}\right)^T \cdot \frac{\partial \bm{s}}{\partial \theta}
+	\end{gather}
+	The gradient contribution from every given pixel $(j, k)$ from the attribution map is equivalent.
+This \underline{holds irrespective of whether the given pixel contains the target feature}; consequently, any region within the image can contribute towards the correct class.
+
+	If the region used to contribute towards the correct class contains a spurious factor, learning said spurious factor reduces cross-entropy loss.
+\end{proof}
+
+\subsubsection{Empirical Case Study -- Oxford-IIIT Pets}
 Using contrastive activation maps, we can evaluate the attention maps generated by a \underline{well-trained} model (ImageNet weights, fine-tuned to $99.9\%$ training accuracy, $99.6\%$ validation accuracy) on Oxford-IIIT Pets:
@@
 \end{theorem}
-\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from \ref{classprobdiff}, we have:
-	\begin{align}
-		d &= [s_i - s_c]_i \\
-		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
-		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
-	\end{align}
-	As an aside: we note that when $i = c$, the contrastive map obtained is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
-\newline \\
-	Using our result from \ref{classprobdiff}, we have:
-	\begin{gather}
-		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})}
-	\end{gather}
-	For a target class $c$, cross-entropy loss is defined as:
+\begin{proof} Following the setup from Theorem \ref{ce-spur-corr}, for a target class $c$, cross-entropy loss is defined as:
 	\begin{gather}

### 2025-03-05 paper/main.tex
updated proof for spurious-motivations in cross-entropy, fixed syncbatchnorm desyncs using `torch.compile`
plus minor dataloader syntax and eval path fixes

--- 2025-02-15 paper/main.tex
@@
 
-Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image.
+Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image (Section \ref{ce-motiv-spurious}).
 We further introduce an interpretable training objective with ContrastiveCAMs, which is equivalent to Core Risk Minimization \citep{singla2022core}.
 
-Next, we present a ContrastiveCAM-based intervention to the model's logits, applied during training, which enables cross-entropy loss to implicitly optimize for multiple objectives.
+Next, we present a ContrastiveCAM-based intervention to the model's logits, applied during training, which enables cross-entropy loss to implicitly optimize for multiple objectives (Theorems \ref{ace}, \ref{abce}).
 It must: a) \textbf{learn an invariance} to a user-specified non-target region \underline{despite any presence of strong spurious factors}, and b) \textbf{learn to generate contrast} within the user-specified target region to fulfill the classification task.
@@
 	\begin{gather}
-		d = [s_i - s_c]_i \\
+		\bm{d} = [s_i - s_c]_i \\
 		\bm{s} = [s_c + d_i]_i
@@
 
-\subsection{Observation: Cross-Entropy Motivates Learning from Spurious Correlations}
+\subsection{Cross-Entropy Motivates Learning from Spurious Correlations} \label{ce-motiv-spurious}
 
-Using contrastive activation maps, we can evaluate the attention maps generated by a \underline{well-trained} model (ImageNet weights, fine-tuned to $99.9\%$ training accuracy, $99.6\%$ validation accuracy) on Oxford-IIIT Pets:
+\begin{theorem}\label{ce-spur-corr}
+	Cross-Entropy motivates learning from spurious correlations.
+\end{theorem}
+\begin{proof} Using the definition of HiResCAMs, ContrastiveCAMs and the elementwise difference setup from Theorem \ref{classprobdiff}, we have:
+	\begin{align}
+		d &= [s_i - s_c]_i \\
+		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
+		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
+	\end{align}
+	As an aside: we note that when $i = c$, the contrastive map obtained is $\bm{0} \in \mathbb{R}^{D_1 \times D_2}$, and instead of $|C| - 1$ contrastive maps we now obtain $|C|$ contrastive maps.
+
+	Using our result from Theorem~\ref{classprobdiff}, we have:
+	\begin{gather}
+		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} = \sum_i \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)
+	\end{gather}
+	For a target class $c$, cross-entropy loss is defined as:
+	\begin{gather}
+		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \sum_i \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right) \right)
+	\end{gather}
+	Vectorizing parameters $\theta \in \mathbb{R}^P$ as well as the contrastive attribution maps $\tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}} \in \mathbb{R}^{D_1 D_2}$, we can use the above result to compute gradient contributions from each pixel within the attribution map:
+	\begin{align}
+		\nabla_\theta H &= \underbrace{\begin{bmatrix} \frac{\partial H}{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}} \end{bmatrix}_i}_{\mathbb{R}^{1 \times |C|}}
+			\cdot \underbrace{\begin{bmatrix} \begin{bmatrix} \frac{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}}{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}} \end{bmatrix}_{i}^T \end{bmatrix}_{j,k}}_{\mathbb{R}^{|C| \times D_1 D_2}}
+			\cdot \underbrace{\frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}}_{\mathbb{R}^{D_1D_2 \times P}} \\
+		&= - \begin{bmatrix} \frac{\text{exp}\left( \sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}\right)}{\sum_j \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, j)}}\right)} \end{bmatrix}_i \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta} \\
+			&= - \sigma\left( \begin{bmatrix} \sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \end{bmatrix}_i \right) \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}
+	\end{align}
+	Each element within the second term is constant $\forall j \in \mathbb{N}(D_1),~k \in \mathbb{N}(D_2)$, indicating the gradient contribution from every given pixel $(j, k)$ from the attribution map is equivalent.
+This \underline{holds irrespective of whether the given pixel contains the target feature}; consequently, any region within the image can contribute towards the correct class.
+\end{proof}
+	Let $\gamma \in [0, 1]$ be the proportion of the attribution map containing the target.
+The so-called `valid' gradient contributions will only come from the regions containing the target, and smaller targets from images taken from far away result in a smaller $\gamma$, and as a consequence a greater proportion of non-target regions are established as new priors over which the classification is determined.
+
+	If the region used to contribute towards the correct class contains a spurious factor with a high correlation across targets, this prior will motivate learning said spurious factor, which reduces cross-entropy loss at the cost of misrepresenting the true classification target.
+
+\subsubsection{Empirical Case Study -- Oxford-IIIT Pets}
+Using contrastive activation maps, we can evaluate the attention maps generated by a \underline{well-trained} ResNet-50 model (ImageNet weights, fine-tuned to $99.9\%$ training accuracy, $99.6\%$ validation accuracy) on Oxford-IIIT Pets:
 
@@
 \begin{theorem}\label{ace}
-	We can integrate backgroud invariance to the definition of cross-entropy using the following formulation:
+	We can integrate background invariance to the definition of cross-entropy using the following formulation:
 	\begin{gather}
@@
 \begin{theorem}\label{abce}
-	We can integrate backgroud invariance to the definition of binary cross-entropy using the following formulation:
+	We can integrate background invariance to the definition of binary cross-entropy using the following formulation:
 	\begin{gather}

### 2025-03-05 src/data/hard_imagenet.py

@@ -106,7 +106,7 @@ def get_generators():
 
     datasets = [Dataset(split=split, ft=True, balanced_subset=const.HARD_INET_BALANCED_SUBSET, trim_masks=const.HARD_INET_TRIM_MASKS, device='cpu') for split in const.SPLITS[:2]]
     dataloaders = *[DataLoader(dataset, shuffle=None if const.DDP else True, sampler=DistributedSampler(dataset) if const.DDP else None, num_workers=const.N_WORKERS,
-                               pin_memory=True, batch_size=const.BATCH_SIZE) for dataset, in datasets], None
+                               pin_memory=True, batch_size=const.BATCH_SIZE) for dataset in datasets], None
 
     const.SPLITS[1] = 'valid'
     return dataloaders

### 2025-03-05 src/evals/oxford/sample_contrastive_cams.py

@@ -26,7 +26,7 @@ def visualize(model, gen, norm=None):
         plt.imshow(F.interpolate(cam[None, None, :], const.IMAGE_SIZE, mode='bilinear')[0, 0].cpu(), cmap='jet', alpha=0.5, norm=norm)
 
     plt.tight_layout()
-    fig.savefig(const.DATA_DIR / 'evals' / f'{model.name}_cam.png')
+    fig.savefig(const.DATA_DIR / 'evals' / f'{model.name.replace("/", "_")}_cam.png')
     plt.show()
 
 

### 2025-03-05 src/model/arch.py

@@ -15,6 +15,7 @@ class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
         if input.dim() != 4:
             raise ValueError(f"expected 4D input (got {input.dim()}D input)")
 
+    @torch.compile
     def forward(self, input):
         self._check_input_dim(input)
 

### 2025-03-23 src/const.py
added dyt as drop-in batchnorm replacement, greater tolerance for faithfulness verifier when using lower-precision matmul

@@ -20,7 +20,7 @@ N_WORKERS = 3
 N_CHANNELS = 3
 BBOX_MAP = False
 BINARY_CLS = True
-DATASET = 'hardimagenet'
+DATASET = 'imagenet'
 CAM_SIZE = (14, 14)
 VAL_CROP_SIZE = 224
 VAL_RESIZE_SIZE = 232
@@ -47,8 +47,8 @@ RANDOMIZED_FLATTEN = False
 UPSAMPLING_LEVEL = 1  # -ve changes direction not magnitude
 XL_BACKBONE = False
 DISABLE_BN = False
-MODIFY_BN = False
 AFFINE_BN = False
+MODIFY_BN = None
 ACTIVATIONS = 'ReLU'
 
 # training

### 2025-03-23 src/model/arch.py

@@ -3,44 +3,15 @@
 from .resnet import Bottleneck
 from .activations import *
 from src import const
+from .norms import *
 from torch import nn
 import torchvision
 import torch
 
 
-# code modified from: https://github.com/ptrblck/pytorch_misc/blob/master/batch_norm_manual.py
-class ModifiedBN2d(torch.nn.modules.batchnorm._BatchNorm):
-    def __init__(self, num_features, eps=1e-5, momentum=0.1, affine=True, track_running_stats=True):
-        super(ModifiedBN2d, self).__init__(num_features, eps, momentum, affine, track_running_stats)
-        torch._dynamo.config.force_parameter_static_shapes = False
-        self.update_proxy_stats = False
-
-    def _check_input_dim(self, input):
-        if input.dim() != 4:
-            raise ValueError(f"expected 4D input (got {input.dim()}D input)")
-
-    @torch.compile(dynamic=True)
-    @staticmethod
-    def functional(input, running_mean, running_var, eps, weight, bias, affine):
-        input = (input - running_mean[None, :, None, None]) / (torch.sqrt(running_var[None, :, None, None] + eps))
-        if affine: input = input * weight[None, :, None, None] + bias[None, :, None, None]
-
-        return input
-
-    @torch.compiler.disable
-    def forward(self, input):
-        self._check_input_dim(input)
-
-        if self.update_proxy_stats:
-            self.running_mean = input.mean([0, 2, 3])
-            self.running_var = input.var([0, 2, 3], unbiased=False)
-
-        return ModifiedBN2d.functional(input, self.running_mean, self.running_var, self.eps, self.weight, self.bias, self.affine)
-
-
 class Model(nn.Module):
     def __init__(self, randomized_flatten=const.RANDOMIZED_FLATTEN, multilabel=False, logits_only=False, backbone_acts=const.ACTIVATIONS,
-                 disable_bn=const.DISABLE_BN, modified_bn=const.MODIFY_BN, register_backward_hook=False, hardinet_eval=False, xl_backbone=const.XL_BACKBONE,
+                 disable_bn=const.DISABLE_BN, modified_bn=None, register_backward_hook=False, hardinet_eval=False, xl_backbone=const.XL_BACKBONE,
[…]

### 2025-03-23 src/model/norms.py

@@ -1,0 +1,59 @@
+#!/usr/bin/env python3
+
+from torch import nn
+import torch
+
+
+# code copied from: https://github.com/jiachenzhu/DyT/blob/main/dynamic_tanh.py
+class DynamicTanh(nn.Module):
+    def __init__(self, normalized_shape, channels_last=False, alpha_init_value=0.5):
+        super().__init__()
+        self.normalized_shape = normalized_shape
+        self.alpha_init_value = alpha_init_value
+        self.channels_last = channels_last
+
+        self.alpha = nn.Parameter(torch.ones(1) * alpha_init_value)
+        self.weight = nn.Parameter(torch.ones(normalized_shape))
+        self.bias = nn.Parameter(torch.zeros(normalized_shape))
+
+    def forward(self, x):
+        x = torch.tanh(self.alpha * x)
+        if self.channels_last:
+            x = x * self.weight + self.bias
+        else:
+            x = x * self.weight[:, None, None] + self.bias[:, None, None]
+        return x
+
+    def extra_repr(self):
+        return f"normalized_shape={self.normalized_shape}, alpha_init_value={self.alpha_init_value}, channels_last={self.channels_last}"
+
+
+# code modified from: https://github.com/ptrblck/pytorch_misc/blob/master/batch_norm_manual.py
+class ModifiedBN2d(nn.modules.batchnorm._BatchNorm):
+    def __init__(self, num_features, eps=1e-5, momentum=0.1, affine=True, track_running_stats=True):
+        super(ModifiedBN2d, self).__init__(num_features, eps, momentum, affine, track_running_stats)
+        torch._dynamo.config.force_parameter_static_shapes = False
+        self.update_proxy_stats = False
+
+    def _check_input_dim(self, input):
+        if input.dim() != 4:
+            raise ValueError(f"expected 4D input (got {input.dim()}D input)")
+
+    # @torch.compile(dynamic=True)
+    @torch.compiler.disable  # use if no gradients are forwarded to running stats
+    @staticmethod
+    def functional(input, running_mean, running_var, eps, weight, bias, affine):
+        input = (input - running_mean[None, :, None, None]) / (torch.sqrt(running_var[None, :, None, None] + eps))
+        if affine: input = input * weight[None, :, None, None] + bias[None, :, None, None]
+
+        return input
+
+    @torch.compiler.disable
+    def forward(self, input):
+        self._check_input_dim(input)
+
+        if self.update_proxy_stats:
[…]

### 2025-03-23 src/model/train.py

@@ -48,14 +48,16 @@ def configure(model_name):
     const.USE_ZERO = 'zero' in const.MODEL_NAME and const.DDP
     const.EMA = 'ema' in const.MODEL_NAME
     const.DISABLE_BN = 'no_bn' in const.MODEL_NAME
-    const.MODIFY_BN = 'mod_bn' in const.MODEL_NAME
     const.XL_BACKBONE = 'largemodel' in const.MODEL_NAME
     const.PRETRAINED_BACKBONE = 'pretrained' in const.MODEL_NAME
+
+    if 'causalbn' in const.MODEL_NAME: const.MODIFY_BN = 'Causal'
+    if 'dytbn' in const.MODEL_NAME: const.MODIFY_BN = 'DyT'
 
     if 'elu' in const.MODEL_NAME: const.ACTIVATIONS = 'ELU'
     elif 'eeu' in const.MODEL_NAME: const.ACTIVATIONS = 'EEU'
     elif 'extsigmoid' in const.MODEL_NAME: const.ACTIVATIONS = 'ExtendedSigmoid'
-    elif 'dyt' in const.MODEL_NAME: const.ACTIVATIONS = 'DyT'
+    elif 'dytacts' in const.MODEL_NAME: const.ACTIVATIONS = 'DyT'
     else: const.ACTIVATIONS = 'ReLU'
 
     if ('revups' in const.MODEL_NAME and const.UPSAMPLING_LEVEL > 0) or ('revups' not in const.MODEL_NAME and const.UPSAMPLING_LEVEL < 0): const.UPSAMPLING_LEVEL *= -1

### 2025-03-24 paper/main.tex
added paper section for stricter core risk minimization objective formulation, plus minor typos and rephrasing

@@
 We demonstrate that the normalizing behavior of softmax during the forward pass introduces conditions where an interpretation obtained from HiResCAMs fails to convey the true factors that contribute towards a prediction.
-To correct this, we propose an interpretability technique -- \textbf{ContrastiveCAMs} -- which reconstructs provably faithful activation mappings in the multiclass setting.
+To correct this, we propose an interpretability technique -- \textbf{ContrastiveCAMs} (Section \ref{sec:define-cont-cams}) -- which reconstructs provably faithful activation mappings in the multiclass setting.
 
@@
 Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image (Section \ref{ce-motiv-spurious}).
-We further introduce an interpretable training objective with ContrastiveCAMs, which is equivalent to Core Risk Minimization \citep{singla2022core}.
+We further introduce an interpretable training objective with ContrastiveCAMs, which is a stricter counterpart of Core Risk Minimization \citep{singla2022core} (Section \ref{sec:constraining-optimization}).
 
@@
 \subsection{Defining the Contrastive Optimization Objective}
+\label{sec:define-cont-cams}
 
@@
 \end{gather}
-Ignoring the bias terms, our new optimization objective is:
+Ignoring the bias terms, the new optimization objective becomes:
 \begin{gather}
@@
 To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result.
-Therefore, the multiclass optimization objective under this paradigm is as follows:
+Therefore, the multiclass optimization objective under this construction reads:
 \begin{gather}
@@
 
-The primary distinction between the contrastive optimization objective over standard risk is that it preserves spatial information.
-We know at a pointwise level how each feature contributes to the final prediction.
+The primary distinction between the contrastive construction over standard risk is that it preserves spatial information.
+We know at a pointwise level how each feature within the contrastive map contributes to the final prediction.
 
-We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.
-The model through the process of training should learn to distinguish and subsequently ignore the background in determining the final prediction.
+\paragraph{Core Risk Minimization \citep{singla2022core}} proposes a generalization of Emprical Risk Minimization, the objective is to minimize prediction error while non-core regions are injected with gaussian noise:
+\begin{equation}
+	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]
+\end{equation}
+Here $\ell$ represents cross-entropy, $\bm{x}$ represents the input, $\bm{h}$ is the set of core features corresponding to the input encoded as a binary mask, $y$ is the label corresponding to the input and $f_\theta$ is the parameterized network.
+
+We can extend this to constrain the way optimization occurs.
+In principle, the objective is consistent -- we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.
+
+The model should, through the process of training learn to distinguish and subsequently ignore the background in determining the final prediction.
+Analogous to \citep{singla2022core}, we define the core-constrained optimization objective as follows:
+\begin{equation}\label{icrm}
+	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} \left[ \int_{\mathbb{R}^{C \times D_1 \times D_2}} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y)~d\bm{z} \right]
+\end{equation}
+
+For the experiments in this paper, we consider the segmentation mask of the image as a proxy for core features that causally determine the prediction.
+However, the same framework can be used to apply constrain learning over an arbitrary mask $\bm{h}$ (and subsequently invariance over mask $1 - \bm{h}$), which can vary depending on the task.
 
 % TODO: formal analysis -- empirical performance test on scale augmentation + norm-change in feature representation.
+%In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target motivates learning scale-robust representations of the target.
 
-In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target also motivates learning scale-robust representations of the target.
+In the following subsections, we outline cost functions designed to minimize (\ref{icrm}) using contrastive maps to directly learn invariance to spurious features while generating contrast over the target region.
 
@@
 
-\subsection{Training Detail}
+\subsection{Training Details}
 
@@
 		CORM & 92.91 & 76.20 & 69.12 & 68.32 & -0.08 \\
-		DRF & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
+		DFR & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
 		CORM + DFR & 91.31 & 72.59 & 63.64 & 63.90 & -0.23 \\

### 2025-03-24 paper/main.tex
added section on sliced wasserstein loss, reformulated ICRM as a constrained optimization problem
refactored contrastive cams as a custom variable to clean-up tex

@@
 \newtheorem{remark}[theorem]{Remark}
+
+\newcommand{\contcam}[2]{\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(#1, #2)}}
+\newcommand{\vectorproj}[2][]{\text{proj}_{\vect{#1}}\vect{#2}}
+\newcommand{\vect}{\mathbf}
 
@@
 Per \cite{draelos2020use}, a HiResCAM is defined as follows:
-\begin{gather}
+\begin{equation}
 	\tilde{\mathcal{\bm{A}}}_c^{\text{HiResCAM}} = \sum^F_{f=1} \frac{\partial s_c}{\partial \bm{A}^f} \odot \bm{A}^f
-\end{gather}
+\end{equation}
 
@@
 For the CAM architecture, \cite{draelos2020use} (section 3.6.2) shows that:
-\begin{gather}
+\begin{equation}
 	\sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} \\
 	s_c = \frac{\bm{w}_c \bm{A}_F}{D_1 D_2} + b_c = \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c
-\end{gather}
+\end{equation}
 Therefore, we can recreate the entire input-dependent part of the score using HiResCAMs. Doing this for each class $c$, we obtain the bias-removed logit vector $\in \mathbb{R}^{c}$.
@@
 \begin{proof} Individual class probabilities for logit vector $\bm{s}$ are defined as:
-	\begin{gather}
+	\begin{equation}
 		\sigma_c(\bm{s}) = \frac{e^{s_c}}{\sum_i e^{s_i}}
-	\end{gather}
+	\end{equation}
 	We define our logit vector in terms of the elementwise difference to a target class $c$:
-	\begin{gather}
+	\begin{equation}
 		\bm{d} = [s_i - s_c]_i \\
 		\bm{s} = [s_c + d_i]_i
-	\end{gather}
+	\end{equation}
 	Based on this definition, class probabilities can now be computed as:
@@
 \begin{corollary} Since class probabilities depend on logit differences, the class probabilities are identitical for any two pairs of logit vectors with identical logit differences.
-	\begin{gather}
+	\begin{equation}
 		\frac{e^{s_c}}{\sum_i e^{s_i}} = \frac{e^{s_c+r}}{\sum_i e^{s_i+r}} \forall r \in \mathbb{R}
-	\end{gather}
+	\end{equation}
 \end{corollary}
@@
 For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
-\begin{gather}
+\begin{equation}
 	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
-\end{gather}
+\end{equation}
 Ignoring the bias terms, the new optimization objective becomes:
-\begin{gather}
-	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} ~~\text{ where }~~ \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c')} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
-\end{gather}
+\begin{equation}
+	\max_{\theta} \sum^{D_1,D_2}_{i,j} \contcam{c}{c'}{}_{, i, j} \quad \text{where} \quad \contcam{c}{c'} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
+\end{equation}
 This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information in the form of the activation map instead of flattening the feature map.
@@
 	Given a set of classes $C$ with $c_t$ being the target class and $c_{t'}$ being a sampled non-target class, we define ContrastiveCAMs as follows:
-	\begin{gather}
-		\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c_t, c_{t'})} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c \setminus c_t}
-	\end{gather}
+	\begin{equation}
+		\contcam{c_t}{c_{t'}} := \left\{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c_{t'}}^{\text{HiResCAM}}\right\}^{|C|-1}_{c_{t'} \in c \setminus c_t}
+	\end{equation}
 \end{definition}
@@
 Therefore, the multiclass optimization objective under this construction reads:
-\begin{gather}
+\begin{equation}
 	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|C| - 1)
-\end{gather}
+\end{equation}
 
@@
 		&= \left[\sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - \sum^{D_1,D_2}_{j,k} \tilde{\mathcal{\bm{A}}}_{c,j,k}^{\text{HiResCAM}}\right]_i \\
-		&= \left[ -\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right]_i
+		&= \left[ -\sum \contcam{c}{i} \right]_i
 	\end{align}
@@
 	Using our result from Theorem~\ref{classprobdiff}, we have:
-	\begin{gather}
-		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} = \sum_i \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)
-	\end{gather}
+	\begin{equation}
+		\sigma_c(\bm{s}) = \frac{1}{\sum_i e^{d_i}} = \frac{1}{\sum_i \text{exp}({-\sum \contcam{c}{i}})} = \sum_i \text{exp}\left({\sum \contcam{c}{i}}\right)
+	\end{equation}
 	For a target class $c$, cross-entropy loss is defined as:
-	\begin{gather}
-		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \sum_i \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right) \right)
-	\end{gather}
+	\begin{equation}
+		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \sum_i \text{exp}\left({\sum \contcam{c}{i}}\right) \right)
+	\end{equation}
 	Vectorizing parameters $\theta \in \mathbb{R}^P$ as well as the contrastive attribution maps $\tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}} \in \mathbb{R}^{D_1 D_2}$, we can use the above result to compute gradient contributions from each pixel within the attribution map:
@@
 			\cdot \underbrace{\frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}}_{\mathbb{R}^{D_1D_2 \times P}} \\
-		&= - \begin{bmatrix} \frac{\text{exp}\left( \sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}\right)}{\sum_j \text{exp}\left({\sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, j)}}\right)} \end{bmatrix}_i \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta} \\
-			&= - \sigma\left( \begin{bmatrix} \sum \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \end{bmatrix}_i \right) \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}
+		&= - \begin{bmatrix} \frac{\text{exp}\left( \sum \contcam{c}{i}\right)}{\sum_j \text{exp}\left({\sum \contcam{c}{j}}\right)} \end{bmatrix}_i \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta} \\
+			&= - \sigma\left( \begin{bmatrix} \sum \contcam{c}{i} \end{bmatrix}_i \right) \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}
 	\end{align}
@@
 \begin{equation}
-	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]
+	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]
 \end{equation}
@@
 The model should, through the process of training learn to distinguish and subsequently ignore the background in determining the final prediction.
-Analogous to \citep{singla2022core}, we define the core-constrained optimization objective as follows:
-\begin{equation}\label{icrm}
-	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} \left[ \int_{\mathbb{R}^{C \times D_1 \times D_2}} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y)~d\bm{z} \right]
+Analogous to \citep{singla2022core}, we define the contrastive-constrained optimization objective as follows:
+\begin{equation}\label{ccrm}
+	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} \left[ \ell(f_\theta(\bm{x}, y) \right] \quad \text{s.t.} \quad \sum_{c' \in C} \| \contcam{c}{c'} \odot (1 - h) \| = 0
 \end{equation}
@@
 
-In the following subsections, we outline cost functions designed to minimize (\ref{icrm}) using contrastive maps to directly learn invariance to spurious features while generating contrast over the target region.
+In the following subsections, we outline a family of cost functions for binary, mutliclass and mutliple-class classification, designed to minimize (\ref{ccrm}) using contrastive maps to learn invariance to spurious features while generating contrast over the target region.
 
@@
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)}))}_{\text{minimize divergence from target shape}} \\
-		&+ \underbrace{\frac{\lambda_4}{(|C| - 1) \left(D_1 D_2 - \sum^{D_1,D_2}_{i,j}h_{i,j}\right)} || (1 - h) \odot \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)}||^2_F}_{\text{minimize background influence}} \\
-		&- \underbrace{\frac{\lambda_5}{(|C| - 1) \sum^{D_1,D_2}_{i,j}h_{i,j}} \sum^{|C|-1,D_1,D_2}_{i,j,k} \left(\tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c,i)} \odot h\right)_{j,k} }_{\text{maximize foreground influence}}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \contcam{c}{i}))}_{\text{minimize divergence from target shape}} \\
+		&+ \underbrace{\frac{\lambda_4}{(|C| - 1) \left(D_1 D_2 - \sum^{D_1,D_2}_{i,j}h_{i,j}\right)} || (1 - h) \odot \contcam{c}{i}||^2_F}_{\text{minimize background influence}} \\
+		&- \underbrace{\frac{\lambda_5}{(|C| - 1) \sum^{D_1,D_2}_{i,j}h_{i,j}} \sum^{|C|-1,D_1,D_2}_{i,j,k} \left(\contcam{c}{i} \odot h\right)_{j,k} }_{\text{maximize foreground influence}}
 	\end{split}
@@
 	We can integrate background invariance to the definition of cross-entropy using the following formulation:
-	\begin{gather}
-		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}|\right)} \right)
-	\end{gather}
+	\begin{equation}
+		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum h \odot \contcam{c}{i} + \sum |(1-h) \odot \contcam{c}{i}|\right)} \right)
+	\end{equation}
 \end{theorem}
 \begin{proof} Following the setup from Theorem \ref{ce-spur-corr}, for a target class $c$, cross-entropy loss is defined as:
-	\begin{gather}
-		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum^{}\tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}})} \right)
-	\end{gather}
+	\begin{equation}
+		H(\theta, \mathcal{D}) = -\log \left( \sigma_c(\bm{s}) \right) = -\log \left( \frac{1}{\sum_i \text{exp}({-\sum \contcam{c}{i}})} \right)
+	\end{equation}
 	We can break down each contrastive map to foreground and background components:
-	\begin{gather}
-		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} - \sum (1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}\right)} \right)
-	\end{gather}
+	\begin{equation}
+		H(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum h \odot \contcam{c}{i} - \sum (1-h) \odot \contcam{c}{i}}\right)} \right)
+	\end{equation}
 	The current loss formulation motivates improving contrast using either the foreground or the background.
 We penalize contrast on the background, motivating learning invariance to noise in the form of the background.
-	\begin{gather}
-		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
-	\end{gather}
+	\begin{equation}
+		\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left({-\sum h \odot \contcam{c}{i} + \sum |(1-h) \odot \contcam{c}{i}}|\right)} \right)
+	\end{equation}
 	This gives us our ablated cross-entropy formulation.
@@
 
-\begin{gather}
+\begin{equation}
 	\begin{split}
-		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \tilde{\mathcal{\bm{A}}}^{\textit{contrastive}}_{(c, i)})) \\
-		&- \lambda_4 \log \left( \frac{1}{\sum_i \text{exp}\left({-\sum^{}h \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |(1-h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}}|\right)} \right)
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \contcam{c}{i})) \\
+		&- \lambda_4 \log \left( \frac{1}{\sum_i \text{exp}\left({-\sum h \odot \contcam{c}{i} + \sum |(1-h) \odot \contcam{c}{i}}|\right)} \right)
 	\end{split}
-\end{gather}
+\end{equation}
 
@@
 	We can integrate background invariance to the definition of binary cross-entropy using the following formulation:
-	\begin{gather}
+	\begin{equation}
 		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
-	\end{gather}
+	\end{equation}
 \end{theorem}
@@
 Therefore, it's definition is retained.
-	\begin{gather}
+	\begin{equation}
 		H(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
-	\end{gather}
+	\end{equation}
 	The current formulation motivates activating either the foreground or the background for positive classification, and motivates de-activating every pixel of the non-positive class.
 We penalize activation on the background for the positive class only:
-	\begin{gather}
+	\begin{equation}
 		\mathcal{J}(\theta, \mathcal{D}) = - y_i \log \left(\sigma\left(h \odot \sum_{j,k}\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}} - (1-h) \odot \sum_{j,k}|\tilde{\mathcal{\bm{A}}}_{i,j,k}^{\text{HiResCAM}}|\right) \right) - (1 - y_i) \log \left( \bm{s}_i \right)
-	\end{gather}
+	\end{equation}
 	This gives us the ablated binary cross-entropy formulation.
@@
 Similar to ACE, we define a divergence term for the target class to motivate learning the foreground as part of the training process.
-\begin{gather}
+\begin{equation}
 	\begin{split}
@@
 	\end{split}
-\end{gather}
+\end{equation}
 
 \subsubsection{Spatial Sliced Wasserstein Loss}
+
+The application of Optimal Transport theory as a robust metric between distributions enables vast applications within image processing \citep{papadakis2015optimal}.
+Robust gradients in turn improve convergence, which we leverage by adapting the spatial sliced wasserstein loss proposed in \citet{heitz2021sliced} to learn the optimal contrastive map as a regularizer term over (\ref{ace}).
+\begin{equation}
+	\begin{split}
+		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \mathbb{E}_{\mathbf{d} \sim \mathcal{N}(0, 1)}\left[\left\| \text{sort}(\vectorproj[d]{(\text{concat}(\contcam{c}{c'}, \bm{s}}))) - \text{sort}(\vectorproj[d]{\text{concat}(h, s)}) \right\|^2 \right] \\
+		&-\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum h \odot \contcam{c}{i} + \sum |(1-h) \odot \contcam{c}{i}|\right)} \right)
+	\end{split}
+\end{equation}
+Where $\mathbf{s}$ is a spatial prior that ensures learned distributions correspond to spatial regions that contain the target.
+In practice, we use sample $\approx 20$ directions, and exclude over-activation in target regions when computing the norm.
+
+We observe that using divergence / distance based regularization to learn segmentation maps as an implicit part of optimization improves the generalization performance of the network.
+We hypothesize this being a consequence of inducing contrast over regions beyond just those most relevant for classification within dataset-specific samples.
+We present empricial evidence to support this claim across all our benchmarks in Section \ref{results}.
 
@@
 \section{Results}
+\label{results}
 
@@
 \subsection{Benchmark: Hard-Imagenet}
-
-\begin{table}[H]
+\begin{table}[h!]
+	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
 	\centering
-	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
+	\hspace*{-4em}
 	\begin{tabular}{c|c|ccc|c}
@@
 Let segmentation mask $h$ take the following form:
-\begin{gather*}
+\begin{equation*}
 	h := \begin{cases}
@@
 	\end{cases}
-\end{gather*}
+\end{equation*}
 
 Then, Ablated Cross Entropy (\ref{ace}) is formulated as follows:
-\begin{gather}
-	\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum \mathbbm{1}_{c}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} + \sum |\mathbbm{1}_{-1}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)}| + \sum \mathbbm{1}_{i}(h) \odot \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, i)} \right)} \right)
-\end{gather}
+\begin{equation}
+	\mathcal{J}(\theta, \mathcal{D}) = -\log \left( \frac{1}{\sum_i \text{exp}\left(-\sum \mathbbm{1}_{c}(h) \odot \contcam{c}{i} + \sum |\mathbbm{1}_{-1}(h) \odot \contcam{c}{i}| + \sum \mathbbm{1}_{i}(h) \odot \contcam{c}{i} \right)} \right)
+\end{equation}
 

### 2025-03-27 paper/main.tex
updated paper to introduce contrast-constrained optimization as a generalization of core risk minimization, minor code consistency improvements

@@
 \newcommand{\vect}{\mathbf}
+
+\DeclareMathOperator{\Tr}{Tr}
+\DeclareMathOperator{\Cov}{Cov}
+\DeclareMathOperator{\Concat}{Concat}
+\DeclareMathOperator*{\argsup}{arg\,sup}
+\DeclareMathOperator*{\arginf}{arg\,inf}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\indep}{\perp \!\!\! \perp}
 
@@
 	\begin{align}
-		\nabla_\theta H &= \underbrace{\begin{bmatrix} \frac{\partial H}{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}} \end{bmatrix}_i}_{\mathbb{R}^{1 \times |C|}}
-			\cdot \underbrace{\begin{bmatrix} \begin{bmatrix} \frac{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}}{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}} \end{bmatrix}_{i}^T \end{bmatrix}_{j,k}}_{\mathbb{R}^{|C| \times D_1 D_2}}
-			\cdot \underbrace{\frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}}_{\mathbb{R}^{D_1D_2 \times P}} \\
+		\nabla_\theta H &= \begin{bmatrix} \frac{\partial H}{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}} \end{bmatrix}_i
+			\cdot \begin{bmatrix} \begin{bmatrix} \frac{\partial \sum \tilde{\mathcal{\bm{A}}}_{(c, i)}^{\text{contrastive}}}{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}} \end{bmatrix}_{i}^T \end{bmatrix}_{j,k}
+			\cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta} \\
 		&= - \begin{bmatrix} \frac{\text{exp}\left( \sum \contcam{c}{i}\right)}{\sum_j \text{exp}\left({\sum \contcam{c}{j}}\right)} \end{bmatrix}_i \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta} \\
-			&= - \sigma\left( \begin{bmatrix} \sum \contcam{c}{i} \end{bmatrix}_i \right) \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}
+		\nabla_\theta H &= - \sigma\left( \begin{bmatrix} \sum \contcam{c}{i} \end{bmatrix}_i \right) \cdot \bm{1}_{|C| \times D_1 D_2} \cdot \frac{\partial \tilde{\mathcal{\bm{A}}}_{(c, i), j, k}^{\text{contrastive}}}{\partial \theta}
 	\end{align}
@@
 \paragraph{Core Risk Minimization \citep{singla2022core}} proposes a generalization of Emprical Risk Minimization, the objective is to minimize prediction error while non-core regions are injected with gaussian noise:
-\begin{equation}
-	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]
+\begin{equation}\label{crm}
+	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} [ \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]]
 \end{equation}
@@
 
-We can extend this to constrain the way optimization occurs.
+\paragraph{Contrast-Constrained Risk Minimization} We extend this notion to constrain the way optimization occurs.
 In principle, the objective is consistent -- we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.
 
-The model should, through the process of training learn to distinguish and subsequently ignore the background in determining the final prediction.
-Analogous to \citep{singla2022core}, we define the contrastive-constrained optimization objective as follows:
+The model should, through the process of training learn to distinguish and subsequently ignore the non-target regions in determining the final prediction.
+Analogous to \citep{singla2022core}, we define the contrast-constrained optimization objective as follows:
 \begin{equation}\label{ccrm}
-	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} \left[ \ell(f_\theta(\bm{x}, y) \right] \quad \text{s.t.} \quad \sum_{c' \in C} \| \contcam{c}{c'} \odot (1 - h) \| = 0
+	\hat{\theta} = \argmin_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} \left[ \ell(f_\theta(\bm{x}, y) \right] \quad \text{s.t.} \quad \sum_{c' \in C} \| \contcam{c}{c'} \odot (1 - \bm{h}) \| = 0
 \end{equation}
+In comparing this objective with (\ref{crm}), notice that when the desired constraint is satisfied, obtaining invariance to the non-target regions implies invariance to ablations made to the non-target regions:
+\begin{equation}
+	f_{\hat{\theta}}(\bm{x} + \bm{z} \odot (1 - \bm{h})) = f_{\hat{\theta}}(\bm{x})~\forall z \sim \mathcal{N}(0, \sigma^2 \bm{I})
+\end{equation}
+Further, this implies that:
+\begin{equation}
+	\mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} [\ell(f_{\hat{\theta}}(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]] =
+	\mathbb{E}_{(\bm{x}, \bm{h}, y) \sim \mathcal{D}} [ \ell(f_{\hat{\theta}}(\bm{x}), y) ]
+\end{equation}
+Revealing Contrast-Constrained Risk Minimization as a generalization of Core Risk Minimization.
 
 For the experiments in this paper, we consider the segmentation mask of the image as a proxy for core features that causally determine the prediction.
-However, the same framework can be used to apply constrain learning over an arbitrary mask $\bm{h}$ (and subsequently invariance over mask $1 - \bm{h}$), which can vary depending on the task.
+However, the same framework can be used to apply constrained learning over an arbitrary mask $\bm{h}$ (consequently, invariance over mask $1 - \bm{h}$), which can vary depending on the task.
 
@@
 
-In the following subsections, we outline a family of cost functions for binary, mutliclass and mutliple-class classification, designed to minimize (\ref{ccrm}) using contrastive maps to learn invariance to spurious features while generating contrast over the target region.
+In the following subsections, we outline a family of cost functions for binary, mutliclass and mutliple-class classification, designed to minimize (\ref{ccrm}) using contrastive maps to learn invariance to spurious features while generating contrast over the target region in favor of the correct class.
 
@@
 		\mathcal{\bm{J}}(\theta, \mathcal{D}) := &\ \frac{1}{|C|} \sum^{|C|}_{i} \underbrace{\lambda_1 D_{KL}(\sigma_{\text{softmax}}(\lambda_3 h)\ ||\ \sigma_{\text{softmax}}(\lambda_2 \contcam{c}{i}))}_{\text{minimize divergence from target shape}} \\
-		&+ \underbrace{\frac{\lambda_4}{(|C| - 1) \left(D_1 D_2 - \sum^{D_1,D_2}_{i,j}h_{i,j}\right)} || (1 - h) \odot \contcam{c}{i}||^2_F}_{\text{minimize background influence}} \\
+		&+ \underbrace{\frac{\lambda_4}{(|C| - 1) \left(D_1 D_2 - \sum^{D_1,D_2}_{i,j}h_{i,j}\right)} || (1 - \bm{h}) \odot \contcam{c}{i}||^2_F}_{\text{minimize background influence}} \\
 		&- \underbrace{\frac{\lambda_5}{(|C| - 1) \sum^{D_1,D_2}_{i,j}h_{i,j}} \sum^{|C|-1,D_1,D_2}_{i,j,k} \left(\contcam{c}{i} \odot h\right)_{j,k} }_{\text{maximize foreground influence}}

### 2025-03-27 src/const.py

@@ -45,11 +45,11 @@ else: N_CLASSES = 2 if BINARY_CLS else 37
 PRETRAINED_BACKBONE = False
 RANDOMIZED_FLATTEN = False
 UPSAMPLING_LEVEL = 1  # -ve changes direction not magnitude
+ACTIVATIONS = 'ReLU'
 XL_BACKBONE = False
 DISABLE_BN = False
 AFFINE_BN = False
 MODIFY_BN = None
-ACTIVATIONS = 'ReLU'
 
 # training
 DEVICE = 'cuda:0' if torch.cuda.is_available() else 'cpu'

### 2025-03-27 src/model/arch.py

@@ -86,7 +86,7 @@ class Model(nn.Module):
                         self.affine = False
 
         self.to(self.device)
-        self.initialize_and_verify()
+        if not hardinet_eval: self.initialize_and_verify()
 
     def disable_batchnorms(self):
         for x in self.modules():
@@ -135,7 +135,7 @@ class Model(nn.Module):
         return (segmentation_map.values > self.segmentation_threshold).to(torch.uint8) * (segmentation_map.indices + 1).to(torch.uint8)
 
     def get_contrastive_cams(self, y, cams):
-        return torch.index_select(cams.view(-1, *cams.shape[2:]), 0, y.argmax(1) + (torch.arange(cams.size(0), device=const.DEVICE) * cams.size(1))).repeat(1, cams.size(1), 1).view(*cams.shape) - cams
+        return torch.index_select(cams.view(-1, *cams.shape[2:]), 0, y.argmax(1) + (torch.arange(cams.size(0), device=self.device) * cams.size(1))).repeat(1, cams.size(1), 1).view(*cams.shape) - cams
 
     def _bp_free_hi_res_cams(self):  # required to obtain gradients on self.linear.weight
         return (self.linear.weight @ self.feature_rect.flatten(2)).unflatten(2, self.feature_rect.shape[2:]) / self.feature_rect.shape[-1]**2
@@ -159,7 +159,7 @@ class Model(nn.Module):
                 module.reset_running_stats()
                 module.update_proxy_stats = True
 
-        for X, y in gen: self(X.to(const.DEVICE))
+        for X, y in gen: self(X.to(self.device))
 
         for module in self.modules():
             if module._get_name() in ['ModifiedBN2d', 'BatchNorm2d']:

### 2025-03-27 src/model/norms.py

@@ -40,7 +40,7 @@ class ModifiedBN2d(nn.modules.batchnorm._BatchNorm):
             raise ValueError(f"expected 4D input (got {input.dim()}D input)")
 
     # @torch.compile(dynamic=True)
-    @torch.compiler.disable  # use if no gradients are forwarded to running stats
+    @torch.compiler.disable  # needed if gradients are forwarded to running stats
     @staticmethod
     def functional(input, running_mean, running_var, eps, weight, bias, affine):
         input = (input - running_mean[None, :, None, None]) / (torch.sqrt(running_var[None, :, None, None] + eps))

### 2025-04-07 paper/main.tex
overwrite running statistics for causal batchnorm only, reverted SWD to p=2, updated paper to NIPS 2025 style file and added additional benchmarks

@@
 % ready for submission
-\usepackage{neurips_2024}
+\usepackage{neurips_2025}
 
@@
 		\midrule
+		DyTBN (Baseline) & 93.07 & 76.67 & 71.73 & 68.27 & -0.22 \\
+		\midrule
 		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
@@
 		ACE + SWD + RevUPs (Ours) & 75.86 & \bf 10.40 & \bf 09.86 & \bf 13.19 & \bf 0.57 \\
+		ACE + SWD + DyTBN (Ours) & 87.73 & 45.86 & 34.80 & 40.93 & 0.16 \\
+		\midrule
+		Gold Standard & 100.00 & 06.67 & 06.67 & 06.67 & 1.00 \\
 		\bottomrule

### 2025-04-07 src/model/activations.py

@@ -17,9 +17,9 @@ class LazyDyT(nn.modules.lazy.LazyModuleMixin, nn.Module):
         if self.has_uninitialized_params():
             with torch.no_grad():
                 channels = input.size(1)
-                self.alpha = nn.Parameter(torch.full((channels,), float(self.init_alpha)))
-                self.gamma = nn.Parameter(torch.ones(channels))
-                self.beta = nn.Parameter(torch.zeros(channels))
+                self.alpha = nn.Parameter(torch.full((channels,), float(self.init_alpha), device=self.device))
+                self.gamma = nn.Parameter(torch.ones(channels, device=self.device))
+                self.beta = nn.Parameter(torch.zeros(channels, device=self.device))
 
     @torch.compile(dynamic=True)
     def forward(self, x):

### 2025-04-07 src/model/arch.py

@@ -82,9 +82,11 @@ class Model(nn.Module):
         self.backbone = torch.compile(self.backbone, backend='cudagraphs')
 
         if disable_bn: self.disable_batchnorms()
-        if modified_bn:
+        if modified_bn or backbone_acts == 'DyT':
             for x in self.modules():
-                if x._get_name() == 'ModifiedBN2d':
+                if x._get_name() == 'LazyDyT':
+                    x.device = self.device
+                elif x._get_name() == 'ModifiedBN2d':
                     x.reset_parameters()
                     x.reset_running_stats()
 
@@ -113,7 +115,7 @@ class Model(nn.Module):
     def initialize_and_verify(self):
         with torch.no_grad():
             x = torch.randn(100, *const.IMAGE_SHAPE, device=self.device)
-            if self.modified_bn: self.overwrite_tracked_statistics(((x, None),))
+            if self.modified_bn == 'Causal': self.overwrite_tracked_statistics(((x, None),))
 
             logits, cam = self(x)
             cam_logits = cam.view(*cam.shape[:2], -1).sum(2)

### 2025-04-07 src/model/loss.py

@@ -65,7 +65,7 @@ class ContrastiveLoss(nn.Module):
             divergence[(1 - target_mask[:, 1]).to(torch.bool).unsqueeze(1).repeat(1, n_directions, 1) & (divergence > 0)] = 0
         else:   # (pos_only & multilabel) | contrastive
             divergence[(sorted_mask_projections != 0) & (divergence < 0)] = 0
-        return divergence.abs().mean()
+        return divergence.pow(2).mean()
 
     def wasserstein(self, cc, fg_mask, y, y_pred):
         fg_mask = fg_mask.to(torch.float)

### 2025-04-07 src/model/overfit.py

@@ -18,7 +18,7 @@ import sys
 if __name__ == '__main__':
     multilabel = len(sys.argv) == 2
     data = sbd(mode='segmentation') if multilabel else oxford_iiit_pet('train')
-    model = Model(multilabel=multilabel, backbone_acts='ELU', modified_bn='Causal', logits_only=True)
+    model = Model(multilabel=multilabel, backbone_acts='ELU', modified_bn='DyT', load_pretrained_weights=False, logits_only=True)
     optim = torch.optim.Adam(model.parameters(), lr=1E-4)
     criterion = ContrastiveLoss(model.get_contrastive_cams, multilabel=multilabel, divergence=const.DIVERGENCE, pos_only=const.POS_ONLY, pos_weight=data.reweight if multilabel else None)
     # criterion = torch.nn.BCEWithLogitsLoss(pos_weight=data.reweight)
@@ -53,7 +53,7 @@ if __name__ == '__main__':
         else: print(loss.item(), F.binary_cross_entropy_with_logits(y_pred[0], y[1], pos_weight=data.reweight).detach().item())
 
         optim.step()
-        if model.modified_bn: model.overwrite_tracked_statistics(((benchmark_batch, None),))
+        if model.modified_bn == 'Causal': model.overwrite_tracked_statistics(((benchmark_batch, None),))
 
     print(y_pred[0].detach(), y[1])
 

### 2025-04-07 src/model/train.py

@@ -164,7 +164,8 @@ def fit(model, optimizer, scheduler, criterion, train, val, is_multilabel=False,
 
                         del batch_loss
                         torch.cuda.empty_cache()
-            except (KeyboardInterrupt, torch.OutOfMemoryError):
+            except (KeyboardInterrupt, torch.OutOfMemoryError) as exception:
+                print(exception)
                 break
 
             for split in const.SPLITS[:2]: metrics[f'{split}_acc'] = (sync_and_compute(metrics[f'{split}_acc']) if const.DDP else metrics[f'{split}_acc'].compute()).item()
@@ -261,7 +262,8 @@ if __name__ == '__main__':
         benchmark_batch = torch.vstack([benchmark_batch[0][random.sample(benchmark_batch[1][1][:, i].nonzero().flatten().tolist(), n_samples)] for i in range(benchmark_batch[1][1].size(-1))])
 
     model = Model(is_contrastive=is_contrastive, multilabel=is_multilabel, xl_backbone=const.XL_BACKBONE, upsampling_level=const.UPSAMPLING_LEVEL,
-                  logits_only=True, disable_bn=const.DISABLE_BN, modified_bn=const.MODIFY_BN, backbone_acts=const.ACTIVATIONS, device=const.DEVICE)
+                  load_pretrained_weights=const.PRETRAINED_BACKBONE, n_classes=const.N_CLASSES, logits_only=True,
+                  disable_bn=const.DISABLE_BN, modified_bn=const.MODIFY_BN, backbone_acts=const.ACTIVATIONS, device=const.DEVICE)
 
     ema = optim.swa_utils.AveragedModel(model, device=const.DEVICE, avg_fn=optim.swa_utils.get_ema_avg_fn(1 - min(1, (1 - const.EMA_DECAY) * const.BATCH_SIZE * const.EMA_STEPS / const.EPOCHS)), use_buffers=True) if const.EMA else None
 
